---
title: "TCR-Toolkit-SCRATCH: Master Summary"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
params:
  project_name: "TCR-Toolkit-SCRATCH"
  outdir: "Master_Summary_Report"

  # Root of the staged per-module table directories (tables_dir/<module>/<file>).
  # Replaces the previous ~30 individually-wired file params: modules publish whole
  # tables/ directories, so new upstream tables become available here without another
  # round of process-input plumbing.
  tables_dir: "intables"

  seurat_rds: ""
  export_cells_file: ""

  # Column mapping / resolution
  label_col: ""
  sample_col: "sample"
  patient_col: "patient"
  condition_col: "condition"
  timepoint_col: "timepoint"
  clone_id_col: "clone_id"
  clone_size_col: "clone_size"
  consensus_cluster_col: "consensus_cluster"

  label_candidates: "annot;Annotation;celltype;CellType;predicted.celltype.l2;celltypist;azimuth_labels;seurat_clusters"
  sample_candidates: "sample;META_SAMPLE;orig.ident;sample_id"
  patient_candidates: "patient;META_PATIENT;patient_id;subject;donor"
  condition_candidates: "condition;META_TIMECOND;group;status"
  timepoint_candidates: "timepoint;visit;day;META_TIMECOND"
  clone_id_candidates: "clone_id;CTaa;clonotype;clone"
  clone_size_candidates: "clone_size;CloneSize;clone_n"
  consensus_cluster_candidates: "consensus_cluster;consensus;consensus_id"

  tcrdist_radius: 24
  top_n_annotations: 20
  top_n_clusters: 20
  top_n_genes: 25
  base_size: 12
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(plotly)
  library(reactable)
  library(htmltools)
})

dir.create(file.path(params$outdir, "tables"),  recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, "figures"), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

# ── Input access ────────────────────────────────────────────────────────────
# Every upstream module stages its tables under <tables_dir>/<module>/. A module that
# did not run on this route simply has no directory, so T() returns NULL and the
# corresponding section renders an explanatory note rather than silently nothing.
T <- function(module, name) {
  p <- file.path(params$tables_dir, module, name)
  if (!file.exists(p) || file.size(p) == 0) return(NULL)
  out <- tryCatch({
    if (tolower(tools::file_ext(name)) == "csv") fread(p) else fread(p, sep = "\t")
  }, error = function(e) NULL)
  if (is.null(out) || nrow(out) == 0) return(NULL)
  as.data.frame(out)
}

# `-P label_col=` (an empty params.label_col) reaches R as NULL, and nzchar(NULL) is
# logical(0), which makes if() error rather than return FALSE. Normalise every incoming
# string param to a length-1 character before testing it.
as_str1 <- function(x) {
  if (is.null(x) || length(x) == 0) return("")
  x <- as.character(x)[1]
  if (is.na(x)) "" else trimws(x)
}

# First match wins across several modules. CONSENSUS computes method_presence /
# method_cluster_counts over CoNGA and consensus labels as well, so its copies are
# strictly richer than CLUSTER_ROLLUP's GIANA/GLIPH2/TCRdist-only fallback.
# Count distinct values, ignoring NA and empty strings. VDJ_QC leaves patient_id blank
# when no GEX object is supplied, so a naive n_distinct() on that column returns 1 for
# "" rather than falling through to a source that actually carries the value.
n_distinct_real <- function(x) {
  if (is.null(x)) return(NA_integer_)
  x <- trimws(as.character(x))
  x <- x[!is.na(x) & nzchar(x)]
  if (!length(x)) return(NA_integer_)
  length(unique(x))
}

# First source with a real value wins.
coalesce_count <- function(...) {
  for (v in list(...)) if (!is.na(v) && v > 0) return(as.character(v))
  NA_character_
}

Tfirst <- function(modules, name) {
  for (m in modules) {
    d <- T(m, name)
    if (!is.null(d)) return(d)
  }
  NULL
}

have <- function(x) !is.null(x) && nrow(x) > 0

note <- function(txt) knitr::asis_output(paste0("\n::: {.callout-note appearance=\"simple\"}\n",
                                                txt, "\n:::\n\n"))

save_table <- function(df, filename) {
  if (have(df)) fwrite(df, file.path(params$outdir, "tables", filename), sep = "\t")
  invisible(df)
}

theme_ms <- function(base_size = params$base_size) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title      = element_text(face = "bold", size = base_size + 1, hjust = 0),
      axis.title      = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text      = element_text(face = "bold"),
      legend.title    = element_text(face = "bold")
    )
}

# ── Interactive output helpers ──────────────────────────────────────────────
# Figures are built with plot_ly() directly rather than ggplotly(). The container pairs
# ggplot2 4.0.2 with plotly 4.10.4, and that plotly predates ggplot2's 4.x internals, so
# ggplotly() dies with "subscript out of bounds" on most plots. Native plot_ly() calls do
# not touch ggplot internals and are version-robust.
#
# Every figure supports: hover for exact values, drag to zoom, double-click to reset,
# click legend entries to toggle series.
pal <- c("#2c7fb8", "#d95f02", "#7570b3", "#1b9e77",
         "#e7298a", "#66a61e", "#e6ab02", "#a6761d")

pconf <- function(p) {
  plotly::config(p, displaylogo = FALSE,
                 modeBarButtonsToRemove = c("select2d", "lasso2d"),
                 toImageButtonOptions = list(format = "png", scale = 2))
}

ttl <- function(t) list(text = t, x = 0, xanchor = "left",
                        font = list(size = params$base_size + 4))

# Horizontal bar. df: cat (chr), val (num), hover (chr)
hbar <- function(df, title, xlab, colour = "#2c7fb8", height = 440, tickformat = NULL) {
  df <- df[order(df$val), , drop = FALSE]
  df$cat <- factor(df$cat, levels = unique(df$cat))
  pconf(plotly::plot_ly(df, x = ~val, y = ~cat, type = "bar", orientation = "h",
                        marker = list(color = colour),
                        text = ~hover, hovertemplate = "%{text}<extra></extra>") %>%
        plotly::layout(title = ttl(title), height = height,
                       margin = list(l = 140, t = 60),
                       xaxis = list(title = xlab, tickformat = tickformat),
                       yaxis = list(title = "")))
}

# Stacked horizontal bar. df: cat, val, grp, hover
sbar <- function(df, title, xlab, height = 470, tickformat = ".0%") {
  pconf(plotly::plot_ly(df, x = ~val, y = ~cat, color = ~grp, colors = pal,
                        type = "bar", orientation = "h",
                        text = ~hover, hovertemplate = "%{text}<extra></extra>") %>%
        plotly::layout(barmode = "stack", title = ttl(title), height = height,
                       margin = list(l = 140, t = 60),
                       xaxis = list(title = xlab, tickformat = tickformat),
                       yaxis = list(title = ""),
                       legend = list(orientation = "h", y = -0.15)))
}

# Grouped vertical bar. df: cat, val, grp, hover
gbar <- function(df, title, xlab, ylab, height = 470) {
  pconf(plotly::plot_ly(df, x = ~cat, y = ~val, color = ~grp, colors = pal,
                        type = "bar",
                        text = ~hover, hovertemplate = "%{text}<extra></extra>") %>%
        plotly::layout(barmode = "group", title = ttl(title), height = height,
                       margin = list(t = 60),
                       xaxis = list(title = xlab), yaxis = list(title = ylab),
                       legend = list(orientation = "h", y = -0.2)))
}

# Box plot with points. df: grp, val, hover
boxby <- function(df, title, ylab, height = 470, logy = FALSE, angle = 0) {
  pconf(plotly::plot_ly(df, x = ~grp, y = ~val, color = ~grp, colors = pal,
                        type = "box", boxpoints = "all", jitter = 0.4, pointpos = 0,
                        marker = list(size = 4, opacity = 0.55),
                        text = ~hover, hoverinfo = "text") %>%
        plotly::layout(title = ttl(title), height = height, showlegend = FALSE,
                       margin = list(t = 60, b = 110),
                       xaxis = list(title = "", tickangle = angle),
                       yaxis = list(title = ylab,
                                    type = if (logy) "log" else "linear")))
}

# Scatter. df: x, y, grp, hover  (grp may be a constant)
scat <- function(df, title, xlab, ylab, height = 470, size = 11) {
  pconf(plotly::plot_ly(df, x = ~x, y = ~y, color = ~grp, colors = pal,
                        type = "scatter", mode = "markers",
                        marker = list(size = size, opacity = 0.85),
                        text = ~hover, hovertemplate = "%{text}<extra></extra>") %>%
        plotly::layout(title = ttl(title), height = height, margin = list(t = 60),
                       xaxis = list(title = xlab), yaxis = list(title = ylab)))
}

# Line/step. df: x, y, grp, hover
lplot <- function(df, title, xlab, ylab, logx = TRUE, logy = TRUE, height = 470) {
  pconf(plotly::plot_ly(df, x = ~x, y = ~y, color = ~grp, colors = pal,
                        type = "scatter", mode = "lines",
                        text = ~hover, hovertemplate = "%{text}<extra></extra>") %>%
        plotly::layout(title = ttl(title), height = height, margin = list(t = 60),
                       xaxis = list(title = xlab, type = if (logx) "log" else "linear"),
                       yaxis = list(title = ylab, type = if (logy) "log" else "linear")))
}

itable <- function(df, caption = NULL, pageSize = 10, ...) {
  if (!have(df)) return(note("No data available."))
  reactable(
    df,
    searchable = TRUE, sortable = TRUE, filterable = FALSE,
    highlight = TRUE, compact = TRUE, striped = TRUE,
    defaultPageSize = pageSize, showPageSizeOptions = TRUE,
    pageSizeOptions = c(5, 10, 25, 50, 100),
    defaultColDef = colDef(
      format = colFormat(digits = 3),
      minWidth = 90,
      headerStyle = list(background = "#f7f7f8", fontWeight = 600)
    ),
    ...
  )
}
```

```{r}
#| label: load-all
# ── VDJ QC ──
vdj_compact    <- T("vdj_qc", "vdj_qc_per_sample_compact.tsv")
vdj_retention  <- T("vdj_qc", "qc_contigs_before_after_summary.tsv")
vdj_rank       <- T("vdj_qc", "clone_rank_abundance.tsv")
vdj_pairing    <- T("vdj_qc", "pairing_status_all.tsv")
vdj_cdr3len    <- T("vdj_qc", "cdr3_lengths_all.tsv")
vdj_vusage     <- T("vdj_qc", "v_gene_usage.tsv")
vdj_jusage     <- T("vdj_qc", "j_gene_usage.tsv")
vdj_samplesheet<- T("vdj_qc", "sample_sheet_resolved.tsv")

# ── Pseudobulk QC gate ──
pb_qc          <- T("pseudobulk", "pseudobulk_qc_summary.csv")
pb_vfam        <- T("pseudobulk", "pseudobulk_v_family.csv")

# ── Shared bulk engine ──
sample_stats   <- T("sample", "sample_stats.csv")
sample_vfam    <- T("sample", "v_family.csv")
sample_jfam    <- T("sample", "j_family.csv")
sharing        <- T("compare", "cdr3_sharing.tsv")

# ── Clustering rollups ──
giana_roll     <- T("rollup", "giana_summary_rollup.tsv")
gliph2_roll    <- T("rollup", "gliph2_summary_rollup.tsv")
tcrdist_roll   <- T("rollup", "tcrdist3_summary_rollup.tsv")
method_presence<- Tfirst(c("consensus", "rollup"), "method_presence_summary.tsv")
method_counts  <- Tfirst(c("consensus", "rollup"), "method_cluster_counts.tsv")
anno_giana     <- T("rollup", "annotation_giana_summary.tsv")
anno_gliph2    <- T("rollup", "annotation_gliph2_summary.tsv")
anno_tcrdist   <- T("rollup", "annotation_tcrdist3_summary.tsv")
giana_detail   <- T("rollup", "giana_cluster_detail.tsv")
gliph2_detail  <- T("rollup", "gliph2_motif_detail.tsv")
tcrdist_detail <- T("rollup", "tcrdist_cluster_detail.tsv")
giana_vgene    <- T("rollup", "giana_vgene_usage.tsv")
gliph2_vgene   <- T("rollup", "gliph2_vgene_usage.tsv")

# ── Repertoire ──
rep_rollup     <- T("repertoire", "repertoire_summary_rollup.tsv")
diversity      <- T("repertoire", "diversity_by_sample.tsv")
div_patient    <- T("repertoire", "diversity_by_patient.tsv")
clone_burden   <- T("repertoire", "clone_burden_cells_by_sample.tsv")
overlap_mat    <- T("repertoire", "sample_overlap_matrix.tsv")
top_shared     <- T("repertoire", "top_shared_clones.tsv")
shareability   <- T("repertoire", "clone_shareability.tsv")
pheno_flux     <- T("repertoire", "phenotypic_flux.tsv")

# ── GEX-dependent ──
tcell_roll     <- T("tcell", "tcell_summary_rollup.tsv")
tcell_anno     <- T("tcell", "tcell_per_annotation_summary.tsv")
tcell_topclone <- T("tcell", "tcell_top_clones.tsv")
tcell_overlap  <- T("tcell", "barcode_overlap_by_sample.tsv")
merged_vdj     <- T("mergedvdj", "post_qc_summary.tsv")
conga_roll     <- T("conga", "conga_summary_rollup.tsv")
conga_clusters <- T("conga", "conga_cluster_summary.tsv")
anno_conga     <- T("conga", "annotation_conga_summary.tsv")
cons_roll      <- T("consensus", "consensus_summary_rollup.tsv")
tcri_roll      <- T("tcri", "tcri_summary_rollup.tsv")
tcri_scores    <- T("tcri", "high_tcri_clone_summary.tsv")
anno_tcri      <- T("tcri", "annotation_tcri_summary.tsv")
cons_clusters  <- T("consensus", "consensus_cluster_summary.tsv")
anno_cons      <- T("consensus", "annotation_consensus_summary.tsv")

# ── Per-cell export ──
export_cells <- NULL
.ec <- as_str1(params$export_cells_file)
if (nzchar(.ec) && file.exists(.ec) && file.size(.ec) > 0) {
  export_cells <- tryCatch(as.data.frame(fread(.ec, sep = "\t")),
                           error = function(e) NULL)
}

pick_col <- function(df, preferred, candidates) {
  if (is.null(df) || !ncol(df)) return(NULL)
  preferred <- as_str1(preferred)
  cand <- trimws(unlist(strsplit(gsub('"', "", as_str1(candidates)), ";", fixed = TRUE)))
  cand <- cand[nzchar(cand)]
  if (nzchar(preferred) && preferred %in% names(df)) return(preferred)
  hit <- intersect(cand, names(df))
  if (length(hit)) hit[[1]] else NULL
}
sample_col  <- pick_col(export_cells, params$sample_col,  params$sample_candidates)
patient_col <- pick_col(export_cells, params$patient_col, params$patient_candidates)
label_col   <- pick_col(export_cells, params$label_col,   params$label_candidates)
clone_col   <- pick_col(export_cells, params$clone_id_col, params$clone_id_candidates)
cons_col    <- pick_col(export_cells, params$consensus_cluster_col, params$consensus_cluster_candidates)

has_gex <- !is.null(tcell_roll) || !is.null(cons_roll) || !is.null(conga_roll) || !is.null(tcri_roll)
route   <- if (has_gex) "Full single-cell (GEX + VDJ)" else "VDJ-only (no GEX object)"
```

::: {.callout-tip appearance="simple"}
**All figures in this report are interactive.** Hover a point or bar for exact values, drag
to zoom, double-click to reset, and click legend entries to show or hide series. Tables are
sortable and searchable, with adjustable page size.
:::

# Overview

What was analysed, and how the run was routed. The route is detected from whether a GEX
annotated object was supplied: without one, the T-cell integration, CoNGA and consensus
steps are skipped and the sections that depend on them say so explicitly.

```{r}
#| label: overview
ov <- data.frame(
  Metric = c("Project", "Analysis route", "Samples", "Patients",
             "Cells / clonotype rows", "Unique clonotypes", "TCRdist radius"),
  Value = c(
    params$project_name,
    route,
    coalesce_count(
      if (have(vdj_compact)) nrow(vdj_compact) else NA_integer_,
      if (!is.null(export_cells) && !is.null(sample_col)) n_distinct_real(export_cells[[sample_col]]) else NA_integer_),
    coalesce_count(
      if (have(vdj_compact) && "patient_id" %in% names(vdj_compact)) n_distinct_real(vdj_compact$patient_id) else NA_integer_,
      if (!is.null(export_cells) && !is.null(patient_col)) n_distinct_real(export_cells[[patient_col]]) else NA_integer_),
    if (!is.null(export_cells)) format(nrow(export_cells), big.mark = ",") else NA,
    if (!is.null(export_cells) && !is.null(clone_col)) format(dplyr::n_distinct(export_cells[[clone_col]]), big.mark = ",") else NA,
    as.character(params$tcrdist_radius)
  ),
  stringsAsFactors = FALSE
)
save_table(ov, "master_overview.tsv")
itable(ov, pageSize = 10)
```

## Receptor retention

The full single-cell route pseudobulks from the **post-merge** export, so a receptor whose
barcode never matched a GEX cell is absent from every downstream result — not only the
cell-level steps, but GIANA, GLIPH2, TCRdist3 and the repertoire statistics too. Every
receptor is nevertheless conserved in `bridge/merged_vdj_object/`, flagged with whether it
reached the GEX analysis.

At a high match rate this is immaterial. At a low one it silently removes a large share of
the data, so the rate is reported here rather than left in a module table.

```{r}
#| label: receptor-retention
#| results: asis
if (!have(tcell_overlap)) {
  note(paste("No barcode overlap table — either the VDJ-only route (no GEX join is",
             "performed, so every receptor is analysed) or T-cell integration did not run."))
} else {
  d <- tcell_overlap
  nb <- intersect(c("tcr_barcodes","n_tcr"), names(d))
  no <- intersect(c("overlap","matched","matched_keys"), names(d))
  if (!length(nb) || !length(no)) {
    note("Overlap table present but lacks the expected barcode/overlap columns.")
  } else {
    tot <- sum(as.numeric(d[[nb[1]]]), na.rm = TRUE)
    hit <- sum(as.numeric(d[[no[1]]]), na.rm = TRUE)
    rate <- if (tot > 0) hit / tot else NA_real_
    save_table(d, "master_receptor_retention.tsv")

    if (!is.na(rate) && rate < 0.75) {
      cat(sprintf(paste0("\n::: {.callout-warning appearance=\"simple\"}\n",
                  "**%s of %s receptors (%.1f%%) matched a GEX cell.** The remaining %s are ",
                  "excluded from every downstream result on this route. They are preserved in ",
                  "`bridge/merged_vdj_object/` with `gex_matched = FALSE`. Below ~75%%, ",
                  "consider whether barcode conventions are reconciling correctly — the ",
                  "per-sample rates below will show if it is one sample or all of them.\n:::\n\n"),
                  format(hit, big.mark=","), format(tot, big.mark=","), 100*rate,
                  format(tot-hit, big.mark=",")))
    } else {
      cat(sprintf(paste0("\n::: {.callout-note appearance=\"simple\"}\n",
                  "**%s of %s receptors (%.1f%%) matched a GEX cell.** The %s unmatched are ",
                  "excluded from downstream results but preserved in ",
                  "`bridge/merged_vdj_object/` with `gex_matched = FALSE`.\n:::\n\n"),
                  format(hit, big.mark=","), format(tot, big.mark=","), 100*rate,
                  format(tot-hit, big.mark=",")))
    }
  }
}
```

```{r}
#| label: retention-by-sample
if (!have(tcell_overlap)) {
  invisible(NULL)
} else {
  d  <- tcell_overlap
  sc <- intersect(c("sample","Sample"), names(d))
  rc <- intersect(c("overlap_rate","rate"), names(d))
  nb <- intersect(c("tcr_barcodes","n_tcr"), names(d))
  no <- intersect(c("overlap","matched","matched_keys"), names(d))
  if (length(sc) && length(nb) && length(no)) {
    r <- if (length(rc)) as.numeric(d[[rc[1]]]) else
         as.numeric(d[[no[1]]]) / as.numeric(d[[nb[1]]])
    dd <- data.frame(cat = as.character(d[[sc[1]]]), val = r,
                     hover = paste0(d[[sc[1]]], "<br>",
                                    format(as.numeric(d[[no[1]]]), big.mark=","), " of ",
                                    format(as.numeric(d[[nb[1]]]), big.mark=","),
                                    " matched<br>", percent(r, accuracy = 0.1)))
    hbar(dd, "Receptors matching a GEX cell, by sample", "Match rate",
         height = 400, tickformat = ".0%")
  }
}
```

## Module coverage

Which analysis modules contributed output to this report. `FALSE` for the GEX-dependent
modules is expected on the VDJ-only route and does not indicate a failure.

```{r}
#| label: module-availability
mods <- list(
  "VDJ QC"             = vdj_compact,
  "Pseudobulk QC"      = pb_qc,
  "Sample statistics"  = sample_stats,
  "Clonotype sharing"  = sharing,
  "GIANA"              = giana_roll,
  "GLIPH2"             = gliph2_roll,
  "TCRdist3"           = tcrdist_roll,
  "Repertoire"         = rep_rollup,
  "T-cell integration" = tcell_roll,
  "CoNGA"              = conga_roll,
  "Consensus"          = cons_roll,
  "TCRi"               = tcri_roll
)
avail <- data.frame(module = names(mods),
                    available = vapply(mods, have, logical(1)),
                    row.names = NULL)
save_table(avail, "module_output_availability.tsv")

d <- avail %>%
  mutate(status = ifelse(available, "present", "absent"),
         cat = module, val = as.integer(available),
         hover = paste0(module, ": ", status))
hbar(d, "Module output availability", "", colour = "#2c7fb8", height = 460)
```

# Sample QC and the pseudobulk gate

Before any clustering, every sample passes a QC gate on clonotype and cell counts. Samples
falling below `pseudobulk_qc_min_clones` / `pseudobulk_qc_min_cells` are dropped from all
downstream analysis, so this is the first place to look if a sample is missing later on.

```{r}
#| label: pseudobulk-gate
if (!have(pb_qc)) {
  note("No pseudobulk QC summary was staged for this run.")
} else {
  save_table(pb_qc, "master_pseudobulk_qc.tsv")
  d <- pb_qc %>%
    transmute(x = n_clones, y = n_cells, grp = as.character(pass),
              hover = paste0(sample, "<br>clones: ", n_clones,
                             "<br>cells: ", n_cells, "<br>", pass))
  p <- scat(d, "Pseudobulk QC gate (dashed lines = thresholds)", "Clonotypes", "Cells")
  p %>% plotly::layout(shapes = list(
    list(type = "line", x0 = pb_qc$min_clones[1], x1 = pb_qc$min_clones[1],
         y0 = 0, y1 = max(pb_qc$n_cells) * 1.05,
         line = list(dash = "dash", color = "grey")),
    list(type = "line", x0 = 0, x1 = max(pb_qc$n_clones) * 1.05,
         y0 = pb_qc$min_cells[1], y1 = pb_qc$min_cells[1],
         line = list(dash = "dash", color = "grey"))))
}
```

```{r}
#| label: pseudobulk-gate-table
itable(pb_qc, pageSize = 10)
```

# VDJ QC

Contig-level quality control from Cell Ranger output: how many contigs survived filtering,
how many cells carry a properly paired receptor, and the shape of the clone size
distribution. Low pairing or heavy multi-chain burden here propagates into every downstream
clustering result.

## Per-sample summary

```{r}
#| label: vdj-compact
if (!have(vdj_compact)) note("VDJ QC did not contribute a per-sample summary.") else {
  save_table(vdj_compact, "master_vdj_qc_per_sample.tsv")
  itable(vdj_compact, pageSize = 10)
}
```

## Contig retention

Fraction of contigs surviving QC filtering, per sample. A sample retaining markedly less
than its peers usually indicates a library or sequencing-depth problem rather than biology.

```{r}
#| label: vdj-retention
if (!have(vdj_retention)) note("No contig retention table available.") else {
  d <- vdj_retention
  ycol <- intersect(c("retained_pct", "retained_prop", "pct_retained"), names(d))
  if (!length(ycol)) note("Retention table present but has no recognisable retention column.") else {
    v <- as.numeric(d[[ycol[1]]]); if (max(v, na.rm = TRUE) <= 1) v <- 100 * v
    dd <- data.frame(cat = d$sample, val = v,
                     hover = paste0(d$sample, "<br>retained: ", round(v, 1), "%"))
    hbar(dd, "Contig retention after QC", "Retained (%)")
  }
}
```

## Chain pairing

Share of cells by chain-pairing status. Only cells with a paired receptor contribute a
usable clonotype, so this sets the effective ceiling on everything downstream.

```{r}
#| label: vdj-pairing
if (!have(vdj_pairing) || !all(c("sample", "pairing") %in% names(vdj_pairing))) {
  note("No pairing status table available.")
} else {
  d <- vdj_pairing %>%
    count(sample, pairing, name = "cells") %>%
    group_by(sample) %>% mutate(frac = cells / sum(cells)) %>% ungroup()
  save_table(d, "master_pairing_by_sample.tsv")
  dd <- d %>% transmute(cat = sample, val = frac, grp = pairing,
                        hover = paste0(sample, "<br>", pairing, "<br>",
                                       cells, " cells (", percent(frac, accuracy = 0.1), ")"))
  sbar(dd, "Chain pairing status by sample", "Fraction of cells")
}
```

## CDR3 length distribution

CDR3β length distribution across all productive contigs. A healthy repertoire is centred
near 14–15 amino acids; sharp spikes can indicate contamination or a dominant clone.

```{r}
#| label: cdr3-length
if (!have(vdj_cdr3len)) note("No CDR3 length table available.") else {
  d <- vdj_cdr3len
  lc <- intersect(c("cdr3_length", "length"), names(d))
  if (!length(lc)) note("CDR3 length table has no length column.") else {
    d$.len <- suppressWarnings(as.integer(d[[lc[1]]]))
    if ("chain" %in% names(d)) d <- d[d$chain %in% c("TRB", "TRA", "TRD", "TRG"), , drop = FALSE]
    d <- d[!is.na(d$.len) & d$.len > 0 & d$.len < 40, , drop = FALSE]
    if (!nrow(d)) note("No usable CDR3 lengths after filtering.") else {
      dd <- if ("chain" %in% names(d)) count(d, .len, chain, name = "n") else
              cbind(count(d, .len, name = "n"), chain = "all")
      dd <- dd %>% transmute(cat = .len, val = n, grp = chain,
                             hover = paste0(chain, "<br>length ", .len, "<br>",
                                            format(n, big.mark = ","), " contigs"))
      gbar(dd, "CDR3 length distribution", "CDR3 length (aa)", "Contigs")
    }
  }
}
```

## V and J gene usage

Germline gene usage per sample. Systematic shifts between samples can reflect real
biology, but can equally flag a batch or primer effect worth ruling out.

```{r}
#| label: gene-usage
usage_box <- function(d, gcol, title) {
  if (!have(d)) return(NULL)
  fc <- intersect(c("freq", "frequency", "prop", "n"), names(d))
  if (!length(fc) || !(gcol %in% names(d))) return(NULL)
  d$.f <- as.numeric(d[[fc[1]]])
  top <- d %>% group_by(.data[[gcol]]) %>%
    summarise(m = mean(.f, na.rm = TRUE), .groups = "drop") %>%
    slice_max(m, n = params$top_n_genes, with_ties = FALSE) %>% pull(.data[[gcol]])
  dd <- d[d[[gcol]] %in% top, , drop = FALSE]
  ord <- dd %>% group_by(.data[[gcol]]) %>% summarise(m = median(.f), .groups = "drop") %>%
    arrange(desc(m)) %>% pull(.data[[gcol]])
  out <- data.frame(grp = factor(dd[[gcol]], levels = ord), val = dd$.f,
                    hover = paste0(dd[[gcol]], "<br>", dd$sample, "<br>", signif(dd$.f, 3)))
  boxby(out, title, "Usage frequency", height = 520, angle = -45)
}
pv <- usage_box(vdj_vusage, "v_gene", paste0("Top ", params$top_n_genes, " V genes"))
if (is.null(pv)) note("No V gene usage table available.") else pv
```

```{r}
#| label: j-usage
pj <- usage_box(vdj_jusage, "j_gene", paste0("Top ", params$top_n_genes, " J genes"))
if (is.null(pj)) note("No J gene usage table available.") else pj
```

# Repertoire diversity and clonal structure

Diversity summarises how evenly the repertoire is distributed across clonotypes. Low
Shannon entropy with high clonality indicates a repertoire dominated by a few expanded
clones — typically the signal of interest in an antigen-driven response.

## Diversity by sample

```{r}
#| label: diversity
if (!have(diversity)) {
  note("No diversity table available. This is produced by the Repertoire module.")
} else {
  save_table(diversity, "master_diversity_by_sample.tsv")
  metrics <- intersect(c("shannon", "simpson", "inv_simpson", "richness"), names(diversity))
  d <- diversity %>%
    pivot_longer(all_of(metrics), names_to = "metric", values_to = "value") %>%
    transmute(grp = metric, val = value,
              hover = paste0(sample, "<br>", metric, ": ", signif(value, 4)))
  boxby(d, "Repertoire diversity across samples", "Metric value")
}
```

```{r}
#| label: diversity-table
itable(diversity, pageSize = 10)
```

## Sample statistics

Clonality, productive fraction and convergence per sample, from the shared bulk engine.
Convergent recombination — distinct nucleotide sequences encoding the same CDR3 amino acid
sequence — is a useful independent marker of antigen-driven selection.

```{r}
#| label: sample-stats
if (!have(sample_stats)) note("No sample statistics table available.") else {
  save_table(sample_stats, "master_sample_stats.tsv")
  itable(sample_stats, pageSize = 10)
}
```

```{r}
#| label: clonality-plot
if (!have(sample_stats) || !("clonality" %in% names(sample_stats))) {
  note("No clonality column in the sample statistics table.")
} else {
  d <- sample_stats %>%
    transmute(cat = sample, val = clonality,
              hover = paste0(sample, "<br>clonality: ", signif(clonality, 3),
                             "<br>clonotypes: ", num_clones))
  hbar(d, "Clonality by sample", "Clonality", colour = "#7570b3")
}
```

## Clonal burden

How much of each sample's cell mass sits in small versus expanded clones. A sample shifting
mass into the large-clone bins is undergoing clonal expansion.

```{r}
#| label: clone-burden
if (!have(clone_burden)) {
  note("No clone burden table available. This is produced by the Repertoire module.")
} else {
  save_table(clone_burden, "master_clone_burden.tsv")
  fc <- intersect(c("frac_cells", "frac_clones"), names(clone_burden))
  if (!("clone_bin" %in% names(clone_burden)) || !length(fc)) {
    note("Clone burden table lacks the expected clone_bin / fraction columns.")
  } else {
    d <- clone_burden
    d$.f <- as.numeric(d[[fc[1]]])
    dd <- d %>% transmute(cat = sample, val = .f, grp = clone_bin,
                          hover = paste0(sample, "<br>", clone_bin, "<br>",
                                         percent(.f, accuracy = 0.1)))
    sbar(dd, "Clonal burden across samples", "Fraction of cells")
  }
}
```

## Clone rank abundance

Clone size against rank. A steep head indicates a few dominant clones; a long flat tail
indicates a diverse, largely unexpanded repertoire.

```{r}
#| label: rank-abundance
ra <- if (have(vdj_rank)) vdj_rank else T("repertoire", "clone_rank_abundance.tsv")
if (!have(ra)) note("No clone rank abundance table available.") else {
  rc <- intersect(c("rank", "clone_rank"), names(ra))
  nc <- intersect(c("n", "cells", "count", "clone_size", "size"), names(ra))
  if (!length(rc) || !length(nc)) note("Rank abundance table lacks rank/size columns.") else {
    d <- ra
    d$.r <- as.numeric(d[[rc[1]]]); d$.n <- as.numeric(d[[nc[1]]])
    d <- d[!is.na(d$.r) & !is.na(d$.n) & d$.n > 0 & d$.r > 0, , drop = FALSE]
    grp <- if ("sample" %in% names(d)) as.character(d$sample) else "all"
    dd <- data.frame(x = d$.r, y = d$.n, grp = grp,
                     hover = paste0(grp, "<br>rank ", d$.r, "<br>", d$.n, " cells"))
    lplot(dd, "Clone rank abundance (log-log)", "Clone rank", "Clone size")
  }
}
```

# Repertoire overlap and clonotype sharing

Clonotypes seen in more than one sample. Within a patient this reflects genuine clonal
persistence across timepoints or tissues; across patients it usually reflects public
clonotypes with high generation probability rather than a shared response.

## Sample overlap

```{r}
#| label: overlap-heatmap
if (!have(overlap_mat)) {
  note("No sample overlap matrix available. This is produced by the Repertoire module.")
} else {
  save_table(overlap_mat, "master_sample_overlap.tsv")
  m <- overlap_mat
  rn <- as.character(m[[1]]); m[[1]] <- NULL
  mat <- as.matrix(m); rownames(mat) <- rn
  plot_ly(
    x = colnames(mat), y = rownames(mat), z = mat,
    type = "heatmap", colors = colorRamp(c("white", "#fee391", "#cb181d")),
    hovertemplate = "%{y} vs %{x}<br>overlap: %{z:.3f}<extra></extra>"
  ) %>%
    layout(title = "Sample repertoire overlap",
           xaxis = list(title = "", tickangle = -45), yaxis = list(title = "")) %>%
    config(displaylogo = FALSE)
}
```

## Most shared clonotypes

```{r}
#| label: top-shared
if (!have(top_shared)) note("No top shared clones table available.") else {
  save_table(top_shared, "master_top_shared_clones.tsv")
  itable(top_shared, pageSize = 10)
}
```

## Sharing versus generation probability

Each clonotype's sharing breadth against its OLGA generation probability. Clonotypes shared
widely *and* easy to generate are likely public/convergent rather than response-specific —
this plot is the quickest way to tell those apart.

```{r}
#| label: sharing-pgen
if (!have(sharing)) note("No clonotype sharing table available (produced by COMPARE).") else {
  d <- sharing
  pg <- intersect(c("log10_pgen", "pgen"), names(d))
  ns <- intersect(c("total_samples", "n_samples"), names(d))
  if (!length(pg) || !length(ns)) note("Sharing table lacks pgen / sample-count columns.") else {
    d$.p <- as.numeric(d[[pg[1]]]); d$.n <- as.integer(d[[ns[1]]])
    if (pg[1] == "pgen") d$.p <- log10(pmax(d$.p, 1e-300))
    d <- d[is.finite(d$.p) & !is.na(d$.n), , drop = FALSE]
    dd <- data.frame(grp = factor(d$.n, levels = sort(unique(d$.n))), val = d$.p,
                     hover = paste0(d$junction_aa, "<br>samples: ", d$.n,
                                    "<br>log10 pgen: ", signif(d$.p, 4)))
    boxby(dd, "Generation probability by sharing breadth", "log10 pgen")
  }
}
```

# Sequence clustering

Three complementary approaches to grouping clonotypes by receptor similarity: **GIANA**
(isometric distance on CDR3β + V gene), **GLIPH2** (shared local motif enrichment) and
**TCRdist3** (pairwise TCR distance, thresholded at radius `r params$tcrdist_radius`).
GIANA and GLIPH2 run per patient; TCRdist3 runs per sample.

## Method coverage

Fraction of cells whose clonotype was assigned to a cluster by each method, and how many
clusters each produced. Large disagreements between methods are expected — they optimise
for different notions of similarity — but a method at or near zero is worth investigating.

```{r}
#| label: method-coverage
if (!have(method_presence)) {
  note("No method coverage table available.")
} else {
  save_table(method_presence, "master_method_presence.tsv")
  d <- method_presence %>%
    transmute(cat = method, val = frac_cells,
              hover = paste0(method, "<br>", percent(frac_cells, accuracy = 0.1), " of cells"))
  hbar(d, "Cell coverage by method", "Fraction of cells", height = 380, tickformat = ".0%")
}
```

```{r}
#| label: method-cluster-counts
if (!have(method_counts)) {
  note("No cluster count table available.")
} else {
  save_table(method_counts, "master_method_cluster_counts.tsv")
  d <- method_counts %>%
    transmute(cat = method, val = n_clusters,
              hover = paste0(method, "<br>", n_clusters, " clusters"))
  hbar(d, "Clusters detected by method", "Clusters", colour = "#d95f02", height = 380)
}
```

## Per-method summaries

```{r}
#| label: cluster-rollups
rolls <- list(GIANA = giana_roll, GLIPH2 = gliph2_roll, TCRdist3 = tcrdist_roll)
rolls <- rolls[vapply(rolls, have, logical(1))]
if (!length(rolls)) note("No clustering rollups available.") else {
  combined <- bind_rows(lapply(names(rolls), function(n)
    data.frame(method = n, rolls[[n]], stringsAsFactors = FALSE)))
  save_table(combined, "master_cluster_rollups.tsv")
  itable(combined, pageSize = 15,
         groupBy = "method",
         columns = list(method = colDef(name = "Method")))
}
```

::: {.callout-note appearance="simple"}
GIANA receives one row per unique clonotype. Before this was enforced, the same clonotype
seen in several samples of a patient arrived as duplicate rows and GIANA reported those
duplicates as clusters — inflating cluster counts and producing no output at all for
patients with a single sample.
:::

## Cluster composition

Every cluster from each method, sized by membership. `n_unique_cdr3` separates real
sequence similarity from duplicate detection: a cluster whose members are all the same
CDR3β is the same clonotype seen in several samples, not a similarity group.

```{r}
#| label: cluster-sizes
#| results: asis
det <- list(GIANA = giana_detail, GLIPH2 = gliph2_detail, TCRdist3 = tcrdist_detail)
det <- det[vapply(det, have, logical(1))]
if (!length(det)) {
  note("No cluster detail tables available.")
} else {
  rows <- lapply(names(det), function(nm) {
    d <- det[[nm]]
    data.frame(method = nm,
               size   = as.numeric(d$n_members),
               stringsAsFactors = FALSE)
  })
  d <- bind_rows(rows) %>%
    count(method, size, name = "n_clusters") %>%
    transmute(cat = size, val = n_clusters, grp = method,
              hover = paste0(method, "<br>", n_clusters, " cluster(s) of size ", size))
  save_table(d, "master_cluster_size_distribution.tsv")
  gbar(d, "Cluster size distribution by method", "Members per cluster", "Clusters")
}
```

```{r}
#| label: gliph2-top-motifs
if (!have(gliph2_detail)) {
  note("No GLIPH2 motif detail available.")
} else {
  save_table(gliph2_detail, "master_gliph2_motif_detail.tsv")
  d <- gliph2_detail %>%
    slice_max(n_members, n = params$top_n_clusters, with_ties = FALSE) %>%
    transmute(cat = paste0(cluster, "  (", patient, ")"),
              val = n_members,
              hover = paste0("motif ", cluster, "<br>patient ", patient,
                             "<br>", n_members, " members, ", n_unique_cdr3,
                             " distinct CDR3b<br>", n_samples, " sample(s)"))
  hbar(d, paste0("Top ", params$top_n_clusters, " GLIPH2 motif clusters"),
       "Members", colour = "#d95f02", height = 560)
}
```

```{r}
#| label: cross-sample-reach
#| results: asis
rows <- list()
for (nm in names(det)) {
  d <- det[[nm]]
  if (!("n_samples" %in% names(d))) next
  rows[[nm]] <- data.frame(
    method = nm,
    frac   = mean(as.numeric(d$n_samples) > 1, na.rm = TRUE),
    n      = nrow(d), stringsAsFactors = FALSE)
}
if (!length(rows)) {
  note("No per-cluster sample counts available.")
} else {
  d <- bind_rows(rows)
  save_table(d, "master_cluster_cross_sample.tsv")
  dd <- d %>% transmute(cat = method, val = frac,
                        hover = paste0(method, "<br>",
                                       percent(frac, accuracy = 0.1),
                                       " of ", n, " clusters span >1 sample"))
  hbar(dd, "Clusters spanning more than one sample", "Fraction of clusters",
       height = 380, tickformat = ".0%")
}
```

A cluster spanning several samples is the interesting case — the same or similar receptor
recovered independently, rather than confined to one library.

```{r}
#| label: giana-cluster-table
if (!have(giana_detail)) {
  note("No GIANA cluster detail available.")
} else {
  save_table(giana_detail, "master_giana_cluster_detail.tsv")
  itable(giana_detail, pageSize = 10)
}
```

## V-gene usage within clusters

Germline V-gene composition of the clustered clonotypes, per patient. A motif family
dominated by one V gene is a stronger structural signal than one drawn from many.

```{r}
#| label: cluster-vgene
#| results: asis
vg <- list(GIANA = giana_vgene, GLIPH2 = gliph2_vgene)
vg <- vg[vapply(vg, have, logical(1))]
if (!length(vg)) {
  note("No V-gene usage tables available for the clustering methods.")
} else {
  d <- bind_rows(lapply(names(vg), function(nm)
        data.frame(method = nm, vg[[nm]], stringsAsFactors = FALSE))) %>%
    group_by(method) %>%
    slice_max(n_clonotypes, n = params$top_n_genes, with_ties = FALSE) %>%
    ungroup()
  save_table(d, "master_cluster_vgene_usage.tsv")
  dd <- d %>% transmute(cat = vgene, val = n_clonotypes, grp = method,
                        hover = paste0(method, "<br>", vgene, "<br>patient ", patient,
                                       "<br>", n_clonotypes, " clonotypes"))
  gbar(dd, "V-gene usage among clustered clonotypes", "", "Clonotypes", height = 520)
}
```

## Per-annotation clustering coverage

Fraction of cells clustered by each method, broken down by cell-type annotation. Requires
cell-type labels from a GEX object.

```{r}
#| label: annotation-panels
panels <- list(GIANA = anno_giana, GLIPH2 = anno_gliph2, TCRdist3 = anno_tcrdist,
               CoNGA = anno_conga, Consensus = anno_cons, TCRi = anno_tcri)
rows <- list()
for (nm in names(panels)) {
  d <- panels[[nm]]
  if (!have(d)) next
  ac <- if ("annot" %in% names(d)) "annot" else names(d)[1]
  fc <- intersect(c("frac_giana", "frac_gliph", "frac_tcrdist", "frac_clustered",
                    "frac_high", "frac_consensus", "frac_tcri", "frac_high_tcri"), names(d))
  if (!length(fc)) next
  rows[[nm]] <- data.frame(method = nm, annot = as.character(d[[ac]]),
                           frac = as.numeric(d[[fc[1]]]), stringsAsFactors = FALSE)
}
if (!length(rows)) {
  note(paste("No per-annotation summaries available. These require cell-type labels from a",
             "GEX object, so this section is empty on the VDJ-only route."))
} else {
  d <- bind_rows(rows) %>%
    group_by(method) %>% slice_max(frac, n = params$top_n_annotations, with_ties = FALSE) %>%
    ungroup()
  save_table(d, "master_annotation_coverage.tsv")
  dd <- d %>% transmute(cat = annot, val = frac, grp = method,
                        hover = paste0(method, "<br>", annot, "<br>",
                                       percent(frac, accuracy = 0.1)))
  gbar(dd, "Clustering coverage by cell-type annotation", "", "Fraction of cells", height = 560)
}
```

# GEX-dependent analyses

These sections require an annotated Seurat object (`--input_annotated_object`) and are
skipped entirely on the VDJ-only route.

## T-cell integration

Fusion of VDJ contigs with the GEX object, matching receptors to transcriptome-defined cell
states.

```{r}
#| label: tcell
if (!have(tcell_roll)) {
  note("T-cell integration did not run on this route (no GEX object supplied).")
} else {
  save_table(tcell_roll, "master_tcell_rollup.tsv")
  itable(tcell_roll, pageSize = 15)
}
```

```{r}
#| label: tcell-annot
if (!have(tcell_anno)) {
  note("No per-annotation T-cell integration summary available.")
} else {
  d <- tcell_anno
  nc <- intersect(c("n_cells", "cells", "n"), names(d))
  ac <- if ("annot" %in% names(d)) "annot" else names(d)[1]
  if (!length(nc)) note("T-cell annotation summary lacks a cell-count column.") else {
    dd <- data.frame(cat = as.character(d[[ac]]), val = as.numeric(d[[nc[1]]]))
    dd$hover <- paste0(dd$cat, "<br>", format(dd$val, big.mark = ","), " cells")
    hbar(dd, "Cells per annotation", "Cells", height = 560)
  }
}
```

## CoNGA

Graph-level correspondence between TCR similarity and transcriptional similarity. Clusters
scoring highly here contain cells that are neighbours in *both* spaces — the strongest
available evidence of antigen-driven clonal programmes.

```{r}
#| label: conga
if (!have(conga_roll) && !have(conga_clusters)) {
  note("CoNGA did not run on this route (requires a GEX object).")
} else {
  if (have(conga_roll)) save_table(conga_roll, "master_conga_rollup.tsv")
  if (have(conga_clusters)) save_table(conga_clusters, "master_conga_clusters.tsv")
  itable(if (have(conga_roll)) conga_roll else conga_clusters, pageSize = 15)
}
```

## Consensus clustering

Reconciliation of the GIANA, GLIPH2, TCRdist3 and CoNGA labels into a single consensus
cluster per cell. Consensus clusters supported by several independent methods are the most
trustworthy candidates for follow-up.

```{r}
#| label: consensus
if (!have(cons_clusters)) {
  note("Consensus clustering did not run on this route (requires a GEX object).")
} else {
  save_table(cons_clusters, "master_consensus_clusters.tsv")
  d <- cons_clusters
  nc <- intersect(c("n_cells", "cells", "n"), names(d))
  cc <- intersect(c("consensus_cluster", "cluster"), names(d))
  if (!length(nc) || !length(cc)) itable(d, pageSize = 15) else {
    dd <- data.frame(cat = as.character(d[[cc[1]]]), val = as.numeric(d[[nc[1]]]))
    dd <- dd[order(-dd$val), , drop = FALSE]
    dd <- head(dd, params$top_n_clusters)
    dd$hover <- paste0("cluster ", dd$cat, "<br>", dd$val, " cells")
    hbar(dd, paste0("Top ", params$top_n_clusters, " consensus clusters"), "Cells", height = 560)
  }
}
```

```{r}
#| label: consensus-table
itable(cons_clusters, pageSize = 10)
```

## TCRi

Immunogenicity scoring: how strongly each clone's receptor resembles known
antigen-reactive receptors. High-scoring clones are the candidates worth following up
experimentally. Requires the GEX object, so this section is empty on the VDJ-only route.

```{r}
#| label: tcri
#| results: asis
if (!have(tcri_roll)) {
  note("TCRi did not run on this route (requires a GEX object).")
} else {
  save_table(tcri_roll, "master_tcri_rollup.tsv")
  itable(tcri_roll, pageSize = 15)
}
```

```{r}
#| label: tcri-clones
if (!have(tcri_scores)) {
  note("No high-TCRi clone summary available.")
} else {
  save_table(tcri_scores, "master_tcri_high_clones.tsv")
  d  <- tcri_scores
  nc <- intersect(c("n_cells", "cells", "clone_size", "n"), names(d))
  cc <- intersect(c("clone_id", "CTaa", "clonotype"), names(d))
  if (!length(nc) || !length(cc)) {
    itable(d, pageSize = 10)
  } else {
    dd <- data.frame(cat = as.character(d[[cc[1]]]), val = as.numeric(d[[nc[1]]]))
    dd <- head(dd[order(-dd$val), , drop = FALSE], params$top_n_clusters)
    dd$hover <- paste0(dd$cat, "<br>", dd$val, " cells")
    hbar(dd, paste0("Top ", params$top_n_clusters, " high-TCRi clones"), "Cells", height = 560)
  }
}
```

# Reference tables

Every table rendered above is also written to `Master_Summary_Report/tables/` as TSV for
downstream use.

```{r}
#| label: written-tables
files <- list.files(file.path(params$outdir, "tables"), pattern = "\\.tsv$")
if (!length(files)) note("No tables were written.") else {
  itable(data.frame(table = sort(files), stringsAsFactors = FALSE), pageSize = 25)
}
```

# Session info

```{r}
#| label: session
writeLines(capture.output(sessionInfo()),
           file.path(params$outdir, "sessionInfo.master_summary.txt"))
sessionInfo()
```
