---
title: "SCRATCH-TCR: T-Cell Integration Report"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
params:
  # ======================================================
  # Inputs
  # ======================================================
  seurat_rds: "data/project_annotated_object.RDS"
  contigs_file: "data/contigs_after_qc.tsv"
  sample_sheet: ""
  metadata_file: ""
  outdir: "TCell_Integration_Report"

  # Optional upstream artifacts
  clonotypes_file: ""
  vdj_qc_summary_file: ""
  barcode_harmonization_required: true

  # ======================================================
  # General output controls
  # ======================================================
  data_dir: "data"
  tables_dir: "tables"
  figures_dir: "figures"
  save_tables: true
  save_figures: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  # ======================================================
  # Metadata mapping in Seurat
  # ======================================================
  label_col: ""
  sample_col: "orig.ident"
  patient_col: ""
  condition_col: ""
  timepoint_col: ""
  batch_col: ""

  sample_candidates: !expr c("META_SAMPLE","orig.ident","sample","sample_id","Sample","SampleID")
  patient_candidates: !expr c("META_PATIENT","patient_id","patient","Patient","subject","donor","case_id")
  condition_candidates: !expr c("META_TIMECOND","condition","Condition","group","Group","status","timepoint","Timepoint")
  timepoint_candidates: !expr c("META_TIMECOND","timepoint","Timepoint","visit","Visit","day","Day")
  batch_candidates: !expr c("META_BATCH","batch","Batch","library","Library","run","Run","lane","Lane")

  label_candidates: !expr c(
    "predicted.celltype.l2","predicted.celltype.l1","azimuth_labels",
    "celltypist","celltypist_label","majority_voting",
    "scType","sctype","sctype_label","celltype","CellType","annot","Annotation")

  # ======================================================
  # Contig mapping
  # ======================================================
  contig_sample_col: "sample"
  contig_barcode_col: "barcode"
  contig_chain_col: "chain"
  contig_cdr3_col: "cdr3"
  contig_v_col: "v_gene"
  contig_j_col: "j_gene"
  contig_clone_col: "raw_clonotype_id"

  contig_sample_candidates: !expr c("sample","sample_id","orig.ident")
  contig_barcode_candidates: !expr c("barcode","cell_id","cell","CB")
  contig_chain_candidates: !expr c("chain","locus")
  contig_cdr3_candidates: !expr c("cdr3","cdr3_aa","CDR3aa","junction_aa")
  contig_v_candidates: !expr c("v_gene","v_call","bestVGene")
  contig_j_candidates: !expr c("j_gene","j_call","bestJGene")
  contig_clone_candidates: !expr c("raw_clonotype_id","clonotype_id","clone_id")

  # ======================================================
  # Integration / scRepertoire controls
  # ======================================================
  cells_mode: "T-AB"           # T-AB / T-GD / both
  filter_to_t_ab: true
  combine_id: ""
  clone_call_preference: "aa"  # aa / nt / strict / gene
  keep_na_clonotypes: false

  # ======================================================
  # Barcode harmonization controls
  # ======================================================
  harmonize_apply_to: "tcr"    # tcr / seurat
  harmonization_overlap_threshold: 0.80
  minimum_final_overlap_fraction: 0.50
  clean_duplicated_sample_prefix: true

  # ======================================================
  # T-cell subset controls
  # ======================================================
  subset_tcells: true
  tcell_regex: "(?i)(^|[^A-Za-z])(t[ -]?cell|cd4|cd8|treg|trm|tem|naive t|memory t|exhausted t|cytotoxic t|gamma.?delta|gd t)([^A-Za-z]|$)"
  use_ident_if_label_missing: true

  # ======================================================
  # UMAP / plotting controls
  # ======================================================
  reduction_use: "umap"
  make_umap_if_missing: true
  umap_dims_max: 30
  umap_nfeatures: 3000
  raster_large_umap: true
  label_clusters: true

  # ======================================================
  # Clone summarization controls
  # ======================================================
  min_clone_size_plot: 2
  top_n_clones_umap: 12
  top_n_clone_table: 50
  clone_bins_quantiles: !expr c(0, 0.2, 0.4, 0.6, 0.8, 1.0)
  force_quantile_clone_bins: true

  # ======================================================
  # Report toggles
  # ======================================================
  show_barcode_harmonization: true
  show_pairing_summary: true
  show_clone_rank_plot: true
  show_clone_state_heatmap: true
  show_homeostasis: true
  show_diversity: true
  show_sample_panels: true
  show_condition_panels: true
  show_patient_panels: true
  show_timepoint_panels: true
  show_top_clone_umaps: true

  report_label: "T-Cell Integration"
---

# ========================================================
## Integrates post-QC VDJ contigs with an annotated Seurat object using scRepertoire and a robust fast-merge strategy.
## Performs barcode cleaning, harmonization checks, and merge diagnostics to maximize reliable TCR↔GEX matching.
## Adds per-cell TCR metadata to Seurat, including clonotype IDs, paired chains, and clone sizes.
## Builds a T-cell-focused integrated object and generates high-quality embedding and clonotype visualizations.
## Exports reusable RDS objects, diagnostic tables, and summary figures for downstream TCRi, CoNGA, and clustering modules.
# ========================================================

## setup
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(dplyr)
  library(data.table)
  library(stringr)
  library(ggplot2)
  library(patchwork)
  library(tidyr)
  library(purrr)
  library(Matrix)
  library(scales)
  library(readr)
  library(ComplexHeatmap)
  library(circlize)
  library(forcats)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(scRepertoire)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$data_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0 && !all(is.na(a))) a else b

# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold"),
#       axis.text = element_text(color = "black"),
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold"),
#       legend.title = element_text(face = "bold"),
#       legend.key = element_blank(),
#       plot.caption = element_text(size = base_size - 2, color = "grey40")
#     )
# }
# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold", size = base_size - 1),
#       # Shrink Y-axis labels slightly
#       axis.text.y = element_text(color = "black", size = base_size - 3), 
#       # Rotate X-axis labels 45 degrees and shrink
#       axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1), 
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold", size = base_size - 2),
#       legend.title = element_text(face = "bold", size = base_size - 2),
#       legend.text = element_text(size = base_size - 3),
#       plot.caption = element_text(size = base_size - 4, color = "grey40")
#     )
# }

theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0),
      axis.title = element_text(face = "bold", size = base_size - 1),
      axis.text.y = element_text(color = "black", size = base_size - 3),
      # Fix for Q2 & Q3: Smaller text and 45-degree rotation for X-axis
      axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
      strip.background = element_rect(fill = "grey95", color = "grey80"),
      strip.text = element_text(face = "bold", size = base_size - 2),
      legend.title = element_text(face = "bold", size = base_size - 2),
      legend.text = element_text(size = base_size - 3),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 4, color = "grey40")
    )
}



save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

save_rds_safe <- function(obj, filename) {
  saveRDS(obj, file.path(params$outdir, params$data_dir, filename))
}

safe_read_table <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "tab", "txt")) {
    fread(path, sep = "\t")
  } else {
    fread(path)
  }
}

require_file <- function(path, msg = NULL) {
  if (!file.exists(path)) stop(msg %||% paste("Missing required file:", path))
  path
}

first_existing_col <- function(df, preferred, candidates = character()) {
  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  NULL
}

has_col <- function(df, x) x %in% colnames(df)

detect_label_col <- function(seu, preferred = "", candidates = character(), use_ident = TRUE) {
  md <- seu@meta.data
  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(md)) return(preferred)
  hit <- intersect(candidates, colnames(md))
  if (length(hit)) return(hit[1])
  if (use_ident) {
    seu$seurat_clusters <- as.character(Idents(seu))
    return("seurat_clusters")
  }
  stop("Could not find a valid annotation column in Seurat metadata.")
}

choose_col <- function(df, primary = NULL, fallback = NULL) {
  nm <- colnames(df)
  if (!is.null(primary) && primary != "" && primary %in% nm) return(primary)
  if (!is.null(fallback) && fallback != "" && fallback %in% nm) return(fallback)
  NULL
}

clean_barcode <- function(x) sub("^([A-Za-z0-9-]+)_\\1_", "\\1_", x)

combineTCR_safe <- function(contig_list, samples, ID = NULL, cells = "T-AB") {
  fun <- get("combineTCR", asNamespace("scRepertoire"))
  fml <- names(formals(fun))
  args <- list()
  if ("contig_list" %in% fml) args$contig_list <- contig_list else if ("contig.list" %in% fml) args$contig.list <- contig_list else args[[1]] <- contig_list
  if ("samples" %in% fml) args$samples <- samples else args[[length(args)+1]] <- samples
  if ("ID" %in% fml) args$ID <- ID else args[[length(args)+1]] <- ID
  if ("cells" %in% fml) args$cells <- cells
  do.call(fun, args)
}

dropNA_clonotypes <- function(combined_obj) {
  detect_clone_col <- function(df) {
    candidates <- c("CTaa", "CTnt", "CTstrict", "CTgene", "clonotype", "cloneType", "cloneCall")
    hit <- intersect(candidates, colnames(df))
    if (length(hit) == 0) return(NA_character_)
    if ("CTaa" %in% hit) return("CTaa")
    hit[1]
  }
  filter_one <- function(df) {
    if (!is.data.frame(df) || nrow(df) == 0) return(df)
    cc <- detect_clone_col(df)
    if (is.na(cc)) return(df)
    keep <- !is.na(df[[cc]]) & df[[cc]] != "" & df[[cc]] != "NA"
    df[keep, , drop = FALSE]
  }
  if (is.list(combined_obj) && !is.data.frame(combined_obj)) {
    out <- lapply(combined_obj, filter_one)
    out <- out[vapply(out, nrow, integer(1)) > 0]
    return(out)
  }
  filter_one(combined_obj)
}

safe_make_umap <- function(seu, reduction_name = "umap", dims = 1:30, nfeatures = 3000, seed = 1234) {
  if ("RNA" %in% names(seu@assays)) DefaultAssay(seu) <- "RNA"
  npcs <- min(max(dims), max(2, ncol(seu) - 1), 50)
  nf <- min(nfeatures, nrow(seu))

  if (!"pca" %in% Reductions(seu)) {
    seu <- FindVariableFeatures(seu, nfeatures = nf, verbose = FALSE)
    seu <- ScaleData(seu, verbose = FALSE)
    seu <- RunPCA(seu, npcs = npcs, verbose = FALSE)
  }

  use_dims <- dims[dims <= npcs]
  if (length(use_dims) < 2) use_dims <- 1:min(10, npcs)

  seu <- FindNeighbors(seu, dims = use_dims, verbose = FALSE)
  seu <- RunUMAP(seu, dims = use_dims, reduction.name = reduction_name, verbose = FALSE)
  seu
}

score_harmonizer_candidates <- function(combined, seu, seu_sample_col = "orig.ident") {
  md <- seu@meta.data
  if (!seu_sample_col %in% colnames(md)) seu_sample_col <- "orig.ident"
  if (!seu_sample_col %in% colnames(md)) stop("Seurat lacks sample column: ", seu_sample_col)

  seu_barcodes <- colnames(seu)
  all_tcr <- data.table::rbindlist(combined, idcol = "sample", fill = TRUE)
  tcr_samples <- as.character(all_tcr$sample)
  tcr_barcodes <- as.character(all_tcr$barcode)

  add_minus1 <- function(x) ifelse(grepl("-\\d+$", x), x, paste0(x, "-1"))
  remove_minus1 <- function(x) sub("-\\d+$", "", x)
  prepend_us <- function(x, s) paste(s, x, sep = "_")
  append_us <- function(x, s) paste(x, s, sep = "_")
  prepend_dash <- function(x, s) paste(s, x, sep = "-")
  append_dash <- function(x, s) paste(x, s, sep = "-")

  candidates <- list(
    list(name = "identity",        f = function(x, s) x),
    list(name = "add_-1",          f = function(x, s) add_minus1(x)),
    list(name = "remove_-1",       f = function(x, s) remove_minus1(x)),
    list(name = "prepend_sample_", f = function(x, s) prepend_us(x, s)),
    list(name = "append_sample_",  f = function(x, s) append_us(x, s)),
    list(name = "prepend-sample",  f = function(x, s) prepend_dash(x, s)),
    list(name = "append-sample",   f = function(x, s) append_dash(x, s))
  )

  score <- lapply(candidates, function(cn) {
    xb <- cn$f(tcr_barcodes, tcr_samples)
    inter <- length(intersect(xb, seu_barcodes))
    data.frame(
      transform = cn$name,
      overlap = inter,
      total = length(xb),
      rate = ifelse(length(xb) > 0, inter / length(xb), NA_real_),
      stringsAsFactors = FALSE
    )
  })
  score <- do.call(rbind, score)
  score <- score[order(-score$overlap, -score$rate), , drop = FALSE]
  list(table = score, selected = score$transform[1])
}

harmonize_barcodes <- function(combined, seu, seu_sample_col = "orig.ident",
                               apply_to = c("tcr", "seurat"), verbose = TRUE) {
  apply_to <- match.arg(apply_to)
  md <- seu@meta.data
  seu_barcodes <- colnames(seu)

  if (!seu_sample_col %in% colnames(md)) {
    warning("'", seu_sample_col, "' not in Seurat metadata; falling back to 'orig.ident'.")
    seu_sample_col <- "orig.ident"
  }
  if (!seu_sample_col %in% colnames(md)) stop("Seurat lacks a per-cell sample column.")

  seu_samples_per_cell <- as.character(md[[seu_sample_col]])
  all_tcr <- data.table::rbindlist(combined, idcol = "sample", fill = TRUE)
  tcr_samples <- as.character(all_tcr$sample)
  tcr_barcodes <- as.character(all_tcr$barcode)

  add_minus1 <- function(x) ifelse(grepl("-\\d+$", x), x, paste0(x, "-1"))
  remove_minus1 <- function(x) sub("-\\d+$", "", x)
  prepend_us <- function(x, s) paste(s, x, sep = "_")
  append_us <- function(x, s) paste(x, s, sep = "_")
  prepend_dash <- function(x, s) paste(s, x, sep = "-")
  append_dash <- function(x, s) paste(x, s, sep = "-")

  cand <- list(
    list(name = "identity",            side = "tcr",    f = function(x, s) x),
    list(name = "add_-1",              side = "tcr",    f = function(x, s) add_minus1(x)),
    list(name = "remove_-1",           side = "tcr",    f = function(x, s) remove_minus1(x)),
    list(name = "prepend_sample_",     side = "tcr",    f = function(x, s) prepend_us(x, s)),
    list(name = "append_sample_",      side = "tcr",    f = function(x, s) append_us(x, s)),
    list(name = "prepend-sample",      side = "tcr",    f = function(x, s) prepend_dash(x, s)),
    list(name = "append-sample",       side = "tcr",    f = function(x, s) append_dash(x, s)),
    list(name = "seurat_add_-1",       side = "seurat", f = function(x, s) add_minus1(x)),
    list(name = "seurat_remove_-1",    side = "seurat", f = function(x, s) remove_minus1(x)),
    list(name = "seurat_prepend_",     side = "seurat", f = function(x, s) prepend_us(x, s)),
    list(name = "seurat_append_",      side = "seurat", f = function(x, s) append_us(x, s)),
    list(name = "seurat_prepend-",     side = "seurat", f = function(x, s) prepend_dash(x, s)),
    list(name = "seurat_append-",      side = "seurat", f = function(x, s) append_dash(x, s))
  )

  score_tbl <- lapply(cand, function(cn) {
    if (cn$side == "tcr") {
      xb <- cn$f(tcr_barcodes, tcr_samples)
      inter <- length(intersect(xb, seu_barcodes))
      list(name = cn$name, side = "tcr", overlap = inter, total = length(xb))
    } else {
      if (apply_to != "seurat") return(NULL)
      xb <- cn$f(seu_barcodes, seu_samples_per_cell)
      inter <- length(intersect(tcr_barcodes, xb))
      list(name = cn$name, side = "seurat", overlap = inter, total = length(xb))
    }
  })
  score_tbl <- score_tbl[!sapply(score_tbl, is.null)]
  score_tbl <- do.call(rbind, lapply(score_tbl, as.data.frame))
  score_tbl$rate <- score_tbl$overlap / pmax(score_tbl$total, 1)

  pool <- subset(score_tbl, side == apply_to)
  best <- pool[which.max(pool$overlap), , drop = FALSE]

  if (verbose) {
    print(pool[order(-pool$overlap, -pool$rate), ], row.names = FALSE)
  }

  if (apply_to == "tcr") {
    combined2 <- lapply(names(combined), function(sid) {
      df <- combined[[sid]]
      df$barcode <- switch(best$name,
        "identity"        = df$barcode,
        "add_-1"          = add_minus1(df$barcode),
        "remove_-1"       = remove_minus1(df$barcode),
        "prepend_sample_" = prepend_us(df$barcode, sid),
        "append_sample_"  = append_us(df$barcode, sid),
        "prepend-sample"  = prepend_dash(df$barcode, sid),
        "append-sample"   = append_dash(df$barcode, sid),
        df$barcode
      )
      df
    })
    names(combined2) <- names(combined)
    return(list(combined = combined2, transformed = best$name, table = pool))
  } else {
    seu2 <- seu
    new_names <- switch(best$name,
      "seurat_add_-1"    = add_minus1(seu_barcodes),
      "seurat_remove_-1" = remove_minus1(seu_barcodes),
      "seurat_prepend_"  = prepend_us(seu_barcodes, seu_samples_per_cell),
      "seurat_append_"   = append_us(seu_barcodes, seu_samples_per_cell),
      "seurat_prepend-"  = prepend_dash(seu_barcodes, seu_samples_per_cell),
      "seurat_append-"   = append_dash(seu_barcodes, seu_samples_per_cell),
      seu_barcodes
    )
    colnames(seu2) <- new_names
    return(list(seurat = seu2, transformed = best$name, table = pool))
  }
}

subset_tcells_safe <- function(seu, label_col, regex, fallback_ident = TRUE) {
  seu <- repair_seurat_for_subset(seu)

  md <- seu@meta.data

  if (!label_col %in% colnames(md)) {
    if (fallback_ident) {
      seu$seurat_clusters <- as.character(Idents(seu))
      label_col <- "seurat_clusters"
    } else {
      stop("Label column not found: ", label_col)
    }
  }

  labs <- as.character(md[[label_col]])
  keep <- grepl(regex, labs, perl = TRUE)

  if (sum(keep, na.rm = TRUE) == 0) {
    warning("No T-cell subset matched the provided regex. Returning repaired original object.")
    return(seu)
  }

  keep_cells <- rownames(md)[keep]
  subset(seu, cells = keep_cells)
}


safe_percent <- function(x, denom) ifelse(denom > 0, x / denom, NA_real_)

repair_seurat_for_subset <- function(seu) {
  suppressPackageStartupMessages({
    library(Seurat)
    library(SeuratObject)
    library(Matrix)
  })

  `%||%` <- function(a, b) if (!is.null(a)) a else b

  get_any_matrix <- function(obj, assay_name) {
    out <- NULL

    # Try Seurat v4/v5 standard accessors first
    out <- tryCatch(Seurat::GetAssayData(obj, assay = assay_name, slot = "counts"), error = function(e) NULL)
    if (!is.null(out) && nrow(out) > 0 && ncol(out) > 0) return(out)

    out <- tryCatch(Seurat::GetAssayData(obj, assay = assay_name, slot = "data"), error = function(e) NULL)
    if (!is.null(out) && nrow(out) > 0 && ncol(out) > 0) return(out)

    # Try Seurat v5 layers explicitly
    lyr_names <- tryCatch(SeuratObject::Layers(obj[[assay_name]]), error = function(e) NULL)
    if (!is.null(lyr_names) && length(lyr_names) > 0) {
      for (lyr in c("counts", "data", lyr_names)) {
        out <- tryCatch(SeuratObject::LayerData(obj[[assay_name]], layer = lyr), error = function(e) NULL)
        if (!is.null(out) && nrow(out) > 0 && ncol(out) > 0) return(out)
      }
    }

    NULL
  }

  assay_names <- names(seu@assays)
  if (length(assay_names) == 0) {
    stop("No assays found in Seurat object.")
  }

  cat("repair_seurat_for_subset(): assays found ->", paste(assay_names, collapse = ", "), "\n")
  cat("repair_seurat_for_subset(): default assay ->", DefaultAssay(seu), "\n")

  mat <- NULL
  chosen_assay <- NULL

  assay_priority <- unique(c(DefaultAssay(seu), "RNA", assay_names))

  for (aa in assay_priority) {
    if (!aa %in% assay_names) next
    candidate <- get_any_matrix(seu, aa)
    if (!is.null(candidate)) {
      mat <- candidate
      chosen_assay <- aa
      break
    }
  }

  if (is.null(mat)) {
    stop("Could not recover a usable assay matrix to rebuild Seurat object.")
  }

  cat("repair_seurat_for_subset(): using assay ->", chosen_assay, "\n")
  cat("repair_seurat_for_subset(): matrix dim ->", nrow(mat), "x", ncol(mat), "\n")

  # Ensure sparse matrix if possible
  if (!inherits(mat, "dgCMatrix")) {
    mat <- tryCatch(as(mat, "dgCMatrix"), error = function(e) Matrix::Matrix(as.matrix(mat), sparse = TRUE))
  }

  # Rebuild a minimal clean Seurat object
  seu_new <- Seurat::CreateSeuratObject(
    counts = mat,
    assay = chosen_assay,
    meta.data = seu@meta.data[colnames(mat), , drop = FALSE]
  )

  DefaultAssay(seu_new) <- chosen_assay

  # Restore reductions only if cell order is compatible
  red_names <- tryCatch(SeuratObject::Reductions(seu), error = function(e) character())
  if (length(red_names) > 0) {
    for (r in red_names) {
      emb <- tryCatch(Seurat::Embeddings(seu, reduction = r), error = function(e) NULL)
      if (is.null(emb)) next
      common_cells <- intersect(colnames(seu_new), rownames(emb))
      if (length(common_cells) < 2) next

      emb2 <- emb[common_cells, , drop = FALSE]
      seu_new <- subset(seu_new, cells = common_cells)

      dr <- SeuratObject::CreateDimReducObject(
        embeddings = emb2,
        key = paste0(toupper(substr(r, 1, 1)), "_"),
        assay = DefaultAssay(seu_new)
      )
      seu_new[[r]] <- dr
    }
  }

  # Restore identities if possible
  old_ids <- tryCatch(Idents(seu), error = function(e) NULL)
  if (!is.null(old_ids)) {
    old_ids <- as.character(old_ids)
    names(old_ids) <- names(Idents(seu))
    common_id_cells <- intersect(colnames(seu_new), names(old_ids))
    if (length(common_id_cells) > 0) {
      Idents(seu_new) <- factor(old_ids[colnames(seu_new)])
    }
  }

  validObject(seu_new)
  seu_new
}

```

##loading inputs
```{r}
#| label: read-inputs
require_file(params$seurat_rds, sprintf("Seurat RDS not found: %s", params$seurat_rds))
require_file(params$contigs_file, sprintf("Filtered contigs file not found: %s", params$contigs_file))

seu <- readRDS(params$seurat_rds)
contigs_all_post <- safe_read_table(params$contigs_file) %>% as.data.frame()
clonotypes_tbl <- safe_read_table(params$clonotypes_file)
metadata_tbl <- safe_read_table(params$metadata_file)
vdj_qc_summary_tbl <- safe_read_table(params$vdj_qc_summary_file)
```

## standardize-contigs
```{r}
#| label: standardize-contigs
sample_col_ctg <- first_existing_col(contigs_all_post, params$contig_sample_col, params$contig_sample_candidates)
barcode_col_ctg <- first_existing_col(contigs_all_post, params$contig_barcode_col, params$contig_barcode_candidates)
chain_col_ctg <- first_existing_col(contigs_all_post, params$contig_chain_col, params$contig_chain_candidates)
cdr3_col_ctg <- first_existing_col(contigs_all_post, params$contig_cdr3_col, params$contig_cdr3_candidates)
v_col_ctg <- first_existing_col(contigs_all_post, params$contig_v_col, params$contig_v_candidates)
j_col_ctg <- first_existing_col(contigs_all_post, params$contig_j_col, params$contig_j_candidates)
clone_col_ctg <- first_existing_col(contigs_all_post, params$contig_clone_col, params$contig_clone_candidates)

need_cols <- c(sample_col_ctg, barcode_col_ctg, chain_col_ctg, cdr3_col_ctg)
if (any(is.null(need_cols))) stop("Could not resolve required contig columns.")

contigs_all_post <- contigs_all_post %>%
  mutate(
    sample = as.character(.data[[sample_col_ctg]]),
    barcode = as.character(.data[[barcode_col_ctg]]),
    chain = as.character(.data[[chain_col_ctg]]),
    cdr3 = as.character(.data[[cdr3_col_ctg]]),
    v_gene = if (!is.null(v_col_ctg)) as.character(.data[[v_col_ctg]]) else NA_character_,
    j_gene = if (!is.null(j_col_ctg)) as.character(.data[[j_col_ctg]]) else NA_character_,
    raw_clonotype_id = if (!is.null(clone_col_ctg)) as.character(.data[[clone_col_ctg]]) else NA_character_
  )
head(contigs_all_post)

```

## SCrepertoire prepare
```{r}
#| label: prepare-screpertoire
if (isTRUE(params$filter_to_t_ab) && any(contigs_all_post$chain %in% c("TRG", "TRD"))) {
  contigs_all_post <- dplyr::filter(contigs_all_post, chain %in% c("TRA", "TRB"))
}

if (params$cells_mode == "T-GD") {
  contigs_all_post <- dplyr::filter(contigs_all_post, chain %in% c("TRG", "TRD"))
}
if (params$cells_mode == "both") {
  contigs_all_post <- dplyr::filter(contigs_all_post, chain %in% c("TRA", "TRB", "TRG", "TRD"))
}

contig_list <- split(contigs_all_post, contigs_all_post$sample)
contig_list <- contig_list[vapply(contig_list, nrow, FUN.VALUE = integer(1)) > 0]
stopifnot(length(contig_list) > 0)

required_cols_screp <- c("barcode", "chain", "cdr3", "v_gene", "j_gene")
bad <- names(contig_list)[!vapply(contig_list, function(df) all(required_cols_screp %in% names(df)), TRUE)]
if (length(bad)) stop("Missing required columns for scRepertoire in: ", paste(bad, collapse = ", "))

sample_vec <- names(contig_list)

combined <- combineTCR_safe(
  contig_list,
  samples = sample_vec,
  ID = if (params$combine_id == "") NULL else params$combine_id,
  cells = params$cells_mode
)

if (!isTRUE(params$keep_na_clonotypes)) {
  combined <- dropNA_clonotypes(combined)
}

save_rds_safe(combined, "scRepertoire_combined.rds")

comb_cells <- vapply(combined, function(x) length(unique(as.character(x$barcode))), integer(1))
comb_cells_tbl <- data.frame(sample = names(comb_cells), cells = comb_cells, stringsAsFactors = FALSE)
save_table_safe(comb_cells_tbl, "scRepertoire_cells_per_sample.tsv")
```

## barcode-preclean
```{r}
#| label: barcode-preclean
if (isTRUE(params$clean_duplicated_sample_prefix)) {
  for (s in names(combined)) combined[[s]]$barcode <- clean_barcode(as.character(combined[[s]]$barcode))
  contigs_all_post$barcode <- clean_barcode(as.character(contigs_all_post$barcode))
}

seu_md <- seu@meta.data

sample_col_seu <- first_existing_col(seu_md, params$sample_col, params$sample_candidates)
patient_col_seu <- first_existing_col(seu_md, params$patient_col, params$patient_candidates)
condition_col_seu <- first_existing_col(seu_md, params$condition_col, params$condition_candidates)
timepoint_col_seu <- first_existing_col(seu_md, params$timepoint_col, params$timepoint_candidates)
batch_col_seu <- first_existing_col(seu_md, params$batch_col, params$batch_candidates)
label_col_seu <- detect_label_col(
  seu,
  preferred = params$label_col,
  candidates = params$label_candidates,
  use_ident = params$use_ident_if_label_missing
)

if (!"META_SAMPLE" %in% colnames(seu@meta.data) && !is.null(sample_col_seu)) seu$META_SAMPLE <- as.character(seu@meta.data[[sample_col_seu]])
if (!"META_PATIENT" %in% colnames(seu@meta.data) && !is.null(patient_col_seu)) seu$META_PATIENT <- as.character(seu@meta.data[[patient_col_seu]])
if (!"META_TIMECOND" %in% colnames(seu@meta.data) && !is.null(condition_col_seu)) seu$META_TIMECOND <- as.character(seu@meta.data[[condition_col_seu]])
if (!"META_BATCH" %in% colnames(seu@meta.data) && !is.null(batch_col_seu)) seu$META_BATCH <- as.character(seu@meta.data[[batch_col_seu]])
```

## Sample Matching Diagnostic

```{r}
#| label: sample-matching-diagnostic
#| warning: false

# VDJ samples (from filtered contigs)
vdj_samples <- sort(unique(as.character(contigs_all_post$sample)))

# GEX samples (from Seurat metadata)
gex_samples <- if (!is.null(sample_col_seu)) {
  sort(unique(as.character(seu@meta.data[[sample_col_seu]])))
} else {
  sort(unique(as.character(seu@meta.data[["orig.ident"]])))
}

matched_samples  <- intersect(vdj_samples, gex_samples)
vdj_only_samples <- setdiff(vdj_samples, gex_samples)
gex_only_samples <- setdiff(gex_samples, vdj_samples)

# Format helper: collapse to comma-separated string, or "none"
fmt <- function(x) if (length(x) == 0) "none" else paste(x, collapse = ", ")

cat("=== Sample Matching Diagnostic ===\n\n")
cat(sprintf("VDJ samples   (%d): %s\n", length(vdj_samples),  fmt(vdj_samples)))
cat(sprintf("GEX samples   (%d): %s\n", length(gex_samples),  fmt(gex_samples)))
cat("\n")
cat(sprintf("Matched       (%d): %s\n", length(matched_samples),  fmt(matched_samples)))
cat(sprintf("VDJ-only      (%d): %s\n", length(vdj_only_samples), fmt(vdj_only_samples)))
cat(sprintf("GEX-only      (%d): %s\n", length(gex_only_samples), fmt(gex_only_samples)))
cat("\n")

if (length(matched_samples) == 0) {
  stop("No VDJ samples match any GEX sample. Check that sample names are consistent between contigs_after_qc.tsv and the Seurat object.")
}

cat(sprintf(">>> Proceeding with %d matched sample(s): %s\n", length(matched_samples), fmt(matched_samples)))

if (length(vdj_only_samples) > 0) {
  cat(sprintf("\nWARNING: %d VDJ sample(s) have no matching GEX data and will be excluded: %s\n",
              length(vdj_only_samples), fmt(vdj_only_samples)))
}
if (length(gex_only_samples) > 0) {
  cat(sprintf("\nNOTE: %d GEX sample(s) have no VDJ data. TCR metrics will be NA for these cells: %s\n",
              length(gex_only_samples), fmt(gex_only_samples)))
}

# Build a summary table for display
sample_match_tbl <- data.frame(
  sample = c(matched_samples, vdj_only_samples, gex_only_samples),
  has_vdj = c(rep(TRUE,  length(matched_samples)),
              rep(TRUE,  length(vdj_only_samples)),
              rep(FALSE, length(gex_only_samples))),
  has_gex = c(rep(TRUE,  length(matched_samples)),
              rep(FALSE, length(vdj_only_samples)),
              rep(TRUE,  length(gex_only_samples))),
  status  = c(rep("Matched",  length(matched_samples)),
              rep("VDJ only", length(vdj_only_samples)),
              rep("GEX only", length(gex_only_samples))),
  stringsAsFactors = FALSE
)
sample_match_tbl <- sample_match_tbl[order(sample_match_tbl$status, sample_match_tbl$sample), ]
knitr::kable(sample_match_tbl, row.names = FALSE,
             caption = "Sample presence in VDJ and GEX data")
```

## overlap-preharm
```{r}
#| label: overlap-preharm
tcr_barcodes_pre <- unique(unlist(lapply(combined, `[[`, "barcode")))
overlap_pre <- length(intersect(tcr_barcodes_pre, colnames(seu)))
total_pre <- length(tcr_barcodes_pre)
overlap_rate_pre <- ifelse(total_pre > 0, overlap_pre / total_pre, NA_real_)

harmonizer_score <- score_harmonizer_candidates(
  combined,
  seu,
  seu_sample_col = sample_col_seu %||% "orig.ident"
)

save_table_safe(harmonizer_score$table, "barcode_harmonizer_candidates.tsv")
head(harmonizer_score$table)

selected_transform <- harmonizer_score$selected
writeLines(as.character(selected_transform), file.path(params$outdir, params$tables_dir, "barcode_harmonizer_selected.txt"))
head(as.character(selected_transform))
```

## harmonization-apply
```{r}
#| label: harmonization-apply
if (isTRUE(params$barcode_harmonization_required) &&
    length(tcr_barcodes_pre) > 0 &&
    overlap_rate_pre < params$harmonization_overlap_threshold) {

  hz <- harmonize_barcodes(
    combined,
    seu,
    seu_sample_col = sample_col_seu %||% "orig.ident",
    apply_to = params$harmonize_apply_to,
    verbose = FALSE
  )

  if (params$harmonize_apply_to == "tcr") {
    combined <- hz$combined
    selected_transform <- hz$transformed
  } else {
    seu <- hz$seurat
    selected_transform <- hz$transformed
  }

  save_table_safe(hz$table, "barcode_harmonizer_candidates_applied.tsv")
  writeLines(as.character(selected_transform), file.path(params$outdir, params$tables_dir, "barcode_harmonizer_applied.txt"))
}
```

## overlap-postharm
```{r}
#| label: overlap-postharm
tcr_barcodes <- unique(unlist(lapply(combined, `[[`, "barcode")))
seu_cells <- colnames(seu)
final_overlap <- length(intersect(tcr_barcodes, seu_cells))
total_tcr <- length(tcr_barcodes)
final_overlap_rate <- ifelse(total_tcr > 0, final_overlap / total_tcr, NA_real_)

if (length(tcr_barcodes) > 0) {
  stopifnot(final_overlap_rate >= params$minimum_final_overlap_fraction)
}

barcode_overlap_summary <- data.frame(
  stage = c("pre_harmonization", "post_harmonization"),
  tcr_unique_barcodes = c(total_pre, total_tcr),
  overlap_with_seurat = c(overlap_pre, final_overlap),
  overlap_rate = c(overlap_rate_pre, final_overlap_rate),
  stringsAsFactors = FALSE
)
save_table_safe(barcode_overlap_summary, "barcode_overlap_summary.tsv")
```

## overlap-by-sample
```{r}
#| label: overlap-by-sample
barcode_overlap_by_sample <- rbindlist(lapply(names(combined), function(sid) {
  bcs <- unique(as.character(combined[[sid]]$barcode))
  ov <- sum(bcs %in% seu_cells)
  data.table(
    sample = sid,
    tcr_barcodes = length(bcs),
    overlap = ov,
    unmapped = length(bcs) - ov,
    overlap_rate = ifelse(length(bcs) > 0, ov / length(bcs), NA_real_)
  )
}))
save_table_safe(barcode_overlap_by_sample, "barcode_overlap_by_sample.tsv")
```

## meta-key-audit
```{r}
#| label: meta-key-audit
md <- seu@meta.data

meta_key_audit <- data.frame(
  key = c("label", "sample", "patient", "condition", "timepoint", "batch"),
  column = c(
    label_col_seu,
    sample_col_seu,
    patient_col_seu,
    condition_col_seu,
    timepoint_col_seu,
    batch_col_seu
  ),
  n_unique = c(
    if (!is.null(label_col_seu)) length(unique(md[[label_col_seu]])) else NA_integer_,
    if (!is.null(sample_col_seu)) length(unique(md[[sample_col_seu]])) else NA_integer_,
    if (!is.null(patient_col_seu)) length(unique(md[[patient_col_seu]])) else NA_integer_,
    if (!is.null(condition_col_seu)) length(unique(md[[condition_col_seu]])) else NA_integer_,
    if (!is.null(timepoint_col_seu)) length(unique(md[[timepoint_col_seu]])) else NA_integer_,
    if (!is.null(batch_col_seu)) length(unique(md[[batch_col_seu]])) else NA_integer_
  ),
  stringsAsFactors = FALSE
)
save_table_safe(meta_key_audit, "meta_key_choices.tsv")
```

## build-per-cell-tcr
```{r}
#| label: build-per-cell-tcr
dt <- as.data.table(contigs_all_post)

for (cc in c("sample", "barcode", "chain", "cdr3", "v_gene", "j_gene", "raw_clonotype_id")) {
  if (cc %in% names(dt)) set(dt, j = cc, value = as.character(dt[[cc]]))
}

alpha_chains <- c("TRA")
beta_chains <- c("TRB")

alpha_by_cell <- dt[chain %in% alpha_chains & !is.na(cdr3),
  .(cdr3a = cdr3[1L], trav = v_gene[1L], traj = j_gene[1L]),
  by = .(sample, barcode)
]
beta_by_cell <- dt[chain %in% beta_chains & !is.na(cdr3),
  .(cdr3b = cdr3[1L], trbv = v_gene[1L], trbj = j_gene[1L]),
  by = .(sample, barcode)
]

per_cell <- merge(alpha_by_cell, beta_by_cell, by = c("sample", "barcode"), all = TRUE)

per_cell[, CTaa := fifelse(
  !is.na(cdr3a) & !is.na(cdr3b), paste0("A:", cdr3a, "|B:", cdr3b),
  fifelse(!is.na(cdr3b), paste0("B:", cdr3b),
    fifelse(!is.na(cdr3a), paste0("A:", cdr3a), NA_character_)
  )
)]

clone_source <- dt[, .(clone_id_raw = raw_clonotype_id[1L]), by = .(sample, barcode)]
per_cell <- merge(per_cell, clone_source, by = c("sample", "barcode"), all.x = TRUE)
```

## evaluate-merge-strategies
```{r}
#| label: evaluate-merge-strategies
seu_md_dt <- as.data.table(seu@meta.data, keep.rownames = "cell_id")
seu_md_dt[, cell_id := as.character(cell_id)]

if (!"META_SAMPLE" %in% colnames(seu_md_dt)) {
  if (!is.null(sample_col_seu)) {
    seu_md_dt[, META_SAMPLE := as.character(get(sample_col_seu))]
  } else {
    seu_md_dt[, META_SAMPLE := NA_character_]
  }
}

seu_k1 <- seu_md_dt[, .(cell_id, META_SAMPLE, key = cell_id)]
seu_k1[, strategy := "cell_id==barcode"]
tcr_k1 <- per_cell[, .(key = barcode, sample, barcode)]

seu_k2 <- copy(seu_md_dt)
seu_k2[, key := sub("^[^_]+_", "", cell_id)]
seu_k2 <- seu_k2[, .(cell_id, META_SAMPLE, key)]
seu_k2[, strategy := "strip_prefix_before_underscore"]
tcr_k2 <- tcr_k1

seu_k3 <- seu_md_dt[, .(cell_id, META_SAMPLE, key = cell_id)]
seu_k3[, strategy := "cell_id==SAMPLE_BARCODE"]
tcr_k3 <- copy(per_cell)[, key := paste(sample, barcode, sep = "_")]
tcr_k3 <- tcr_k3[, .(key, sample, barcode)]

seu_k4 <- seu_md_dt[, .(cell_id, META_SAMPLE, key = paste(META_SAMPLE, cell_id, sep = "_"))]
seu_k4[, strategy := "META_SAMPLE+cell_id==sample+barcode"]
tcr_k4 <- tcr_k3

eval_strategy <- function(seu_tbl, tcr_tbl) {
  s <- unique(seu_tbl$key)
  t <- unique(tcr_tbl$key)
  length(intersect(s, t))
}

counts <- data.table(
  strategy = c(seu_k1$strategy[1], seu_k2$strategy[1], seu_k3$strategy[1], seu_k4$strategy[1]),
  matches = c(
    eval_strategy(seu_k1, tcr_k1),
    eval_strategy(seu_k2, tcr_k2),
    eval_strategy(seu_k3, tcr_k3),
    eval_strategy(seu_k4, tcr_k4)
  )
)
setorder(counts, -matches)
best <- counts$strategy[1]

idx3 <- which(counts$strategy == "cell_id==SAMPLE_BARCODE")
if (length(idx3) && counts$matches[idx3] >= 0.98 * counts$matches[1] &&
    best != "cell_id==SAMPLE_BARCODE") {
  best <- "cell_id==SAMPLE_BARCODE"
}

pick <- function(s) switch(
  s,
  "cell_id==barcode"                    = list(seu = seu_k1, tcr = tcr_k1),
  "strip_prefix_before_underscore"      = list(seu = seu_k2, tcr = tcr_k2),
  "cell_id==SAMPLE_BARCODE"             = list(seu = seu_k3, tcr = tcr_k3),
  "META_SAMPLE+cell_id==sample+barcode" = list(seu = seu_k4, tcr = tcr_k4)
)

pt <- pick(best)
seu_tbl <- pt$seu
tcr_tbl <- pt$tcr

n_seu_keys <- length(unique(seu_md_dt$cell_id))
n_tcr_keys <- nrow(unique(per_cell[, .(sample, barcode)]))
best_matches <- eval_strategy(seu_tbl, tcr_tbl)
theoretical_max_pct <- 100 * n_tcr_keys / n_seu_keys

diag_tbl <- copy(counts)
diag_tbl[, total_cells := n_seu_keys]
diag_tbl[, total_vdj := n_tcr_keys]
diag_tbl[, pct_cells_matched := round(100 * matches / total_cells, 2)]
diag_tbl[, pct_tcr_matched := round(100 * matches / total_vdj, 2)]

save_table_safe(diag_tbl, "tcr_merge_match_diagnostics.tsv")
```

## per-sample-coverage
```{r}
#| label: per-sample-coverage
matched_keys <- intersect(unique(seu_tbl$key), unique(tcr_tbl$key))

seu_sample_raw <- seu@meta.data$META_SAMPLE %||% seu@meta.data[[sample_col_seu]]
seu_sample_base <- sub("\\.\\d+$", "", as.character(seu_sample_raw))
seu_by_samp <- data.table(sample = seu_sample_base)[, .N, by = sample]
setnames(seu_by_samp, "N", "gex_cells")

tcr_by_samp <- unique(per_cell[, .(sample, barcode)])[, .N, by = sample]
setnames(tcr_by_samp, "N", "tcr_cells")

matched_by_samp <- tcr_tbl[key %in% matched_keys, .N, by = sample]
setnames(matched_by_samp, "N", "matched_keys")

per_sample_cov <- Reduce(function(x, y) merge(x, y, by = "sample", all = TRUE),
                         list(seu_by_samp, tcr_by_samp, matched_by_samp))

for (cc in c("gex_cells", "tcr_cells", "matched_keys")) {
  if (!cc %in% names(per_sample_cov)) per_sample_cov[[cc]] <- 0L
}
per_sample_cov[is.na(gex_cells), gex_cells := 0L]
per_sample_cov[is.na(tcr_cells), tcr_cells := 0L]
per_sample_cov[is.na(matched_keys), matched_keys := 0L]

per_sample_cov[, `:=`(
  gex_cov_pct = ifelse(gex_cells > 0, round(100 * matched_keys / gex_cells, 2), 0),
  tcr_cov_pct = ifelse(tcr_cells > 0, round(100 * matched_keys / tcr_cells, 2), 0)
)]

save_table_safe(per_sample_cov, "tcr_merge_per_sample_coverage.tsv")
```

## final-merge-to-seurat
```{r}
#| label: final-merge-to-seurat
tcr_annot <- merge(
  per_cell,
  tcr_tbl[, .(key, sample, barcode)],
  by = c("sample", "barcode"),
  all.y = TRUE
)

bring <- c("CTaa", "cdr3a", "cdr3b", "trav", "traj", "trbv", "trbj", 
           "clone_id_raw", "patient_id", "condition", "timepoint")

for (col in bring) if (!col %in% names(tcr_annot)) tcr_annot[[col]] <- NA_character_

seu_join <- merge(
  seu_tbl[, .(cell_id, META_SAMPLE, key)],
  tcr_annot[, c("key", bring), with = FALSE],
  by = "key",
  all.x = TRUE
)

setkey(seu_join, cell_id)
md_dt <- as.data.table(seu@meta.data, keep.rownames = "cell_id")
setkey(md_dt, cell_id)
md_dt <- seu_join[md_dt]

# Map metadata to internal keys
if (has_col(md_dt, "patient_id")) md_dt[, META_PATIENT := coalesce(as.character(META_PATIENT), as.character(patient_id))]
if (has_col(md_dt, "condition"))  md_dt[, META_TIMECOND := coalesce(as.character(META_TIMECOND), as.character(condition))]
if (has_col(md_dt, "timepoint"))  md_dt[, META_TIMECOND := coalesce(as.character(META_TIMECOND), as.character(timepoint))]

md_dt[, clone_id := CTaa]
md_dt[, has_tcr := !is.na(clone_id)]
md_dt[, clone_size := ifelse(is.na(clone_id), 0L, .N), by = clone_id]
md_dt[, paired_tcr := !is.na(cdr3a) & !is.na(cdr3b)]

# ROBUST BINNING: Always create clone_size_bin
md_dt[, clone_size_bin := NA_character_]
if (any(md_dt$clone_size > 0)) {
  try({
    qs <- unique(quantile(md_dt$clone_size[md_dt$clone_size > 0], probs = params$clone_bins_quantiles, na.rm = TRUE))
    if (length(qs) >= 2) {
      md_dt[clone_size > 0, clone_size_bin := as.character(cut(clone_size, breaks = qs, include.lowest = TRUE))]
    } else {
      md_dt[clone_size > 0, clone_size_bin := as.character(clone_size)]
    }
  }, silent = TRUE)
}

rn <- md_dt$cell_id
md_dt$cell_id <- NULL
seu@meta.data <- as.data.frame(md_dt)
rownames(seu@meta.data) <- rn

save_rds_safe(seu, "seurat_with_TCR_integrated.rds")

```

## screpertoire-merge
```{r}
#| label: screpertoire-merge-optional

screp_combined_expression_ok <- FALSE
seu_screp <- seu
screp_error_msg <- NA_character_
screp_note <- NA_character_

message("--- Attempting optional scRepertoire::combineExpression integration ---")


rebuild_clean_seurat_for_screp <- function(obj) {
  suppressPackageStartupMessages({
    library(Seurat)
    library(SeuratObject)
    library(Matrix)
  })

  `%||%` <- function(a, b) if (!is.null(a)) a else b

  get_any_matrix <- function(x, assay_name) {
    out <- tryCatch(Seurat::GetAssayData(x, assay = assay_name, slot = "counts"), error = function(e) NULL)
    if (!is.null(out) && nrow(out) > 0 && ncol(out) > 0) return(out)

    out <- tryCatch(Seurat::GetAssayData(x, assay = assay_name, slot = "data"), error = function(e) NULL)
    if (!is.null(out) && nrow(out) > 0 && ncol(out) > 0) return(out)

    lyr_names <- tryCatch(SeuratObject::Layers(x[[assay_name]]), error = function(e) NULL)
    if (!is.null(lyr_names) && length(lyr_names) > 0) {
      for (lyr in c("counts", "data", lyr_names)) {
        out <- tryCatch(SeuratObject::LayerData(x[[assay_name]], layer = lyr), error = function(e) NULL)
        if (!is.null(out) && nrow(out) > 0 && ncol(out) > 0) return(out)
      }
    }

    NULL
  }

  assay_names <- names(obj@assays)
  assay_priority <- unique(c(DefaultAssay(obj), "RNA", assay_names))

  mat <- NULL
  chosen_assay <- NULL

  for (aa in assay_priority) {
    if (!aa %in% assay_names) next
    candidate <- get_any_matrix(obj, aa)
    if (!is.null(candidate)) {
      mat <- candidate
      chosen_assay <- aa
      break
    }
  }

  if (is.null(mat)) {
    stop("Could not recover a usable assay matrix for scRepertoire repair.")
  }

  if (!inherits(mat, "dgCMatrix")) {
    mat <- tryCatch(as(mat, "dgCMatrix"), error = function(e) Matrix::Matrix(as.matrix(mat), sparse = TRUE))
  }

  md <- obj@meta.data[colnames(mat), , drop = FALSE]

  seu_new <- Seurat::CreateSeuratObject(
    counts = mat,
    assay = chosen_assay,
    meta.data = md
  )

  DefaultAssay(seu_new) <- chosen_assay

  old_ids <- tryCatch(Idents(obj), error = function(e) NULL)
  if (!is.null(old_ids)) {
    old_ids <- as.character(old_ids)
    names(old_ids) <- names(Idents(obj))
    ids_vec <- old_ids[colnames(seu_new)]
    names(ids_vec) <- colnames(seu_new)
    Idents(seu_new) <- factor(ids_vec)
  }

  validObject(seu_new)
  seu_new
}

combineExpression_safe <- function(combined_obj, seu_obj, clone_call = "aa", add_label = FALSE) {
  fun <- get("combineExpression", asNamespace("scRepertoire"))
  fml <- names(formals(fun))

  args <- list(
    combined_obj,
    seu_obj
  )

  if ("cloneCall" %in% fml)  args$cloneCall  <- clone_call
  if ("proportion" %in% fml) args$proportion <- FALSE
  if ("addLabel" %in% fml)   args$addLabel   <- add_label
  if ("filterMode" %in% fml) args$filterMode <- "union"
  if ("cloneSize" %in% fml) {
    args$cloneSize <- c(Single = 1, Small = 5, Medium = 20, Large = 100, Hyperexpanded = 500)
  }

  do.call(fun, args)
}

tryCatch({

  combined_harmonized <- lapply(names(combined), function(sid) {
    df <- combined[[sid]]

    if (best == "cell_id==SAMPLE_BARCODE") {
      df$barcode <- paste(sid, df$barcode, sep = "_")
    } else if (best == "META_SAMPLE+cell_id==sample+barcode") {
      df$barcode <- df$barcode
    } else if (best == "strip_prefix_before_underscore") {
      df$barcode <- df$barcode
    } else {
      df$barcode <- df$barcode
    }

    df$barcode <- as.character(df$barcode)
    df
  })
  names(combined_harmonized) <- names(combined)

  # rebuild minimal valid Seurat object
  seu_screp <- rebuild_clean_seurat_for_screp(seu_screp)

  if (inherits(seu_screp[[DefaultAssay(seu_screp)]], "Assay5")) {
    seu_screp <- JoinLayers(seu_screp)
  }

  # IMPORTANT: rename Seurat cells to exact barcode namespace expected by scRepertoire
  all_tcr_barcodes <- unique(unlist(lapply(combined_harmonized, function(x) as.character(x$barcode))))
  seurat_cells <- colnames(seu_screp)

  make_target_names <- function(cells, sample_vec, strategy) {
    if (strategy == "cell_id==SAMPLE_BARCODE") {
      return(paste(sample_vec, cells, sep = "_"))
    } else if (strategy == "META_SAMPLE+cell_id==sample+barcode") {
      return(paste(sample_vec, cells, sep = "_"))
    } else {
      return(cells)
    }
  }

  sample_for_cells <- NULL
  if ("META_SAMPLE" %in% colnames(seu_screp@meta.data)) {
    sample_for_cells <- as.character(seu_screp@meta.data$META_SAMPLE)
  } else if (!is.null(sample_col_seu) && sample_col_seu %in% colnames(seu_screp@meta.data)) {
    sample_for_cells <- as.character(seu_screp@meta.data[[sample_col_seu]])
  } else {
    sample_for_cells <- rep(NA_character_, ncol(seu_screp))
  }

  candidate_names <- make_target_names(seurat_cells, sample_for_cells, best)

  overlap_now <- sum(candidate_names %in% all_tcr_barcodes)
  overlap_identity <- sum(seurat_cells %in% all_tcr_barcodes)

  if (overlap_now > overlap_identity) {
    colnames(seu_screp) <- candidate_names
  }

  # keep only exact overlap cells to avoid scRepertoire metadata write-back issues
  common_cells <- intersect(colnames(seu_screp), all_tcr_barcodes)
  if (length(common_cells) == 0) {
    stop("No overlapping cell names between Seurat object and combined_harmonized barcodes after harmonization.")
  }

  seu_screp <- subset(seu_screp, cells = common_cells)

  # final identity repair after subset
  ids_vec <- as.character(Idents(seu_screp))
  names(ids_vec) <- colnames(seu_screp)
  Idents(seu_screp) <- factor(ids_vec)

  validObject(seu_screp)

  seu_screp <- combineExpression_safe(
    combined_harmonized,
    seu_screp,
    clone_call = switch(
      params$clone_call_preference,
      aa = "aa",
      nt = "nt",
      strict = "strict",
      gene = "gene",
      "aa"
    ),
    add_label = FALSE
  )

  added_cols <- intersect(
    c("CTaa", "CTnt", "CTstrict", "CTgene", "strict", "aa", "nt", "gene"),
    colnames(seu_screp@meta.data)
  )

  if (length(added_cols) > 0) {
    screp_combined_expression_ok <- TRUE
    screp_note <- paste("combineExpression succeeded; detected columns:", paste(added_cols, collapse = ", "))
    message("SUCCESS: scRepertoire integration complete.")
    save_rds_safe(seu_screp, "seurat_with_scRepertoire_combineExpression.rds")
  } else {
    screp_error_msg <- "combineExpression returned without error, but no expected scRepertoire clonotype columns were added to Seurat metadata."
    screp_note <- "Optional scRepertoire path returned no usable metadata columns."
    message("WARNING: scRepertoire ran but did not add expected metadata columns.")
  }

}, error = function(e) {
  screp_error_msg <<- conditionMessage(e)
  screp_note <<- "Optional scRepertoire path failed; validated fast-merge outputs remain available."
  message("scRepertoire Error: ", screp_error_msg)
})

screp_diag_tbl <- tibble::tibble(
  metric = c(
    "scRepertoire combineExpression status",
    "selected_merge_strategy",
    "selected_barcode_transform",
    "clone_call_preference",
    "tcr_barcodes_post_harmonization",
    "seurat_cells_before_subset",
    "final_overlap",
    "final_overlap_rate",
    "scRepertoire_note",
    "scRepertoire_error"
  ),
  value = c(
    ifelse(screp_combined_expression_ok, "SUCCESS", "FAILED_OR_EMPTY"),
    as.character(best),
    as.character(selected_transform),
    as.character(params$clone_call_preference),
    as.character(length(tcr_barcodes)),
    as.character(ncol(seu)),
    as.character(final_overlap),
    sprintf("%.4f", final_overlap_rate),
    as.character(screp_note %||% ""),
    as.character(screp_error_msg %||% "")
  )
)

save_table_safe(screp_diag_tbl, "scRepertoire_combineExpression_diagnostics.tsv")

if (isTRUE(screp_combined_expression_ok)) {
  cat("scRepertoire::combineExpression completed successfully.\n")
} else {
  cat("scRepertoire::combineExpression did not produce usable output.\n")
  if (!is.na(screp_error_msg) && nzchar(screp_error_msg)) {
    cat("Reason:\n")
    cat(screp_error_msg, "\n")
  }
  cat("The pipeline continued using the validated fast-merge TCR↔GEX integration outputs.\n")
}
```


## subset-tcells
```{r}
#| label: subset-tcells


seu_t <- repair_seurat_for_subset(seu)

if (isTRUE(params$subset_tcells)) {
  seu_t <- subset_tcells_safe(
    seu_t,
    label_col = label_col_seu,
    regex = params$tcell_regex,
    fallback_ident = params$use_ident_if_label_missing
  )
}

reduction_to_use <- params$reduction_use
if (!(reduction_to_use %in% names(seu_t@reductions)) && isTRUE(params$make_umap_if_missing)) {
  seu_t <- safe_make_umap(
    seu_t,
    reduction_name = params$reduction_use,
    dims = 1:params$umap_dims_max,
    nfeatures = params$umap_nfeatures
  )
}
if (!(reduction_to_use %in% names(seu_t@reductions))) {
  reduction_to_use <- if ("umap" %in% names(seu_t@reductions)) "umap" else if ("tsne" %in% names(seu_t@reductions)) "tsne" else if ("pca" %in% names(seu_t@reductions)) "pca" else stop("No usable embedding found.")
}

```

### tcell-clone-size
```{r}
#| label: tcell-clone-size
if (!"clone_size" %in% colnames(seu_t@meta.data)) {
  seu_t$clone_size <- 0L
} else {
  seu_t$clone_size[is.na(seu_t$clone_size)] <- 0L
}

save_rds_safe(seu_t, "seurat_tcells_with_TCR.rds")
```

## tcell-summaries
```{r}
#| label: tcell-summaries
sample_col_export <- choose_col(seu_t@meta.data, "META_SAMPLE", sample_col_seu)
patient_col_export <- choose_col(seu_t@meta.data, "META_PATIENT", patient_col_seu)
condition_col_export <- choose_col(seu_t@meta.data, "META_TIMECOND", condition_col_seu)
timepoint_col_export <- choose_col(seu_t@meta.data, "META_TIMECOND", timepoint_col_seu)

umap_df <- as.data.frame(Embeddings(seu_t, reduction_to_use)) %>%
  tibble::rownames_to_column("cell_id")

export_cells <- seu_t@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  dplyr::transmute(
    cell_id,
    sample = if (!is.null(sample_col_export)) .data[[sample_col_export]] else NA_character_,
    patient = if (!is.null(patient_col_export)) .data[[patient_col_export]] else NA_character_,
    condition = if (!is.null(condition_col_export)) .data[[condition_col_export]] else NA_character_,
    annot = .data[[label_col_seu]],
    CTaa = dplyr::coalesce(.data[["CTaa"]], NA_character_),
    clone_id = dplyr::coalesce(.data[["clone_id"]], NA_character_),
    clone_size = as.integer(.data[["clone_size"]]),
    clone_size_bin = .data[["clone_size_bin"]],
    paired_tcr = .data[["paired_tcr"]],
    has_tcr = .data[["has_tcr"]]
  ) %>%
  left_join(umap_df, by = "cell_id")

extra_cols <- intersect(c("cdr3a", "cdr3b", "trav", "trbv", "traj", "trbj"), colnames(seu_t@meta.data))
if (length(extra_cols) > 0) {
  export_cells <- export_cells %>%
    left_join(
      seu_t@meta.data %>%
        tibble::rownames_to_column("cell_id") %>%
        dplyr::select(dplyr::all_of(c("cell_id", extra_cols))),
      by = "cell_id"
    )
}
save_table_safe(export_cells, "tcr_export_cells_with_embedding.tsv")

tcell_per_annotation <- export_cells %>%
  count(annot, has_tcr, paired_tcr, name = "n_cells") %>%
  group_by(annot) %>%
  mutate(frac = n_cells / sum(n_cells)) %>%
  ungroup()
save_table_safe(tcell_per_annotation, "tcell_per_annotation_summary.tsv")

top_clones_tbl <- export_cells %>%
  filter(!is.na(clone_id), clone_id != "", clone_size >= params$min_clone_size_plot) %>%
  count(sample, annot, clone_id, clone_size, name = "n_cells") %>%
  arrange(desc(clone_size), desc(n_cells)) %>%
  slice_head(n = params$top_n_clone_table)
save_table_safe(top_clones_tbl, "tcell_top_clones.tsv")

clone_rank_tbl <- export_cells %>%
  filter(!is.na(clone_id), clone_id != "", clone_size >= params$min_clone_size_plot) %>%
  count(sample, clone_id, name = "n_cells") %>%
  group_by(sample) %>%
  arrange(desc(n_cells), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()
save_table_safe(clone_rank_tbl, "tcell_clone_rank_abundance.tsv")

clone_state_tbl <- export_cells %>%
  filter(!is.na(clone_id), clone_id != "") %>%
  count(clone_id, annot, name = "n_cells")
save_table_safe(clone_state_tbl, "tcell_clone_state_occupancy.tsv")

summary_rollup <- tibble(
  metric = c(
    "TCR barcodes pre-harmonization",
    "Pre-harmonization overlap",
    "Post-harmonization overlap",
    "Final overlap rate",
    "Selected merge strategy",
    "T-cell subset cells",
    "TCR-positive T cells",
    "Paired TCR T cells"
  ),
  value = c(
    total_pre,
    overlap_pre,
    final_overlap,
    sprintf("%.2f%%", 100 * final_overlap_rate),
    best,
    ncol(seu_t),
    sum(seu_t$has_tcr, na.rm = TRUE),
    sum(seu_t$paired_tcr, na.rm = TRUE)
  )
)
save_table_safe(summary_rollup, "tcell_summary_rollup.tsv")
```

## overview
```{r}
#| label: overview
overview_tbl <- tibble(
  Metric = c(
    "Report label",
    "Input Seurat cells",
    "Input filtered contigs",
    "Samples represented in VDJ",
    "Annotation column used",
    "Sample column used",
    "Selected harmonization transform",
    "Selected merge strategy",
    "Final barcode overlap rate",
    "T-cell subset cells",
    "TCR-positive T cells",
    "Paired TCR T cells"
  ),
  Value = c(
    params$report_label,
    comma(ncol(seu)),
    comma(nrow(contigs_all_post)),
    comma(length(unique(contigs_all_post$sample))),
    label_col_seu,
    sample_col_seu %||% "NA",
    selected_transform,
    best,
    sprintf("%.2f%%", 100 * final_overlap_rate),
    comma(ncol(seu_t)),
    comma(sum(seu_t$has_tcr, na.rm = TRUE)),
    comma(sum(seu_t$paired_tcr, na.rm = TRUE))
  )
)

kable(overview_tbl, caption = "High-level overview of GEX–TCR integration.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

## harmonization-figures
```{r}
#| label: harmonization-figures
if (isTRUE(params$show_barcode_harmonization)) {
  p_overlap <- ggplot(barcode_overlap_by_sample,
                      aes(x = reorder(sample, overlap_rate), y = overlap_rate)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "TCR to Seurat Barcode Overlap by Sample",
      subtitle = "Post-harmonization overlap rate.",
      x = NULL,
      y = "Overlap rate"
    ) +
    theme_scratch_pub(params$base_size)

  p_map_stack <- barcode_overlap_by_sample %>%
    dplyr::select(sample, overlap, unmapped) %>%
    tidyr::pivot_longer(cols = c(overlap, unmapped), names_to = "status", values_to = "n") %>%
    mutate(status = recode(
      status,
      overlap = "Mapped to Seurat",
      unmapped = "Not found in Seurat"
    )) %>%
    ggplot(aes(x = reorder(sample, n), y = n, fill = status)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Mapped vs Unmapped TCR Barcodes",
      x = NULL,
      y = "TCR barcodes",
      fill = NULL
    ) +
    theme_scratch_pub(params$base_size)

  p_scatter <- ggplot(barcode_overlap_by_sample,
                      aes(x = tcr_barcodes, y = overlap_rate)) +
    geom_point(size = 2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "Overlap Rate vs TCR Barcode Count",
      x = "Unique TCR barcodes per sample",
      y = "Overlap rate"
    ) +
    theme_scratch_pub(params$base_size)

  (p_overlap | p_map_stack) / p_scatter
  save_plot_safe(p_overlap, glue("barcode_overlap_by_sample.{params$figure_format}"))
  save_plot_safe(p_map_stack, glue("barcode_mapping_mapped_unmapped.{params$figure_format}"))
  save_plot_safe(p_scatter, glue("barcode_overlap_vs_tcr_count.{params$figure_format}"))
}
```

## umap-annotation
```{r}
#| label: umap-annotation


# 1. Get the raw labels and handle NAs immediately
raw_labels <- as.character(seu_t@meta.data[[label_col_seu]])
raw_labels[is.na(raw_labels) | raw_labels == ""] <- "Unannotated"

# 2. Create the unique levels (sorted)
annot_levels <- sort(unique(raw_labels))

# 3. Create a clean numeric mapping
annot_map <- setNames(seq_along(annot_levels), annot_levels)

# 4. Map the labels to numbers and assign to Seurat using AddMetaData (safer than $)
plot_numbers <- annot_map[raw_labels]
names(plot_numbers) <- rownames(seu_t@meta.data) # CRITICAL: Align barcodes

seu_t <- AddMetaData(seu_t, metadata = plot_numbers, col.name = "plot_number")

# 5. Build the legend text (e.g., "1: CD4 T cells")
legend_labels <- paste0(seq_along(annot_levels), ": ", annot_levels)

p_umap_annot <- DimPlot(
  seu_t,
  reduction = reduction_to_use,
  group.by = "plot_number",
  label = TRUE,
  label.size = 6,
  repel = TRUE,
  raster = isTRUE(params$raster_large_umap)
) +
  ggtitle("T-cell embedding: Numbered Annotations") +
  theme_scratch_pub(params$base_size) +
  # Map the numbers back to the long strings in the legend
  scale_color_discrete(labels = legend_labels) +
  guides(color = guide_legend(ncol = 1, override.aes = list(size = 4)))

p_umap_annot
save_plot_safe(p_umap_annot, glue("umap_tcells_annotation_numbered.{params$figure_format}"), width = 12, height = 8)
# p_umap_annot <- DimPlot(
#   seu_t,
#   reduction = reduction_to_use,
#   group.by = label_col_seu,
#   label = params$label_clusters,
#   repel = TRUE,
#   raster = isTRUE(params$raster_large_umap)
# ) +
#   ggtitle("T-cell embedding: annotation") +
#   theme_scratch_pub(params$base_size)
# 
# p_umap_annot
# save_plot_safe(p_umap_annot, glue("umap_tcells_annotation.{params$figure_format}"), width = 8, height = 6)
```

## umap-tcr-positive
```{r}
#| label: umap-tcr-positive
p_umap_tcr <- DimPlot(
  seu_t,
  reduction = reduction_to_use,
  group.by = "has_tcr",
  raster = isTRUE(params$raster_large_umap)
) +
  ggtitle("T-cell embedding: TCR-positive vs TCR-negative") +
  theme_scratch_pub(params$base_size)

p_umap_tcr
save_plot_safe(p_umap_tcr, glue("umap_tcells_has_tcr.{params$figure_format}"), width = 8, height = 6)
```

## umap-paired
```{r}
#| label: umap-paired
p_umap_paired <- DimPlot(
  seu_t,
  reduction = reduction_to_use,
  group.by = "paired_tcr",
  raster = isTRUE(params$raster_large_umap)
) +
  ggtitle("T-cell embedding: paired TCR status") +
  theme_scratch_pub(params$base_size)

p_umap_paired
save_plot_safe(p_umap_paired, glue("umap_tcells_paired_tcr.{params$figure_format}"), width = 8, height = 6)
```

## umap-clone-size
```{r}
#| label: umap-clone-size
p_umap_clone <- FeaturePlot(
  seu_t,
  reduction = reduction_to_use,
  features = "clone_size",
  raster = isTRUE(params$raster_large_umap)
) +
  ggtitle("T-cell embedding: clone size") +
  theme_scratch_pub(params$base_size)

p_umap_clone
save_plot_safe(p_umap_clone, glue("umap_tcells_clone_size.{params$figure_format}"), width = 8, height = 6)
```

## pairing-summary
```{r}
#| label: pairing-summary

pairing_data <- export_cells %>%
    mutate(pairing_group = case_when(
      paired_tcr ~ "Paired",
      has_tcr ~ "Unpaired",
      TRUE ~ "No TCR"
    )) %>%
    count(annot, pairing_group, name = "n_cells")

# Always print the table first so something shows up in the report
kable(pairing_data, caption = "Pairing Status Table") %>% kable_styling()

if (isTRUE(params$show_pairing_summary)) {
  p_pairing <- pairing_data %>%
    group_by(annot) %>%
    mutate(frac = n_cells / sum(n_cells)) %>%
    ggplot(aes(x = fct_reorder(annot, frac, .fun = max), y = frac, fill = pairing_group)) +
    geom_col(position = "fill") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    theme_scratch_pub(params$base_size)
  print(p_pairing)
}

# if (isTRUE(params$show_pairing_summary)) {
#   p_pairing <- export_cells %>%
#     mutate(pairing_group = case_when(
#       paired_tcr ~ "Paired",
#       has_tcr ~ "Unpaired",
#       TRUE ~ "No TCR"
#     )) %>%
#     count(annot, pairing_group, name = "n_cells") %>%
#     group_by(annot) %>%
#     mutate(frac = n_cells / sum(n_cells)) %>%
#     ungroup() %>%
#     ggplot(aes(x = fct_reorder(annot, frac, .fun = max), y = frac, fill = pairing_group)) +
#     geom_col(position = "fill") +
#     coord_flip() +
#     scale_y_continuous(labels = percent_format(accuracy = 1)) +
#     labs(
#       title = "Pairing Status Across T-cell States",
#       x = NULL,
#       y = "Fraction of cells",
#       fill = "TCR status"
#     ) +
#     theme_scratch_pub(params$base_size)
# 
#   p_pairing
#   save_plot_safe(p_pairing, glue("pairing_status_by_annotation.{params$figure_format}"), width = 10, height = 7)
# }
```

## Clonal Homeostasis (Expansion)
```{r}
#| label: clonal-expansion-homeostasis
#| fig-width: 10
#| fig-height: 7

homeo_df <- export_cells %>%
  filter(!is.na(sample), sample != "", !is.na(clone_id), clone_id != "", !is.na(clone_size)) %>%
  distinct(sample, clone_id, clone_size) %>%
  mutate(
    Expansion_Class = cut(
      clone_size,
      breaks = c(0, 1, 5, 20, 100, Inf),
      labels = c("Rare (1)", "Small (2-5)", "Medium (6-20)", "Large (21-100)", "Hyperexpanded (>100)"),
      right = TRUE
    )
  ) %>%
  count(sample, Expansion_Class, name = "n_clones") %>%
  tidyr::complete(sample, Expansion_Class, fill = list(n_clones = 0)) %>%
  group_by(sample) %>%
  mutate(fraction = n_clones / sum(n_clones)) %>%
  ungroup()

sample_order_df <- homeo_df %>%
  select(sample, Expansion_Class, fraction) %>%
  distinct() %>%
  tidyr::pivot_wider(
    names_from = Expansion_Class,
    values_from = fraction,
    values_fill = 0
  ) %>%
  distinct(sample, .keep_all = TRUE) %>%
  arrange(
    desc(`Hyperexpanded (>100)`),
    desc(`Large (21-100)`),
    desc(`Medium (6-20)`),
    desc(`Small (2-5)`),
    desc(`Rare (1)`)
  )

sample_levels <- unique(as.character(sample_order_df$sample))

homeo_df <- homeo_df %>%
  mutate(
    sample = factor(as.character(sample), levels = sample_levels),
    Expansion_Class = factor(
      Expansion_Class,
      levels = c("Rare (1)", "Small (2-5)", "Medium (6-20)", "Large (21-100)", "Hyperexpanded (>100)")
    )
  )

p_homeo <- ggplot(homeo_df, aes(x = sample, y = fraction, fill = Expansion_Class)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c(
      "Rare (1)" = "#4575b4",
      "Small (2-5)" = "#91bfdb",
      "Medium (6-20)" = "#fee090",
      "Large (21-100)" = "#fc8d59",
      "Hyperexpanded (>100)" = "#d73027"
    )
  ) +
  theme_scratch_pub(params$base_size) +
  labs(
    title = "Clonal Expansion Homeostasis",
    y = "Relative Abundance",
    x = "Sample",
    fill = "Expansion_Class"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p_homeo)

save_plot_safe(
  p_homeo,
  glue("clonal_homeostasis_expansion.{params$figure_format}"),
  width = 10,
  height = 7
)
```


## top-clone-overlays
```{r}
#| label: top-clone-overlays
if (isTRUE(params$show_top_clone_umaps)) {
  top_clone_ids <- export_cells %>%
    filter(!is.na(clone_id), clone_id != "") %>%
    count(clone_id, name = "n_cells") %>%
    arrange(desc(n_cells)) %>%
    slice_head(n = params$top_n_clones_umap) %>%
    pull(clone_id)

  if (length(top_clone_ids) > 0) {
    seu_t$top_clone_label <- ifelse(seu_t$clone_id %in% top_clone_ids, seu_t$clone_id, "Other")
    p_top_clones <- DimPlot(
      seu_t,
      reduction = reduction_to_use,
      group.by = "top_clone_label",
      raster = isTRUE(params$raster_large_umap)
    ) +
      ggtitle("Top expanded clonotypes on embedding") +
      theme_scratch_pub(params$base_size)

    p_top_clones
    save_plot_safe(p_top_clones, glue("umap_top_clones.{params$figure_format}"), width = 10, height = 8)
  }
}
```

## clone-rank-abundance
```{r}
#| label: clone-rank-abundance
if (isTRUE(params$show_clone_rank_plot) && nrow(clone_rank_tbl) > 0) {
  p_clone_rank <- ggplot(clone_rank_tbl, aes(x = rank, y = n_cells, color = sample)) +
    geom_line(linewidth = 0.9, alpha = 0.8) +
    geom_point(size = 1.2, alpha = 0.9) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
      title = "Clone Rank-Abundance Curve",
      subtitle = glue("Only clones with at least {params$min_clone_size_plot} cells are shown."),
      x = "Clone rank (log10)",
      y = "Cells per clone (log10)",
      color = "Sample"
    ) +
    theme_scratch_pub(params$base_size)

  p_clone_rank
  save_plot_safe(p_clone_rank, glue("clone_rank_abundance.{params$figure_format}"))
}
```

## clone-state-heatmap
```{r}
#| label: clone-state-heatmap
#| fig-width: 12
#| fig-height: 8

if (isTRUE(params$show_clone_state_heatmap) && exists("clone_state_tbl") && nrow(clone_state_tbl) > 0) {
  
  # 1. Identify the top 30 most expanded clones to keep the heatmap readable
  top_state_clones <- clone_state_tbl %>%
    group_by(clone_id) %>%
    summarise(total = sum(n_cells), .groups = "drop") %>%
    arrange(desc(total)) %>%
    slice_head(n = 30) %>%
    pull(clone_id)

  # 2. Pivot data into a wide matrix format
  clone_state_mat <- clone_state_tbl %>%
    filter(clone_id %in% top_state_clones) %>%
    tidyr::pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
    as.data.frame()

  # 3. Set row names and convert to pure numeric matrix
  rownames(clone_state_mat) <- clone_state_mat$clone_id
  clone_state_mat$clone_id <- NULL
  clone_state_mat <- as.matrix(clone_state_mat)

  # 4. Define a clean color scale
  col_fun <- colorRamp2(
    c(0, max(clone_state_mat, na.rm = TRUE) / 2, max(clone_state_mat, na.rm = TRUE)), 
    c("white", "gold", "firebrick")
  )

  # 5. Generate the Heatmap with optimized label sizes
  ht_clone_state <- Heatmap(
    clone_state_mat,
    name = "Cells",
    col = col_fun,
    cluster_rows = TRUE, 
    cluster_columns = TRUE,
    show_row_dend = TRUE,
    show_column_dend = TRUE,
    
    # Visual Polish for Long Labels
    row_names_gp = gpar(fontsize = 7),       # Smaller font for Clone IDs
    column_names_gp = gpar(fontsize = 7),    # Smaller font for ScType labels
    column_names_rot = 45,                   # Rotate labels to prevent overlap
    row_names_side = "left",
    
    column_title = "Clone Occupancy Across Annotated T-cell States",
    column_title_gp = gpar(fontsize = 12, fontface = "bold"),
    
    heatmap_legend_param = list(
      title = "Cell Count",
      at = c(0, max(clone_state_mat)),
      labels = c("0", "Max")
    )
  )

  # 6. Render the plot
  draw(ht_clone_state, heatmap_legend_side = "right")
  
} else {
  cat("Clone-state heatmap skipped: insufficient data or disabled in params.")
}
# if (isTRUE(params$show_clone_state_heatmap) && nrow(clone_state_tbl) > 0) {
#   top_state_clones <- clone_state_tbl %>%
#     group_by(clone_id) %>%
#     summarise(total = sum(n_cells), .groups = "drop") %>%
#     arrange(desc(total)) %>%
#     slice_head(n = 30) %>%
#     pull(clone_id)
# 
#   clone_state_mat <- clone_state_tbl %>%
#     filter(clone_id %in% top_state_clones) %>%
#     tidyr::pivot_wider(names_from = annot, values_from = n_cells, values_fill = 0) %>%
#     as.data.frame()
# 
#   rownames(clone_state_mat) <- clone_state_mat$clone_id
#   clone_state_mat$clone_id <- NULL
#   clone_state_mat <- as.matrix(clone_state_mat)
# 
#   ht_clone_state <- Heatmap(
#     clone_state_mat,
#     name = "Cells",
#     col = colorRamp2(
#       c(0, max(clone_state_mat, na.rm = TRUE) / 2, max(clone_state_mat, na.rm = TRUE)),
#       c("white", "gold", "firebrick")
#     ),
#     cluster_rows = TRUE,
#     cluster_columns = TRUE,
#     row_names_side = "left",
#     column_title = "Clone occupancy across annotated T-cell states",
#     heatmap_legend_param = list(title = "Cells")
#   )
# 
#   draw(ht_clone_state)
# }
```

## screpertoire-figures
```{r}
#| label: screpertoire-figures

# Diagnostic check
if (!exists("seu_screp")) {
  message("!!! DIAGNOSTIC: seu_screp object does not exist. Check the merge step.")
} else if (!"strict" %in% colnames(seu_screp@meta.data)) {
  message("!!! DIAGNOSTIC: 'strict' column (clonotype) not found in metadata. scRepertoire plots may fail.")
}

if (isTRUE(params$show_homeostasis) && exists("export_cells")) {
  cat("### Clonal Homeostasis\n")

  clonal_group_levels <- c(
    "Rare (0 < X <= 1e-04)",
    "Small (1e-04 < X <= 0.001)",
    "Medium (0.001 < X <= 0.01)",
    "Large (0.01 < X <= 0.1)",
    "Hyperexpanded (0.1 < X <= 1)"
  )

  homeo_rel_df <- export_cells %>%
    filter(!is.na(sample), sample != "", !is.na(clone_id), clone_id != "") %>%
    count(sample, clone_id, name = "n_cells") %>%
    group_by(sample) %>%
    mutate(clone_fraction = n_cells / sum(n_cells)) %>%
    ungroup() %>%
    mutate(
      Clonal_Group = cut(
        clone_fraction,
        breaks = c(0, 1e-4, 1e-3, 1e-2, 1e-1, 1),
        labels = clonal_group_levels,
        include.lowest = TRUE,
        right = TRUE
      )
    ) %>%
    group_by(sample, Clonal_Group) %>%
    summarise(relative_abundance = sum(clone_fraction, na.rm = TRUE), .groups = "drop")

  sample_order_df2 <- homeo_rel_df %>%
    select(sample, Clonal_Group, relative_abundance) %>%
    distinct() %>%
    tidyr::pivot_wider(
      names_from = Clonal_Group,
      values_from = relative_abundance,
      values_fill = 0
    ) %>%
    distinct(sample, .keep_all = TRUE)

  # make sure all ordering columns exist even if absent in this dataset
  for (nm in clonal_group_levels) {
    if (!nm %in% colnames(sample_order_df2)) {
      sample_order_df2[[nm]] <- 0
    }
  }

  sample_order_df2 <- sample_order_df2 %>%
    arrange(
      desc(.data[["Hyperexpanded (0.1 < X <= 1)"]]),
      desc(.data[["Large (0.01 < X <= 0.1)"]]),
      desc(.data[["Medium (0.001 < X <= 0.01)"]]),
      desc(.data[["Small (1e-04 < X <= 0.001)"]]),
      desc(.data[["Rare (0 < X <= 1e-04)"]])
    )

  sample_levels2 <- unique(as.character(sample_order_df2$sample))

  homeo_rel_df <- homeo_rel_df %>%
    mutate(
      sample = factor(as.character(sample), levels = sample_levels2),
      Clonal_Group = factor(Clonal_Group, levels = clonal_group_levels)
    )

  p_homeo_rel <- ggplot(homeo_rel_df, aes(x = sample, y = relative_abundance, fill = Clonal_Group)) +
    geom_col(width = 0.8) +
    theme_scratch_pub(params$base_size) +
    labs(
      title = "Clonal Homeostasis",
      x = "Samples",
      y = "Relative Abundance",
      fill = "Clonal Group"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )

  print(p_homeo_rel)

  save_plot_safe(
    p_homeo_rel,
    glue("clonal_homeostasis.{params$figure_format}"),
    width = 10,
    height = 7
  )
}

if (isTRUE(params$show_diversity) && exists("export_cells")) {
  cat("### Clonal Diversity\n")

  div_df <- export_cells %>%
    filter(!is.na(sample), sample != "", !is.na(clone_id), clone_id != "") %>%
    group_by(sample, clone_id) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    summarise(
      shannon = -sum((n / sum(n)) * log(n / sum(n))),
      richness = dplyr::n(),
      .groups = "drop"
    ) %>%
    arrange(desc(shannon))

  div_df <- div_df %>%
    mutate(sample = factor(sample, levels = unique(as.character(sample))))

  p_div <- ggplot(div_df, aes(x = sample, y = shannon)) +
    geom_col(fill = "gray35", width = 0.8) +
    theme_scratch_pub(params$base_size) +
    labs(
      title = "Clonal Diversity (Shannon Index)",
      x = "Sample",
      y = "Shannon Index"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
    )

  print(p_div)

  save_plot_safe(
    p_div,
    glue("clonal_diversity.{params$figure_format}"),
    width = 9,
    height = 6
  )
}

  # p_div <- tryCatch({
  #   scRepertoire::clonalDiversity(seu_screp, 
  #                                 cloneCall = "strict", 
  #                                 group.by = sample_col_seu %||% "orig.ident", 
  #                                 n.boots = 20) + # Lower boots for speed
  #     theme_scratch_pub(params$base_size) +
  #     theme(axis.text.x = element_text(angle = 45, hjust = 1, fontsize = 8)) +
  #     ggtitle("Clonal Diversity")
  # }, error = function(e) {
  #   message("Error in clonalDiversity: ", e$message)
  #   return(NULL)
  # })
  # 
  # if (!is.null(p_div)) {
  #   print(p_div)
  #   save_plot_safe(p_div, glue("clonal_diversity.{params$figure_format}"), width = 9, height = 6)
  # }
# }

# if (isTRUE(params$show_homeostasis) && screp_combined_expression_ok) {
#   p_homeo <- tryCatch({
#     clonalHomeostasis(seu_screp, cloneCall = "strict", group.by = sample_col_seu %||% "orig.ident") +
#       theme_scratch_pub(params$base_size) +
#       ggtitle("Clonal homeostasis")
#   }, error = function(e) NULL)
# 
#   if (!is.null(p_homeo)) {
#     print(p_homeo)
#     save_plot_safe(p_homeo, glue("clonal_homeostasis.{params$figure_format}"), width = 9, height = 6)
#   }
# }
# 
# if (isTRUE(params$show_diversity) && screp_combined_expression_ok) {
#   p_div <- tryCatch({
#     clonalDiversity(seu_screp, cloneCall = "strict", group.by = sample_col_seu %||% "orig.ident") +
#       theme_scratch_pub(params$base_size) +
#       ggtitle("Clonal diversity")
#   }, error = function(e) NULL)
# 
#   if (!is.null(p_div)) {
#     print(p_div)
#     save_plot_safe(p_div, glue("clonal_diversity.{params$figure_format}"), width = 9, height = 6)
#   }
# }
```

## sample-condition-patient
```{r}
#| label: sample-condition-patient
show_sample    <- isTRUE(params$show_sample_panels) && (!is.null(sample_col_seu) || "META_SAMPLE" %in% colnames(seu@meta.data))
show_condition <- isTRUE(params$show_condition_panels) && (!is.null(condition_col_seu) || "META_TIMECOND" %in% colnames(seu@meta.data))
show_patient   <- isTRUE(params$show_patient_panels) && (!is.null(patient_col_seu) || "META_PATIENT" %in% colnames(seu@meta.data))
show_timepoint <- isTRUE(params$show_timepoint_panels) && (!is.null(timepoint_col_seu) || "META_TIMECOND" %in% colnames(seu@meta.data))
```


## sample-summary-plot
```{r}
#| results: asis

if (show_sample) {
  sample_summary <- export_cells %>%
    filter(!is.na(sample)) %>%
    group_by(sample) %>%
    summarise(
      n_cells = n(),
      n_tcr = sum(has_tcr, na.rm = TRUE),
      n_paired = sum(paired_tcr, na.rm = TRUE),
      pct_tcr = safe_percent(n_tcr, n_cells),
      .groups = "drop"
    )

  if (nrow(sample_summary) > 0) {
    # FIX: Use knitr::knit_print() instead of print() to ensure HTML renders correctly
    # and remove the print() call entirely if possible.
    ktab <- kable(
        sample_summary,
        caption = "Sample-level TCR Stats"
      ) %>%
      kable_styling(
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE
      )
    
    # This ensures the table renders as HTML in the final document
    knitr::knit_print(ktab) 
    
    # Add a small break between table and plot
    cat("\n\n")

    p_sample <- sample_summary %>%
      mutate(sample = fct_reorder(sample, pct_tcr)) %>%
      ggplot(aes(x = sample, y = pct_tcr)) +
      geom_col(fill = "steelblue") +
      coord_flip() +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      labs(
        title = "TCR-positive fraction by sample",
        x = NULL,
        y = "Fraction"
      ) +
      theme_scratch_pub(params$base_size) +
      theme(
        # INCREASED: Changed from size=7 to base_size to make sample names legible
        axis.text.y = element_text(size = params$base_size), 
        axis.text.x = element_text(size = params$base_size),
        axis.title.x = element_text(size = params$base_size + 2, face = "bold"),
        plot.title = element_text(size = params$base_size + 4, face = "bold")
      )

    print(p_sample)
    save_plot_safe(p_sample, glue("sample_tcr_positive_fraction.{params$figure_format}"))
  } else {
    message("sample_summary is empty. Check whether the sample column is missing or all NA.")
  }
}

# #| label: sample-summary-plot
# 
# if (show_sample) {
#   sample_summary <- export_cells %>%
#     filter(!is.na(sample)) %>%
#     group_by(sample) %>%
#     summarise(
#       n_cells = n(),
#       n_tcr = sum(has_tcr, na.rm = TRUE),
#       n_paired = sum(paired_tcr, na.rm = TRUE),
#       pct_tcr = safe_percent(n_tcr, n_cells),
#       .groups = "drop"
#     )
# 
#   if (nrow(sample_summary) > 0) {
#     # Print table so we see data
#     # print(kable(sample_summary, caption = "Sample-level TCR Stats") %>% kable_styling())
#     cat(kable(sample_summary, format = "html", caption = "Sample-level TCR Stats") %>% 
#       kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
#       as.character())
#     p_sample <- sample_summary %>%
#       mutate(sample = fct_reorder(sample, pct_tcr)) %>%
#       ggplot(aes(x = sample, y = pct_tcr)) +
#       geom_col(fill = "steelblue") +
#       coord_flip() +
#       scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
#       labs(title = "TCR-positive fraction by sample", x = NULL, y = "Fraction") +
#       theme_scratch_pub(params$base_size) +
#       theme(axis.text.y = element_text(size = 7)) # Small font for many samples
# 
#     print(p_sample)
#     save_plot_safe(p_sample, glue("sample_tcr_positive_fraction.{params$figure_format}"))
#   } else {
#     message("!!! DIAGNOSTIC: sample_summary is empty. Check if 'sample' column has NAs.")
#   }
# }

# if (show_sample) {
#   sample_summary <- export_cells %>%
#     group_by(sample) %>%
#     summarise(
#       n_cells = n(),
#       n_tcr = sum(has_tcr, na.rm = TRUE),
#       n_paired = sum(paired_tcr, na.rm = TRUE),
#       pct_tcr = safe_percent(n_tcr, n_cells),
#       pct_paired = safe_percent(n_paired, n_cells),
#       median_clone_size = median(clone_size, na.rm = TRUE),
#       .groups = "drop"
#     )
# 
#   save_table_safe(sample_summary, "sample_level_tcell_integration_summary.tsv")
# 
#   p_sample <- sample_summary %>%
#     mutate(sample = fct_reorder(sample, pct_tcr)) %>%
#     ggplot(aes(x = sample, y = pct_tcr)) +
#     geom_col(fill = "steelblue") +
#     coord_flip() +
#     scale_y_continuous(labels = percent_format(accuracy = 1)) +
#     labs(
#       title = "TCR-positive fraction by sample",
#       x = NULL,
#       y = "TCR-positive cells"
#     ) +
#     theme_scratch_pub(params$base_size)
# 
#   p_sample
#   save_plot_safe(p_sample, glue("sample_tcr_positive_fraction.{params$figure_format}"))
# }
```

## condition-summary-plot
```{r}
#| label: condition-summary-plot
if (show_condition) {
  condition_summary <- export_cells %>%
    group_by(condition) %>%
    summarise(
      n_cells = n(),
      n_tcr = sum(has_tcr, na.rm = TRUE),
      n_paired = sum(paired_tcr, na.rm = TRUE),
      pct_tcr = safe_percent(n_tcr, n_cells),
      pct_paired = safe_percent(n_paired, n_cells),
      .groups = "drop"
    )

  save_table_safe(condition_summary, "condition_level_tcell_integration_summary.tsv")

  p_condition <- condition_summary %>%
    mutate(condition = fct_reorder(condition, pct_tcr)) %>%
    ggplot(aes(x = condition, y = pct_tcr)) +
    geom_col(fill = "purple4") +
    coord_flip() +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = "TCR-positive fraction by condition",
      x = NULL,
      y = "TCR-positive cells"
    ) +
    theme_scratch_pub(params$base_size)

  p_condition
  save_plot_safe(p_condition, glue("condition_tcr_positive_fraction.{params$figure_format}"))
}
```

## patient-summary-plot

## patient-summary-plot
```{r}
#| label: patient-summary-plot

# Ensure the patient column is defined from the params
# If not already resolved in a 'resolve-columns' chunk:
curr_patient_col <- params$patient_col %||% "patient_id"

if (show_patient && curr_patient_col %in% colnames(export_cells)) {
  
  patient_summary <- export_cells %>%
    # Use the parameter from the params object
    group_by(across(all_of(curr_patient_col))) %>% 
    summarise(
      n_cells = n(),
      n_tcr = sum(has_tcr, na.rm = TRUE),
      n_paired = sum(paired_tcr, na.rm = TRUE),
      pct_tcr = safe_percent(n_tcr, n_cells),
      pct_paired = safe_percent(n_paired, n_cells),
      .groups = "drop"
    ) %>%
    # Rename using the parameter
    rename(patient = !!sym(curr_patient_col))

  if (nrow(patient_summary) > 0) {
    save_table_safe(patient_summary, "patient_level_tcell_integration_summary.tsv")

    p_patient <- patient_summary %>%
      mutate(patient = fct_reorder(patient, pct_tcr)) %>%
      ggplot(aes(x = patient, y = pct_tcr)) +
      geom_col(fill = "darkgreen") +
      coord_flip() +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      labs(
        title = "TCR-positive fraction by patient",
        subtitle = paste("Column used:", curr_patient_col),
        x = NULL,
        y = "TCR-positive fraction"
      ) +
      theme_scratch_pub(params$base_size) +
      theme(
        axis.text.y = element_text(size = params$base_size),
        axis.text.x = element_text(size = params$base_size),
        axis.title.x = element_text(size = params$base_size + 2, face = "bold"),
        plot.title = element_text(size = params$base_size + 4, face = "bold")
      )

    # Explicitly print inside the if block
    print(p_patient) 
    
    save_plot_safe(p_patient, glue("patient_tcr_positive_fraction.{params$figure_format}"))
  }
} else {
  message("Skipping patient plot: column '", curr_patient_col, "' not found in data.")
}
```

## warnings
```{r}
#| label: warnings
warn_tbl <- tibble(
  warning = c(
    "Low final barcode overlap",
    "Few T-cell matches after merge",
    "Low paired TCR fraction",
    "Annotation column auto-detected",
    "Sample column missing",
    "No scRepertoire combineExpression output"
  ),
  triggered = c(
    isTRUE(final_overlap_rate < 0.6),
    isTRUE(sum(seu_t$has_tcr, na.rm = TRUE) < 100),
    isTRUE(mean(seu_t$paired_tcr, na.rm = TRUE) < 0.3),
    isTRUE(params$label_col == ""),
    isTRUE(is.null(sample_col_seu)),
    isTRUE(!screp_combined_expression_ok)
  ),
  interpretation = c(
    "Final TCR to GEX overlap is below 60%. Check barcode prefix/suffix conventions and sample naming.",
    "Very few T cells retained matched TCR information. Inspect upstream VDJ QC and join strategy diagnostics.",
    "Paired TCR fraction is low. This may reflect assay quality, chain filtering, or incomplete pairing.",
    "Annotation column was auto-detected rather than explicitly specified.",
    "No reliable sample column was detected in Seurat metadata. Some grouped summaries may be incomplete.",
    "Optional scRepertoire combineExpression integration was not produced. The pipeline continued using the validated fast-merge TCR↔GEX integration outputs."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current integration settings.")
} else {
  kable(
    warn_tbl %>% select(-triggered),
    caption = "Automatically generated integration warnings and notes."
  ) %>%
    kable_styling(
      full_width = TRUE,
      bootstrap_options = c("striped", "hover", "condensed")
    )
}

# Show the actual scRepertoire error message if available
if (!isTRUE(screp_combined_expression_ok) &&
    exists("screp_error_msg") &&
    !is.null(screp_error_msg) &&
    !is.na(screp_error_msg) &&
    nzchar(screp_error_msg)) {

  cat("\n\n### scRepertoire combineExpression diagnostic message\n")
  cat("`", screp_error_msg, "`\n", sep = "")
}
```

## session-info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.tcell_integration.txt"))
sessionInfo()
```
