---
title: "TCRtoolkit — Integrated Pipeline Design"
subtitle: "Bulk + single-cell TCR analysis under one shared engine"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "100%")

# Diagrams are rendered to static PNGs (via Graphviz) so they embed in BOTH
# HTML and PDF. Requires: DiagrammeR, DiagrammeRsvg, magick.
for (p in c("DiagrammeR", "DiagrammeRsvg", "magick")) {
  if (!requireNamespace(p, quietly = TRUE))
    install.packages(p, repos = "https://cloud.r-project.org", quiet = TRUE)
}

fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)

# Render a Graphviz DOT string to a PNG and return the path for include_graphics().
dot_png <- function(dot, name) {
  svg <- DiagrammeRsvg::export_svg(DiagrammeR::grViz(dot))
  svg_file <- file.path(fig_dir, paste0(name, ".svg"))
  png_file <- file.path(fig_dir, paste0(name, ".png"))
  writeLines(svg, svg_file)
  img <- magick::image_convert(magick::image_read(svg_file), "png")
  magick::image_write(img, png_file)
  knitr::include_graphics(png_file)
}
```

# Overview

TCRtoolkit is a Nextflow DSL2 pipeline that analyses T-cell receptor (TCR) repertoires from **bulk** and **single-cell** data under a single entry point, selected with `--mode` (default `bulk`).

The core design principle is **one shared bulk analysis engine**. Single-cell data is *pseudobulked* into clonotype tables and then run through the exact same OLGA → GIANA / GLIPH2 / TCRdist3 → diversity code as bulk input, so results from both modalities are directly comparable and there is no duplicated analysis logic.

# Design Principles

- **One engine, two modalities.** The bulk repertoire engine is shared; single-cell adds a head (QC + pseudobulk) and, when a GEX object is present, a tail (cell-level clustering/reports).
- **Pseudobulk belongs to single-cell.** Collapsing cells into clonotype tables lives entirely in the single-cell path; the bulk path is untouched.
- **Pool by sample, cluster by patient, phenotype downstream.** (See *Pseudobulk pooling*.)
- **Additive, non-breaking.** Bulk behaviour is unchanged; the single-cell modality reuses the bulk subworkflows through additive entry points.

# Architecture & Modality Dispatch

`--mode` selects the modality. The two modalities are colour-coded by **module source**: the shared **tcrtoolkit** engine (green) runs both bulk input and single-cell pseudobulk; **SCRATCH** single-cell modules (purple) provide the QC/integration head and the cell-level reporting tail; **bridges** (amber) convert between the two.

```{r architecture}
dot_png('
digraph {
  rankdir=TB; bgcolor="white"; compound=true; nodesep=0.3; ranksep=0.55;
  node [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=10, penwidth=1.4, margin="0.16,0.07"];
  edge [fontname="Helvetica", fontsize=8.5, color="#94a3b8", arrowsize=0.8];

  START [label="nextflow run .   --mode", shape=box, style="filled,rounded", fillcolor="#0f172a", fontcolor="white", fontsize=13, penwidth=0];

  subgraph cluster_bulk {
    label="  BULK TCR  "; labeljust="l"; fontname="Helvetica-Bold"; fontsize=13; fontcolor="#166534";
    style="rounded,filled"; fillcolor="#f0fdf4"; color="#22c55e"; penwidth=1.6;
    BIN [label="AIRR / Adaptive\nsamplesheet", fillcolor="#e2e8f0", color="#64748b"];
    B_ANN [label="ANNOTATE", fillcolor="#dcfce7", color="#16a34a"];
    B_SAM [label="SAMPLE\ndiversity · TCRdist3", fillcolor="#dcfce7", color="#16a34a"];
    B_PAT [label="PATIENT\nGIANA · GLIPH2", fillcolor="#dcfce7", color="#16a34a"];
    B_CMP [label="COMPARE", fillcolor="#dcfce7", color="#16a34a"];
    B_REP [label="REPORT", fillcolor="#dcfce7", color="#16a34a"];
    BIN -> B_ANN; B_ANN -> B_SAM; B_ANN -> B_PAT; B_ANN -> B_CMP; B_SAM -> B_REP;
  }

  subgraph cluster_sc {
    label="  SINGLE-CELL TCR  "; labeljust="l"; fontname="Helvetica-Bold"; fontsize=13; fontcolor="#5b21b6";
    style="rounded,filled"; fillcolor="#faf5ff"; color="#a855f7"; penwidth=1.6;
    SIN [label="Cell Ranger VDJ\n(+ optional GEX Seurat)", fillcolor="#e2e8f0", color="#64748b"];
    S_VQ [label="VDJ_QC", fillcolor="#ede9fe", color="#7c3aed"];
    GEXQ [label="GEX annotated\nobject provided?", shape=diamond, fillcolor="#fde68a", color="#b45309", fontsize=9];
    S_TI [label="TCELL_INTEGRATION", fillcolor="#ede9fe", color="#7c3aed"];
    BR_SC [label="SC_TO_CDR3\npseudobulk", fillcolor="#fef3c7", color="#d97706"];
    BR_VB [label="VDJ_TO_BULK\npseudobulk", fillcolor="#fef3c7", color="#d97706"];
    S_PQ [label="PSEUDOBULK_QC", fillcolor="#fef3c7", color="#d97706"];
    E_ANN [label="ANNOTATE_FROM_CONCAT", fillcolor="#dcfce7", color="#16a34a"];
    E_SAM [label="SAMPLE", fillcolor="#dcfce7", color="#16a34a"];
    E_PAT [label="PATIENT · GIANA · GLIPH2", fillcolor="#dcfce7", color="#16a34a"];
    S_RE [label="REPERTOIRE\n(cell-level if GEX,\nclonotype-level if not)", fillcolor="#ede9fe", color="#7c3aed"];
    S_MS [label="MASTER_SUMMARY\n(full if GEX, excl. CoNGA if not)", fillcolor="#ede9fe", color="#7c3aed"];

    subgraph cluster_gex {
      label="Requires GEX object — skipped in VDJ-only";
      fontname="Helvetica-Bold"; fontsize=10; fontcolor="#b45309";
      style="rounded,dashed"; color="#f59e0b"; penwidth=1.8;
      BR_CT [label="CLUSTER_TO_SC\nmap clusters to cells", fillcolor="#fef3c7", color="#d97706"];
      S_CG [label="CoNGA", fillcolor="#ede9fe", color="#7c3aed"];
      S_CN [label="CONSENSUS", fillcolor="#ede9fe", color="#7c3aed"];
      BR_CT -> S_CG; BR_CT -> S_CN;
    }

    SIN -> S_VQ; S_VQ -> GEXQ;
    GEXQ -> S_TI [label="yes  (Full SC)", fontcolor="#7c3aed"]; S_TI -> BR_SC; BR_SC -> S_PQ;
    GEXQ -> BR_VB [label="no  (VDJ-only)", fontcolor="#b45309"]; BR_VB -> S_PQ;
    S_PQ -> E_ANN; E_ANN -> E_SAM; E_ANN -> E_PAT;
    E_PAT -> BR_CT [label="Full SC only", style=dashed, fontcolor="#b45309"];
    E_SAM -> BR_CT [style=dashed];
    E_SAM -> S_RE; E_PAT -> S_RE;
    S_CN -> S_RE [label="enrich (GEX)", style=dashed];
    S_CG -> S_MS [style=dashed]; S_RE -> S_MS;
  }

  START -> BIN [label="bulk (default)", fontcolor="#166534"];
  START -> SIN [label="singlecell", fontcolor="#5b21b6"];

  subgraph cluster_legend {
    label="Module source / container"; fontname="Helvetica-Bold"; fontsize=10;
    style="rounded"; color="#cbd5e1"; margin=8;
    L1 [label="tcrtoolkit module — tcrtoolkit container", fillcolor="#dcfce7", color="#16a34a"];
    L2 [label="SCRATCH module — scratch-tcr container", fillcolor="#ede9fe", color="#7c3aed"];
    L3 [label="bridge (SC to bulk glue)", fillcolor="#fef3c7", color="#d97706"];
    L4 [label="input", fillcolor="#e2e8f0", color="#64748b"];
    L1 -> L2 [style=invis]; L2 -> L3 [style=invis]; L3 -> L4 [style=invis];
  }
}', "architecture")
```

Two things to read from this diagram:

- The single-cell modality **reuses the green tcrtoolkit modules** (ANNOTATE, SAMPLE, PATIENT — including GIANA/GLIPH2/TCRdist3) rather than duplicating them; the SCRATCH modules only add what is genuinely single-cell.
- The **`GEX annotated object provided?`** decision gates only the **cell-level clustering** steps.
  **Without a GEX object (VDJ-only)** the dashed box — **CLUSTER_TO_SC → CoNGA / CONSENSUS** — is
  **skipped** (CoNGA needs gene-expression data; the others map clusters onto cells that don't
  exist without a Seurat). **REPERTOIRE and MASTER_SUMMARY still run** in both routes: cell-level
  and full when a GEX object is present, clonotype-level repertoire and a CoNGA-excluded summary
  when it is not.

# Single-Cell Routes

The presence of an annotated GEX Seurat object (`--input_annotated_object`) selects the route.

| Stage | Full SC (GEX present) | VDJ-only (no GEX) |
|----|:--:|:--:|
| VDJ QC | Yes | Yes |
| T-cell integration | Yes | No (no Seurat) |
| Pseudobulk | Yes (from integrated cells) | Yes (from contigs) |
| Pseudobulk QC gate | Yes | Yes |
| OLGA + GIANA + GLIPH2 + TCRdist3 + diversity | Yes | Yes |
| QC / repertoire report | Yes | Yes |
| Map clusters onto cells (CLUSTER_TO_SC) | Yes | No |
| CoNGA / consensus clustering | Yes | No (needs GEX) |
| Repertoire analysis | Yes (cell-level) | Yes (clonotype-level, adapted) |
| Master summary | Yes (full) | Yes (excludes CoNGA) |

The middle band (pseudobulk → QC → shared engine → report) is **identical** in both routes; only the head (T-cell integration) and tail (cell-level clustering/reports) require the GEX object.

# Pseudobulk Pooling

Single cells are collapsed into per-unit clonotype tables before the shared engine runs.

- **Default — pool by sample.** Each sample is one unit; `patient` is carried as metadata so the shared `PATIENT` step **pools per-patient** for GIANA / GLIPH2 / TCRdist3 clustering.
- **Phenotype is applied downstream**, not as a pre-split: cluster results are mapped back onto cells and stratified by cell state in the CoNGA / repertoire reports.
- **Optional per-cell-type view** — `--pseudobulk_by_phenotype true` creates `{sample}__{phenotype}` units; under-powered units are dropped by the QC gate.

```{r pooling}
dot_png('
digraph {
  rankdir=LR; bgcolor="transparent";
  node [shape=box, style="rounded,filled", fillcolor="white", fontname="Helvetica", fontsize=10];
  edge [fontname="Helvetica", fontsize=9];
  C [label="Annotated cells\n(sample, patient, phenotype, CDR3b)"];
  D [label="pseudobulk_by_phenotype?", shape=diamond, fillcolor="#f1f5f9"];
  E [label="Unit = sample\n(patient carried as metadata)"];
  F [label="Unit = sample__phenotype"];
  G [label="PATIENT pools per patient\nfor clustering"];
  H [label="Phenotype applied downstream\n(cluster mapping, CoNGA, repertoire)"];
  C -> D;
  D -> E [label="false (default)"];
  D -> F [label="true"];
  E -> G; F -> G; G -> H;
}', "pooling")
```

**Rationale:** pooling the full repertoire per patient preserves the statistical power that TCR clustering methods need, and keeps clusters that span cell states intact. Phenotype then enters as *association* (which cell types populate each cluster) rather than fragmenting the repertoire before clustering.

# Container Split

Two container images run the pipeline; processes are routed per-process.

| Container | Runs |
|----|----|
| `ghcr.io/karchinlab/tcrtoolkit:main` (`--container`) | Bulk / shared engine — including **GIANA, GLIPH2, TCRdist3** (tcrtoolkit tools), OLGA, diversity stats, and the pandas bridges / QC gate |
| `syedsazaidi/scratch-tcr` (`--sc_container`) | Single-cell R / Seurat / scanpy steps — VDJ QC, T-cell integration, CoNGA, consensus, repertoire, master summary, cluster mapping |

# Shared Bulk Engine — Additive Entry Points

The single-cell modality reuses the bulk subworkflows through additive, backward-compatible hooks (bulk behaviour unchanged):

- **`ANNOTATE_FROM_CONCAT`** — a variant of `ANNOTATE` that skips the per-sample column-rename step (single-cell data arrives already in canonical clonotype schema) and runs the same sort → dedup → OLGA chain.
- **`PATIENT` cluster emits** — exposes GIANA / GLIPH2 cluster outputs so they can be mapped back onto single cells; grouping key is configurable (`--patient_col`).
- **`SAMPLE` tcrdist emits** — exposes TCRdist3 outputs so the single-cell tail reuses them without a second run.

# Reports

- **Bulk / both single-cell routes:** QC and repertoire HTML (`template_qc.qmd`), plus GLIPH2 clustering when enabled.
- **Full single-cell:** VDJ QC, T-cell integration, CoNGA, consensus clustering, repertoire, and a master summary tying the single-cell analyses together.

------------------------------------------------------------------------

*Render:* open in RStudio and **Knit** (HTML or PDF), or `rmarkdown::render("pipeline_design.Rmd")`. Diagrams render as static images (Graphviz), so they work in both formats. PDF output additionally requires a LaTeX distribution (`tinytex::install_tinytex()` if you don't have one).
