---
title: "SCRATCH-TCR: VDJ QC Analysis"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    df-print: paged
    smooth-scroll: true
    anchor-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  # ======================================================
  # Inputs
  # ======================================================
  sample_sheet: "vdj_samples.txt"
  metadata_file: ""
  outdir: "VDJ_QC"
  input_annotated_object: ""

  # ======================================================
  # Sample sheet columns
  # sample sheet must minimally contain:
  #   sample, path
  # ======================================================
  sample_sheet_sample_col: "sample"
  sample_sheet_path_col: "path"

  # ======================================================
  # Metadata mapping (optional)
  # ======================================================
  meta_barcode_col: "barcode"
  meta_sample_col: "sample"
  meta_patient_col: "patient_id"
  meta_condition_col: "condition"
  meta_timepoint_col: "timepoint"

  meta_barcode_candidates: !expr c("barcode","cell_id","cell","CB")
  meta_sample_candidates: !expr c("sample","sample_id","orig.ident","library_id")
  meta_patient_candidates: !expr c("patient_id","patient","subject_id")
  meta_condition_candidates: !expr c("condition","group","status","DiseaseStatus")
  meta_timepoint_candidates: !expr c("timepoint","time","visit","day")

  # ======================================================
  # QC parameters
  # ======================================================
  chain_mode: "auto"            # auto / T-AB / T-GD / both
  require_productive: true
  require_high_conf: true
  require_full_length: true
  min_umis: 1
  min_reads: 0
  keep_one_per_chain: true
  keep_paired_only: false
  keep_dual_alpha: true
  keep_dual_beta: true
  cdr3_aa_min: 8
  cdr3_aa_max: 25
  drop_stop_or_frames: true
  top_n_samples_table: 50

  # ======================================================
  # Plot/report settings
  # ======================================================
  figures_dir: "figures"
  tables_dir: "tables"
  save_tables: true
  save_figures: true
  figure_format: "png"
  figure_width: 10
  figure_height: 7
  figure_dpi: 300
  base_size: 12

  top_n_v: 20
  top_n_j: 20
  min_clone_size_plot: 1

  show_reads_umis: true
  show_patient_panels: true
  show_condition_panels: true
  show_timepoint_panels: true
  show_clone_rank_plot: true
  show_vj_heatmaps: true

  report_label: "VDJ QC Analysis"
---


##SETUP
```{r}
#| label: setup
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(ggplot2)
  library(scales)
  library(purrr)
  library(glue)
  library(knitr)
  library(kableExtra)
  library(ComplexHeatmap)
  library(circlize)
  library(patchwork)
  library(RColorBrewer)
  library(plotly)
  # library(DT)
  library(htmltools)
  library(htmlwidgets)
  library(tibble)
})

options(stringsAsFactors = FALSE)

dir.create(params$outdir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$figures_dir), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(params$outdir, params$tables_dir), recursive = TRUE, showWarnings = FALSE)

`%||%` <- function(x, y) if (!is.null(x) && length(x) > 0 && !all(is.na(x))) x else y

# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold"),
#       axis.text = element_text(color = "black"),
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold"),
#       legend.title = element_text(face = "bold"),
#       legend.key = element_blank(),
#       plot.caption = element_text(size = base_size - 2, color = "grey40")
#     )
# }
# theme_scratch_pub <- function(base_size = 12) {
#   theme_bw(base_size = base_size) +
#     theme(
#       plot.title = element_text(face = "bold", size = base_size + 2, hjust = 0),
#       plot.subtitle = element_text(size = base_size, hjust = 0),
#       axis.title = element_text(face = "bold", size = base_size - 1),
#       # Shrink Y-axis labels slightly
#       axis.text.y = element_text(color = "black", size = base_size - 3), 
#       # Rotate X-axis labels 45 degrees and shrink
#       axis.text.x = element_text(color = "black", size = base_size - 3, angle = 45, hjust = 1), 
#       panel.grid.minor = element_blank(),
#       panel.grid.major = element_line(linewidth = 0.1, color = "grey90"),
#       strip.background = element_rect(fill = "grey95", color = "grey80"),
#       strip.text = element_text(face = "bold", size = base_size - 2),
#       legend.title = element_text(face = "bold", size = base_size - 2),
#       legend.text = element_text(size = base_size - 3),
#       plot.caption = element_text(size = base_size - 4, color = "grey40")
#     )
# }
theme_scratch_pub <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 3, hjust = 0),
      plot.subtitle = element_text(size = base_size, hjust = 0, color = "grey30"),
      axis.title = element_text(face = "bold", size = base_size),
      axis.text.y = element_text(color = "black", size = base_size - 2),
      axis.text.x = element_text(color = "black", size = base_size - 2, angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey92"),
      strip.background = element_rect(fill = "#eef5f3", color = "#d9e6e1"),
      strip.text = element_text(face = "bold", size = base_size - 1),
      legend.title = element_text(face = "bold", size = base_size - 1),
      legend.text = element_text(size = base_size - 2),
      legend.key = element_blank(),
      plot.caption = element_text(size = base_size - 3, color = "grey40"),
      plot.margin = margin(10, 12, 10, 10)
    )
}

save_plot_safe <- function(plot_obj, filename,
                           width = params$figure_width,
                           height = params$figure_height,
                           dpi = params$figure_dpi) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  ggsave(
    filename = file.path(params$outdir, params$figures_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white",
    limitsize = FALSE
  )
}

save_table_safe <- function(df, filename) {
  if (!isTRUE(params$save_tables)) return(invisible(NULL))
  fwrite(df, file.path(params$outdir, params$tables_dir, filename), sep = "\t")
}

safe_read_table <- function(path) {
  if (is.null(path) || is.na(path) || path == "" || !file.exists(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "tab", "txt")) {
    fread(path, sep = "\t")
  } else {
    fread(path)
  }
}

first_existing_col <- function(df, preferred, candidates = character()) {
  if (!is.null(preferred) && preferred != "" && preferred %in% colnames(df)) return(preferred)
  hits <- intersect(candidates, colnames(df))
  if (length(hits) > 0) return(hits[[1]])
  return(NULL)
}

require_cols <- function(df, cols, object_name = deparse(substitute(df))) {
  miss <- setdiff(cols, colnames(df))
  if (length(miss) > 0) {
    stop(glue("{object_name} is missing required columns: {paste(miss, collapse = ', ')}"))
  }
}

has_col <- function(df, col) {
  !is.null(col) && col %in% colnames(df)
}

to_logical <- function(v) {
  if (is.logical(v)) return(v)
  if (is.numeric(v)) return(v != 0)
  if (is.character(v)) return(tolower(trimws(v)) %in% c("true", "t", "yes", "y", "1"))
  return(as.logical(v))
}

numify <- function(x) suppressWarnings(as.numeric(x))

clean_chain <- function(x) {
  x <- as.character(x)
  x <- toupper(trimws(x))
  x
}

infer_chain_mode <- function(chain_vals) {
  chain_vals <- unique(na.omit(chain_vals))
  has_ab <- any(chain_vals %in% c("TRA","TRB"))
  has_gd <- any(chain_vals %in% c("TRG","TRD"))
  if (has_ab && !has_gd) return("T-AB")
  if (!has_ab && has_gd) return("T-GD")
  if (has_ab && has_gd) return("both")
  "unknown"
}

filter_chain_mode <- function(df, chain_col, mode = "auto") {
  if (!has_col(df, chain_col)) return(df)
  mode_use <- mode
  if (mode_use == "auto") mode_use <- infer_chain_mode(df[[chain_col]])
  if (mode_use == "T-AB") return(df %>% filter(.data[[chain_col]] %in% c("TRA","TRB")))
  if (mode_use == "T-GD") return(df %>% filter(.data[[chain_col]] %in% c("TRG","TRD")))
  df
}

safe_pct <- function(x, denom) ifelse(denom == 0, NA_real_, x / denom)

coalesce_chr <- function(x, y) ifelse(is.na(x) | x == "", y, x)

detect_vdj_file <- function(outs_dir, fname) {
  roots <- c(
    outs_dir,
    file.path(outs_dir, "vdj_t"),
    file.path(outs_dir, "vdj_b")
  )
  fp <- file.path(roots, fname)
  i <- which(file.exists(fp))[1]
  if (is.na(i)) return(NULL)
  fp[[i]]
}

read_samplesheet <- function(path, sample_col = "sample", path_col = "path") {
  stopifnot(file.exists(path))
  df <- safe_read_table(path) %>% as.data.frame()
  require_cols(df, c(sample_col, path_col), "sample_sheet")
  stopifnot(!anyDuplicated(df[[sample_col]]))

  df <- df %>%
    mutate(
      sample = as.character(.data[[sample_col]]),
      raw_path = as.character(.data[[path_col]])
    )

  df$outs <- vapply(df$raw_path, function(p) {
    p <- normalizePath(p, mustWork = FALSE)
    if (dir.exists(file.path(p, "outs"))) {
      file.path(p, "outs")
    } else if (grepl("/outs/?$", p) && dir.exists(p)) {
      p
    } else {
      stop("Could not find 'outs' directory for path: ", p)
    }
  }, character(1))

  df %>% select(sample, outs) %>% distinct()
}

load_vdj_one <- function(sname, outs_dir) {
  contigs_fp <- detect_vdj_file(outs_dir, "filtered_contig_annotations.csv")
  clonotypes_fp <- detect_vdj_file(outs_dir, "clonotypes.csv")
  metrics_fp <- detect_vdj_file(outs_dir, "metrics_summary.csv")

  if (is.null(contigs_fp)) {
    stop("[", sname, "] missing filtered_contig_annotations.csv under ", outs_dir)
  }
  if (is.null(clonotypes_fp)) {
    stop("[", sname, "] missing clonotypes.csv under ", outs_dir)
  }

  contigs <- fread(contigs_fp) %>% as.data.frame()
  clonotypes <- fread(clonotypes_fp) %>% as.data.frame()
  metrics <- NULL
  if (!is.null(metrics_fp) && file.exists(metrics_fp)) {
    metrics <- tryCatch(fread(metrics_fp) %>% as.data.frame(), error = function(e) NULL)
  }

  contigs$sample <- sname
  clonotypes$sample <- sname
  if (!is.null(metrics)) metrics$sample <- sname

  list(
    contigs = contigs,
    clonotypes = clonotypes,
    metrics = metrics
  )
}


metric_card <- function(title, value, subtitle = NULL, color = "#eef5f3") {
  htmltools::div(
    style = paste0(
      "background:", color, "; border:1px solid #d8e3df; border-radius:12px; ",
      "padding:14px 16px; margin:8px 8px 8px 0; min-width:200px; display:inline-block;"
    ),
    htmltools::tags$div(style = "font-size:13px; color:#4b5b57; font-weight:600;", title),
    htmltools::tags$div(style = "font-size:28px; font-weight:700; color:#1f2d2a;", value),
    if (!is.null(subtitle)) htmltools::tags$div(style = "font-size:12px; color:#6b7c77;", subtitle)
  )
}


scroll_kable <- function(df, caption = NULL, height = "360px", font_size = 11) {
  kableExtra::kbl(df, caption = caption) %>%
    kableExtra::kable_styling(
      full_width = FALSE,
      bootstrap_options = c("striped","hover","condensed"),
      font_size = font_size
    ) %>%
    kableExtra::scroll_box(width = "100%", height = height)
}

save_widget_safe <- function(widget, filename) {
  if (!isTRUE(params$save_figures)) return(invisible(NULL))
  htmlwidgets::saveWidget(
    widget = widget,
    file = file.path(params$outdir, params$figures_dir, filename),
    selfcontained = TRUE
  )
}


ggplotly_clean <- function(p, tooltip = c("x", "y")) {
  plotly::ggplotly(p, tooltip = tooltip) %>%
    plotly::layout(
      margin = list(l = 60, r = 30, b = 60, t = 60),
      legend = list(orientation = "v")
    )
}

```

#read input
```{r}
#| label: read-inputs

sample_sheet <- read_samplesheet(
  params$sample_sheet,
  sample_col = params$sample_sheet_sample_col,
  path_col = params$sample_sheet_path_col
)

metadata_tbl <- safe_read_table(params$metadata_file)

if ((is.null(metadata_tbl) || nrow(metadata_tbl) == 0) &&
    !is.null(params$input_annotated_object) &&
    params$input_annotated_object != "" &&
    !grepl("NO_FILE", basename(params$input_annotated_object)) &&
    file.exists(params$input_annotated_object)) {

  suppressPackageStartupMessages({
    library(Seurat)
    library(tibble)
    library(dplyr)
  })

  seu <- readRDS(params$input_annotated_object)

  metadata_tbl <- seu@meta.data %>%
    tibble::rownames_to_column("cell_id_full") %>%
    mutate(
      sample = as.character(orig.ident),
      barcode = if ("raw_cell_id" %in% colnames(.)) {
        sub("-1$", "", as.character(raw_cell_id))
      } else {
        sub("^[^_]+_", "", cell_id_full) |> sub("-1$", "", x = _)
      },
      patient_id = if ("patient_id" %in% colnames(.)) as.character(patient_id) else NA_character_,
      condition  = if ("condition"  %in% colnames(.)) as.character(condition)  else NA_character_,
      timepoint  = if ("timepoint"  %in% colnames(.)) as.character(timepoint)  else NA_character_
    )

  cat("\nLoaded metadata from annotated object\n")
  print(colnames(metadata_tbl))
  print(head(metadata_tbl[, c("sample","barcode","patient_id","condition","timepoint")], 10))
}

```

##load files VDJ
```{r}
#| label: load-vdj
vdj_list <- lapply(seq_len(nrow(sample_sheet)), function(i) {
  load_vdj_one(sample_sheet$sample[i], sample_sheet$outs[i])
})
names(vdj_list) <- sample_sheet$sample

contigs_all_pre <- rbindlist(lapply(vdj_list, `[[`, "contigs"), use.names = TRUE, fill = TRUE) %>% as.data.frame()
clonotypes_all  <- rbindlist(lapply(vdj_list, `[[`, "clonotypes"), use.names = TRUE, fill = TRUE) %>% as.data.frame()

metrics_list <- lapply(vdj_list, `[[`, "metrics")
metrics_list <- metrics_list[!vapply(metrics_list, is.null, logical(1))]
metrics_all <- if (length(metrics_list) > 0) {
  rbindlist(metrics_list, use.names = TRUE, fill = TRUE) %>% as.data.frame()
} else {
  NULL
}
```

## validate-standardize
```{r}
#| label: validate-and-standardize
need_cols <- c("barcode", "chain", "cdr3", "productive")
require_cols(contigs_all_pre, need_cols, "contigs_all_pre")

contigs_all_pre <- contigs_all_pre %>%
  mutate(
    sample = as.character(sample),
    barcode = as.character(barcode),
    cell_id = as.character(barcode),
    chain = clean_chain(chain),
    productive = if ("productive" %in% names(.)) to_logical(productive) else productive,
    high_confidence = if ("high_confidence" %in% names(.)) to_logical(high_confidence) else high_confidence,
    full_length = if ("full_length" %in% names(.)) to_logical(full_length) else full_length,
    umis = if ("umis" %in% names(.)) numify(umis) else NA_real_,
    reads = if ("reads" %in% names(.)) numify(reads) else NA_real_,
    cdr3 = as.character(cdr3),
    cdr3_nt = if ("cdr3_nt" %in% names(.)) as.character(cdr3_nt) else NA_character_,
    v_gene = if ("v_gene" %in% names(.)) as.character(v_gene) else NA_character_,
    j_gene = if ("j_gene" %in% names(.)) as.character(j_gene) else NA_character_,
    raw_clonotype_id = if ("raw_clonotype_id" %in% names(.)) as.character(raw_clonotype_id) else if ("clonotype_id" %in% names(.)) as.character(clonotype_id) else NA_character_
  )
```

##merge-optional-metadata
```{r}
#| label: merge-optional-metadata

contigs_all_pre$patient_id <- NA_character_
contigs_all_pre$condition  <- NA_character_
contigs_all_pre$timepoint  <- NA_character_

if (!is.null(metadata_tbl) && nrow(metadata_tbl) > 0) {
  message("--- Merging VDJ contigs with GEX metadata ---")

  md_sample_col    <- "sample"
  md_barcode_col   <- "barcode"
  md_patient_col   <- if ("patient_id" %in% colnames(metadata_tbl)) "patient_id" else NULL
  md_condition_col <- if ("condition" %in% colnames(metadata_tbl)) "condition" else NULL
  md_timepoint_col <- if ("timepoint" %in% colnames(metadata_tbl)) "timepoint" else NULL

  metadata_prep <- metadata_tbl %>%
    mutate(
      md_sample  = as.character(.data[[md_sample_col]]),
      md_barcode = sub("-1$", "", as.character(.data[[md_barcode_col]]))
    )

  # barcode + sample level mapping
  barcode_meta_map <- metadata_prep %>%
    transmute(
      sample = md_sample,
      barcode_clean = md_barcode,
      patient_id_join = if (!is.null(md_patient_col)) as.character(.data[[md_patient_col]]) else NA_character_,
      condition_join  = if (!is.null(md_condition_col)) as.character(.data[[md_condition_col]]) else NA_character_,
      timepoint_join  = if (!is.null(md_timepoint_col)) as.character(.data[[md_timepoint_col]]) else NA_character_
    ) %>%
    distinct(sample, barcode_clean, .keep_all = TRUE)

  # sample-level fallback
  sample_meta_map <- metadata_prep %>%
    transmute(
      sample = md_sample,
      patient_id_sample = if (!is.null(md_patient_col)) as.character(.data[[md_patient_col]]) else NA_character_,
      condition_sample  = if (!is.null(md_condition_col)) as.character(.data[[md_condition_col]]) else NA_character_,
      timepoint_sample  = if (!is.null(md_timepoint_col)) as.character(.data[[md_timepoint_col]]) else NA_character_
    ) %>%
    distinct(sample, .keep_all = TRUE)

  contigs_all_pre <- contigs_all_pre %>%
    mutate(
      sample = as.character(sample),
      barcode_clean = sub("-1$", "", as.character(barcode))
    ) %>%
    left_join(barcode_meta_map, by = c("sample", "barcode_clean")) %>%
    left_join(sample_meta_map, by = "sample") %>%
    mutate(
      patient_id = dplyr::coalesce(patient_id_join, patient_id_sample),
      condition  = dplyr::coalesce(condition_join, condition_sample),
      timepoint  = dplyr::coalesce(timepoint_join, timepoint_sample)
    ) %>%
    select(-any_of(c(
      "barcode_clean",
      "patient_id_join", "condition_join", "timepoint_join",
      "patient_id_sample", "condition_sample", "timepoint_sample"
    )))

  message(sprintf(
    "Merged metadata: patient_id non-NA = %d / %d ; timepoint non-NA = %d / %d",
    sum(!is.na(contigs_all_pre$patient_id)), nrow(contigs_all_pre),
    sum(!is.na(contigs_all_pre$timepoint)), nrow(contigs_all_pre)
  ))
}
```

# metadata-debug
```{r}
#| label: metadata-debug
cat("\n===== DEBUG MERGED METADATA =====\n")
print(
  contigs_all_pre %>%
    select(sample, barcode, patient_id, condition, timepoint) %>%
    distinct() %>%
    head(20)
)


cat("\nPer-sample merged metadata summary:\n")
print(
  contigs_all_pre %>%
    group_by(sample) %>%
    summarise(
      patient_vals = paste(unique(na.omit(patient_id)), collapse = ";"),
      condition_vals = paste(unique(na.omit(condition)), collapse = ";"),
      timepoint_vals = paste(unique(na.omit(timepoint)), collapse = ";"),
      .groups = "drop"
    )
)
```

##chain mode
```{r}
#| label: chain-mode
detected_chain_mode <- infer_chain_mode(contigs_all_pre$chain)
effective_chain_mode <- if (params$chain_mode == "auto") detected_chain_mode else params$chain_mode

contigs_all_pre <- filter_chain_mode(contigs_all_pre, "chain", params$chain_mode)
```

##QC-filtering
```{r}
#| label: qc-filtering
contigs_all_pre <- contigs_all_pre %>%
  mutate(
    cdr3_len = nchar(cdr3),
    has_stop = ifelse(is.na(cdr3), NA, str_detect(cdr3, "\\*")),
    has_frame_issue = ifelse(is.na(cdr3_nt), FALSE, nchar(cdr3_nt) %% 3 != 0)
  )

contigs_all_post <- contigs_all_pre %>%
  {
    x <- .
    if (params$require_productive && "productive" %in% names(x)) {
      x <- x %>% filter(productive)
    }
    if (params$require_high_conf && "high_confidence" %in% names(x)) {
      x <- x %>% filter(high_confidence)
    }
    if (params$require_full_length && "full_length" %in% names(x)) {
      x <- x %>% filter(full_length)
    }
    if ("umis" %in% names(x)) {
      x <- x %>% filter(is.na(umis) | umis >= params$min_umis)
    }
    if ("reads" %in% names(x)) {
      x <- x %>% filter(is.na(reads) | reads >= params$min_reads)
    }

    x <- x %>% filter(!is.na(cdr3), cdr3 != "")

    if (isTRUE(params$drop_stop_or_frames)) {
      x <- x %>% filter(!grepl("\\*", cdr3))
      if ("cdr3_nt" %in% names(x)) {
        x <- x %>% filter(is.na(cdr3_nt) | nchar(cdr3_nt) %% 3 == 0)
      }
    }

    x <- x %>%
      filter(
        nchar(cdr3) >= params$cdr3_aa_min,
        nchar(cdr3) <= params$cdr3_aa_max
      )

    if (isTRUE(params$keep_one_per_chain)) {
      order_vars <- intersect(c("umis", "reads"), names(x))
      if (length(order_vars) > 0) {
        x <- x %>%
          group_by(sample, barcode, chain) %>%
          arrange(across(all_of(order_vars), desc), .by_group = TRUE) %>%
          slice(1) %>%
          ungroup()
      } else {
        x <- x %>% distinct(sample, barcode, chain, .keep_all = TRUE)
      }
    }

    x
  }

if (!isTRUE(params$keep_dual_alpha)) {
  contigs_all_post <- contigs_all_post %>%
    group_by(sample, barcode) %>%
    filter(!(chain == "TRA" & row_number() > 1)) %>%
    ungroup()
}

if (!isTRUE(params$keep_dual_beta)) {
  contigs_all_post <- contigs_all_post %>%
    group_by(sample, barcode) %>%
    filter(!(chain == "TRB" & row_number() > 1)) %>%
    ungroup()
}

if (isTRUE(params$keep_paired_only)) {
  if (effective_chain_mode == "T-AB") {
    contigs_all_post <- contigs_all_post %>%
      group_by(sample, barcode) %>%
      filter(any(chain == "TRA") & any(chain == "TRB")) %>%
      ungroup()
  } else if (effective_chain_mode == "T-GD") {
    contigs_all_post <- contigs_all_post %>%
      group_by(sample, barcode) %>%
      filter(any(chain == "TRG") & any(chain == "TRD")) %>%
      ungroup()
  }
}
```


## removed-contigs
```{r}
#| label: removed-contigs
key_cols <- intersect(
  c("sample","barcode","chain","cdr3","cdr3_nt","v_gene","j_gene","raw_clonotype_id"),
  colnames(contigs_all_pre)
)

if (length(key_cols) == 0) {
  removed <- contigs_all_pre[0, , drop = FALSE]
} else {
  removed <- dplyr::anti_join(
    contigs_all_pre,
    contigs_all_post,
    by = key_cols
  )
}
```


## QC-Summary before-after
```{r}
#| label: qc-summary-before-after
qc_summary <- contigs_all_pre %>%
  mutate(stage = "before") %>%
  bind_rows(contigs_all_post %>% mutate(stage = "after")) %>%
  group_by(sample, stage) %>%
  summarise(
    cells_with_any = n_distinct(barcode),
    contigs = n(),
    mean_contigs_per_cell = contigs / cells_with_any,
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = stage,
    values_from = c(cells_with_any, contigs, mean_contigs_per_cell)
  )

before_df <- contigs_all_pre %>% count(sample, name = "contigs_before")
after_df  <- contigs_all_post %>% count(sample, name = "contigs_after")

qc_retention <- before_df %>%
  left_join(after_df, by = "sample") %>%
  mutate(
    contigs_after = replace_na(contigs_after, 0L),
    dropped = pmax(contigs_before - contigs_after, 0L),
    retained_prop = ifelse(contigs_before > 0, contigs_after / contigs_before, NA_real_),
    retained_pct = 100 * retained_prop
  ) %>%
  arrange(desc(contigs_before))
```

### Pairing Summary
```{r}
#| label: pairing-summary
pairing_all <- contigs_all_post %>%
  group_by(sample, cell_id = barcode) %>%
  summarise(
    patient_id = first(na.omit(patient_id)) %||% NA_character_,
    condition = first(na.omit(condition)) %||% NA_character_,
    timepoint = first(na.omit(timepoint)) %||% NA_character_,
    n_prod = n(),
    n_TRA = sum(chain == "TRA"),
    n_TRB = sum(chain == "TRB"),
    n_TRG = sum(chain == "TRG"),
    n_TRD = sum(chain == "TRD"),
    has_TRA = any(chain == "TRA"),
    has_TRB = any(chain == "TRB"),
    has_TRG = any(chain == "TRG"),
    has_TRD = any(chain == "TRD"),
    alpha_ct = sum(chain == "TRA"),
    beta_ct = sum(chain == "TRB"),
    gamma_ct = sum(chain == "TRG"),
    delta_ct = sum(chain == "TRD"),
    cdr3a = paste(unique(cdr3[chain == "TRA"]), collapse = ";"),
    cdr3b = paste(unique(cdr3[chain == "TRB"]), collapse = ";"),
    cdr3g = paste(unique(cdr3[chain == "TRG"]), collapse = ";"),
    cdr3d = paste(unique(cdr3[chain == "TRD"]), collapse = ";"),
    trav = paste(unique(v_gene[chain == "TRA"]), collapse = ";"),
    trbv = paste(unique(v_gene[chain == "TRB"]), collapse = ";"),
    trgv = paste(unique(v_gene[chain == "TRG"]), collapse = ";"),
    trdv = paste(unique(v_gene[chain == "TRD"]), collapse = ";"),
    traj = paste(unique(j_gene[chain == "TRA"]), collapse = ";"),
    trbj = paste(unique(j_gene[chain == "TRB"]), collapse = ";"),
    trgj = paste(unique(j_gene[chain == "TRG"]), collapse = ";"),
    trdj = paste(unique(j_gene[chain == "TRD"]), collapse = ";"),
    clone_id = first(na.omit(raw_clonotype_id)) %||% NA_character_,
    max_reads = suppressWarnings(max(reads, na.rm = TRUE)),
    max_umis = suppressWarnings(max(umis, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    pairing = case_when(
      effective_chain_mode == "T-AB" & has_TRA & has_TRB ~ "Alpha+Beta",
      effective_chain_mode == "T-AB" & has_TRA & !has_TRB ~ "Alpha_only",
      effective_chain_mode == "T-AB" & !has_TRA & has_TRB ~ "Beta_only",
      effective_chain_mode == "T-GD" & has_TRG & has_TRD ~ "Gamma+Delta",
      effective_chain_mode == "T-GD" & has_TRG & !has_TRD ~ "Gamma_only",
      effective_chain_mode == "T-GD" & !has_TRG & has_TRD ~ "Delta_only",
      effective_chain_mode == "both" & (has_TRA & has_TRB) ~ "Alpha+Beta",
      effective_chain_mode == "both" & (has_TRG & has_TRD) ~ "Gamma+Delta",
      TRUE ~ "Other"
    ),
    multi_alpha = alpha_ct > 1,
    multi_beta = beta_ct > 1,
    multi_gamma = gamma_ct > 1,
    multi_delta = delta_ct > 1
  )
```



##CDR3 lengths
```{r}
#| label: cdr3-lengths
cdr3_len_all <- contigs_all_post %>%
  filter(!is.na(cdr3), cdr3 != "") %>%
  mutate(
    len = nchar(cdr3),
    chain_grp = case_when(
      chain == "TRA" ~ "TRA",
      chain == "TRB" ~ "TRB",
      chain == "TRG" ~ "TRG",
      chain == "TRD" ~ "TRD",
      TRUE ~ "Other"
    )
  )
```


##vj-usage
```{r}
#| label: vj-usage-and-clone-rank
v_usage <- NULL
j_usage <- NULL

if (isTRUE(params$show_vj_heatmaps) && any(!is.na(contigs_all_post$v_gene))) {
  v_usage <- contigs_all_post %>%
    filter(!is.na(v_gene), v_gene != "") %>%
    count(sample, v_gene, name = "n") %>%
    group_by(sample) %>%
    mutate(freq = n / sum(n)) %>%
    ungroup()

  top_v <- v_usage %>%
    group_by(v_gene) %>%
    summarise(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    slice_head(n = params$top_n_v) %>%
    pull(v_gene)

  v_usage <- v_usage %>% filter(v_gene %in% top_v)
}

if (isTRUE(params$show_vj_heatmaps) && any(!is.na(contigs_all_post$j_gene))) {
  j_usage <- contigs_all_post %>%
    filter(!is.na(j_gene), j_gene != "") %>%
    count(sample, j_gene, name = "n") %>%
    group_by(sample) %>%
    mutate(freq = n / sum(n)) %>%
    ungroup()

  top_j <- j_usage %>%
    group_by(j_gene) %>%
    summarise(total = sum(n), .groups = "drop") %>%
    arrange(desc(total)) %>%
    slice_head(n = params$top_n_j) %>%
    pull(j_gene)

  j_usage <- j_usage %>% filter(j_gene %in% top_j)
}

clone_rank_tbl <- NULL
if (isTRUE(params$show_clone_rank_plot)) {
  clone_rank_tbl <- pairing_all %>%
    filter(!is.na(clone_id), clone_id != "") %>%
    count(sample, clone_id, name = "n_cells") %>%
    group_by(sample) %>%
    arrange(desc(n_cells), .by_group = TRUE) %>%
    mutate(rank = row_number()) %>%
    ungroup() %>%
    filter(n_cells >= params$min_clone_size_plot)
}
```

##Compact Summary
```{r}
#| label: compact-sample-summary
pairing_rates <- pairing_all %>%
  count(sample, pairing, name = "cells") %>%
  group_by(sample) %>%
  mutate(frac = cells / sum(cells)) %>%
  pivot_wider(names_from = pairing, values_from = c(cells, frac), values_fill = 0) %>%
  ungroup()

sample_qc <- pairing_all %>%
  group_by(sample) %>%
  summarise(
    patient_id = first(na.omit(patient_id)) %||% NA_character_,
    condition = first(na.omit(condition)) %||% NA_character_,
    timepoint = first(na.omit(timepoint)) %||% NA_character_,
    n_cells_total = n(),
    n_cells_paired = sum(pairing %in% c("Alpha+Beta","Gamma+Delta")),
    n_cells_multichain = sum(multi_alpha | multi_beta | multi_gamma | multi_delta, na.rm = TRUE),
    pct_cells_paired = safe_pct(n_cells_paired, n_cells_total),
    pct_cells_multichain = safe_pct(n_cells_multichain, n_cells_total),
    n_unique_clones = n_distinct(clone_id[!is.na(clone_id) & clone_id != ""]),
    median_max_reads = suppressWarnings(median(max_reads[is.finite(max_reads)], na.rm = TRUE)),
    median_max_umis = suppressWarnings(median(max_umis[is.finite(max_umis)], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(qc_summary, by = "sample") %>%
  left_join(qc_retention, by = "sample") %>%
  left_join(pairing_rates, by = "sample")
```


## Optional metadata summaries
```{r}
#| label: metadata-panels
show_patient <- isTRUE(params$show_patient_panels) && any(!is.na(sample_qc$patient_id))
show_condition <- isTRUE(params$show_condition_panels) && any(!is.na(sample_qc$condition))
# show_timepoint <- isTRUE(params$show_timepoint_panels) && any(!is.na(sample_qc$timepoint))
show_timepoint <- isTRUE(params$show_timepoint_panels) &&
  any(!is.na(sample_qc$timepoint)) &&
  dplyr::n_distinct(na.omit(sample_qc$timepoint)) >= 2
```

##save outputs
```{r}
#| label: save-tables
save_table_safe(sample_sheet, "sample_sheet_resolved.tsv")
save_table_safe(contigs_all_pre, "contigs_before_qc.tsv")
save_table_safe(contigs_all_post, "contigs_after_qc.tsv")
save_table_safe(removed, "contigs_removed_by_qc.tsv")
save_table_safe(qc_summary, "vdj_qc_summary_before_after.tsv")
save_table_safe(qc_retention, "qc_contigs_before_after_summary.tsv")
save_table_safe(pairing_all, "pairing_status_all.tsv")
save_table_safe(sample_qc, "vdj_qc_per_sample_compact.tsv")
save_table_safe(cdr3_len_all, "cdr3_lengths_all.tsv")
save_table_safe(clonotypes_all, "clonotypes_all.tsv")
if (!is.null(metrics_all)) save_table_safe(metrics_all, "vdj_metrics_summary_all.tsv")
if (!is.null(v_usage)) save_table_safe(v_usage, "v_gene_usage.tsv")
if (!is.null(j_usage)) save_table_safe(j_usage, "j_gene_usage.tsv")
if (!is.null(clone_rank_tbl)) save_table_safe(clone_rank_tbl, "clone_rank_abundance.tsv")
```



## What this module does

::: {.callout-note}
This module performs quality control on Cell Ranger VDJ contig outputs before downstream clonotype analysis.  
It standardizes contig annotations, merges optional sample-level metadata, applies explicit QC filters, and summarizes the retained receptor repertoire at the contig, cell, and sample levels.  
This matters because downstream tools such as GLIPH2, TCRdist3, GIANA, repertoire analysis, and consensus clustering are only as reliable as the receptor calls entering them.  
The report is based primarily on filtered contig annotations, clonotype assignments, and optional sample metadata, and it evaluates retention, pairing, multi-chain burden, CDR3 properties, V/J usage, and clone size structure.
:::

## How to read this report

::: {.callout-tip}
- **Higher contig retention** usually indicates that most receptor calls survived the QC rules and were technically consistent with the expected assay.
- **Higher paired-chain fraction** is generally desirable for alpha-beta or gamma-delta analyses because paired receptors are more interpretable biologically.
- **High multi-chain burden** can reflect true biology in some cases, but can also indicate doublets, ambient contamination, or ambiguous contig recovery.
- **CDR3 length distributions** should look biologically plausible and fairly consistent across samples; strong shifts or extreme tails are potential red flags.
- **Clone rank-abundance curves** reveal whether repertoires are broad versus dominated by a few expanded clonotypes.
- **V/J usage heatmaps** help spot sample-specific biases, batch effects, or biologically enriched receptor programs.
:::

## Red flags and biologically interesting patterns

::: {.callout-warning}
**Potential red flags**
- very low retention after QC
- very low paired fraction
- unusually high multi-chain burden
- missing read or UMI support fields
- strong outlier samples in CDR3 length or V/J usage patterns

**Potentially interesting biology**
- strong clonal dominance in a subset of samples
- reproducible V/J enrichment across related samples
- distinct pairing behavior by patient, condition, or timepoint
- broad versus focused repertoire structure across the cohort
:::



## Overview
```{r}
#| label: Key summary metrics
#| results: asis

n_samples <- n_distinct(sample_sheet$sample)
n_contigs_before <- nrow(contigs_all_pre)
n_contigs_after <- nrow(contigs_all_post)
n_cells_after <- n_distinct(pairing_all$cell_id)
pct_retained_num <- n_contigs_after / max(1, n_contigs_before)
pct_retained <- percent(pct_retained_num, accuracy = 0.1)

paired_frac <- mean(pairing_all$pairing %in% c("Alpha+Beta","Gamma+Delta"), na.rm = TRUE)
multichain_frac <- mean(pairing_all$multi_alpha | pairing_all$multi_beta | pairing_all$multi_gamma | pairing_all$multi_delta, na.rm = TRUE)

top_sample_by_cells <- sample_qc %>%
  arrange(desc(n_cells_total)) %>%
  slice_head(n = 1) %>%
  pull(sample)

top_sample_by_retention <- qc_retention %>%
  arrange(desc(retained_prop)) %>%
  slice_head(n = 1) %>%
  pull(sample)

htmltools::browsable(
  htmltools::tagList(
    metric_card("Resolved samples", comma(n_samples), "libraries in sample sheet"),
    metric_card("Contigs before QC", comma(n_contigs_before), "raw filtered contigs"),
    metric_card("Contigs after QC", comma(n_contigs_after), "passing QC"),
    metric_card("Retention rate", pct_retained, "fraction of contigs retained"),
    metric_card("Cells after QC", comma(n_cells_after), "unique barcodes retained"),
    metric_card("Paired-chain fraction", percent(paired_frac, accuracy = 0.1), "post-QC cells"),
    metric_card("Multi-chain burden", percent(multichain_frac, accuracy = 0.1), "cells with >1 productive chain type"),
    metric_card("Detected chain mode", detected_chain_mode, "inferred from contigs"),
    metric_card("Effective chain mode", effective_chain_mode, "used for filtering"),
    metric_card("Top sample by cells", top_sample_by_cells %||% "NA", "largest retained sample"),
    metric_card("Top sample by retention", top_sample_by_retention %||% "NA", "highest QC retention")
  )
)

overview_tbl <- tibble(
  Metric = c(
    "Report label",
    "Resolved samples",
    "Detected chain mode",
    "Effective chain mode",
    "Contigs before QC",
    "Contigs after QC",
    "Retention rate",
    "Cells after QC",
    "Unique clonotypes loaded"
  ),
  Value = c(
    params$report_label,
    comma(n_samples),
    detected_chain_mode,
    effective_chain_mode,
    comma(n_contigs_before),
    comma(n_contigs_after),
    pct_retained,
    comma(n_cells_after),
    comma(nrow(clonotypes_all))
  )
)
print(
  scroll_kable(
    overview_tbl,
    caption = "High-level overview of VDJ QC analysis.",
    height = "260px"
  )
)

save_table_safe(overview_tbl, "vdj_qc_overview_metrics.tsv")
# kable(overview_tbl, caption = "High-level overview of VDJ QC analysis.") %>%
#   kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover","condensed"))
```

##sample-level Summary
```{r}
#| label: sample-summary-table
#| results: asis

# print(colnames(sample_qc))

top_n_samples_table <- params$top_n_samples_table
if (is.null(top_n_samples_table) || is.na(top_n_samples_table)) {
  top_n_samples_table <- 50
}
top_n_samples_table <- as.integer(top_n_samples_table)

sample_qc_display <- sample_qc %>%
  mutate(
    retained_pct = sprintf("%.1f%%", retained_pct),
    pct_cells_paired = percent(pct_cells_paired, accuracy = 0.1),
    pct_cells_multichain = percent(pct_cells_multichain, accuracy = 0.1)
  )

if ("contigs_after" %in% colnames(sample_qc_display)) {
  sample_qc_display <- sample_qc_display %>%
    arrange(desc(contigs_after))
} else {
  sample_qc_display <- sample_qc_display %>%
    arrange(desc(n_cells_total))
}

sample_qc_display <- sample_qc_display %>%
  slice_head(n = top_n_samples_table)

print(
  scroll_kable(
    sample_qc_display,
    caption = "Per-sample VDJ QC compact summary.",
    height = "420px",
    font_size = 11
  )
)

# kable(sample_qc_display, caption = "Per-sample VDJ QC compact summary.") %>%
#   kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover","condensed","responsive"))

# #| label: sample-summary-table
# print(colnames(sample_qc))
# 
# sample_qc_display <- sample_qc %>%
#   mutate(
#     retained_pct = sprintf("%.1f%%", retained_pct),
#     pct_cells_paired = percent(pct_cells_paired, accuracy = 0.1),
#     pct_cells_multichain = percent(pct_cells_multichain, accuracy = 0.1)
#   )
# 
# if ("contigs_after" %in% colnames(sample_qc_display)) {
#   sample_qc_display <- sample_qc_display %>%
#     arrange(desc(contigs_after))
# } else {
#   sample_qc_display <- sample_qc_display %>%
#     arrange(desc(n_cells_total))
# }
# 
# sample_qc_display <- sample_qc_display %>%
#   slice_head(n = params$top_n_samples_table)
# 
# kable(sample_qc_display, caption = "Per-sample VDJ QC compact summary.") %>%
#   kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover","condensed","responsive"))
```

## QC figure panels {.tabset}

### Retention and support

::: {.callout-note}
These plots summarize how strongly QC filtering affected each sample and whether retained contigs have adequate read and UMI support.
:::


```{r}
#| label: before-after-figures
p_before_counts <- qc_retention %>%
  ggplot(aes(x = reorder(sample, contigs_before), y = contigs_before)) +
  geom_col(fill = "grey50") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Contigs Before QC", x = NULL, y = "Contigs") +
  theme_scratch_pub(params$base_size)

p_after_counts <- qc_retention %>%
  ggplot(aes(x = reorder(sample, contigs_before), y = contigs_after)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Contigs After QC", x = NULL, y = "Contigs") +
  theme_scratch_pub(params$base_size)

p_retained <- qc_retention %>%
  ggplot(aes(x = reorder(sample, retained_prop), y = retained_pct)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  geom_text(aes(label = sprintf("%.1f%%", retained_pct)), hjust = 1.05, size = 3, color = "white") +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "Retention After QC", x = NULL, y = "Retained (%)") +
  theme_scratch_pub(params$base_size)

patch_before_after <- p_before_counts + p_after_counts + p_retained + plot_layout(ncol = 3)
patch_before_after
save_plot_safe(patch_before_after, glue("qc_before_after_retention.{params$figure_format}"), width = 16, height = 5)
```

```{r}
#| label: read-umi-plots
if (isTRUE(params$show_reads_umis) && any(!is.na(contigs_all_post$reads))) {
  p_reads <- contigs_all_post %>%
    filter(!is.na(reads), reads > 0) %>%
    ggplot(aes(x = reads, fill = chain)) +
    geom_histogram(bins = 50, alpha = 0.8, position = "identity") +
    scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
    labs(
      title = "Read Support of Passing Contigs",
      subtitle = "Histogram shown on log10 scale.",
      x = "Reads",
      y = "Contig count",
      fill = "Chain"
    ) +
    theme_scratch_pub(params$base_size)
  ggplotly_clean(p_reads)
  # print(p_reads)
  save_plot_safe(p_reads, glue("reads_support_histogram.{params$figure_format}"))
}

if (isTRUE(params$show_reads_umis) && any(!is.na(contigs_all_post$umis))) {
  p_umis <- contigs_all_post %>%
    filter(!is.na(umis), umis > 0) %>%
    ggplot(aes(x = umis, fill = chain)) +
    geom_histogram(bins = 50, alpha = 0.8, position = "identity") +
    scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
    labs(
      title = "UMI Support of Passing Contigs",
      subtitle = "Histogram shown on log10 scale.",
      x = "UMIs",
      y = "Contig count",
      fill = "Chain"
    ) +
    theme_scratch_pub(params$base_size)
  ggplotly_clean(p_umis)
  # print(p_umis)
  save_plot_safe(p_umis, glue("umis_support_histogram.{params$figure_format}"))
}
```


### Pairing and chain structure

::: {.callout-note}
Paired-chain recovery is one of the most important QC outcomes for TCR analysis. High paired fractions support reliable receptor interpretation, while elevated single-chain or multi-chain patterns can indicate technical dropout or ambiguous receptor assignment.
:::

#### Pairing status by sample
```{r}
#| label: pairing-plot-dropdown
#| warning: false
#| message: false

pairing_plot_tbl <- pairing_all %>%
  count(sample, pairing, name = "n") %>%
  mutate(
    pairing = factor(pairing, levels = unique(pairing))
  )

samples_pair <- sort(unique(pairing_plot_tbl$sample))
pairing_levels <- levels(pairing_plot_tbl$pairing)

# make sure every sample has all pairing categories
pairing_plot_tbl_complete <- tidyr::expand_grid(
  sample = samples_pair,
  pairing = pairing_levels
) %>%
  left_join(pairing_plot_tbl, by = c("sample", "pairing")) %>%
  mutate(n = ifelse(is.na(n), 0, n))

pairing_colors <- c(
  "Alpha_only" = "#F8766D",
  "Alpha+Beta" = "#00BA38",
  "Beta_only"  = "#619CFF"
)

# fallback if labels differ slightly
for (nm in unique(as.character(pairing_plot_tbl_complete$pairing))) {
  if (!nm %in% names(pairing_colors)) {
    pairing_colors[[nm]] <- "#999999"
  }
}

fig_pair <- plotly::plot_ly()

for (i in seq_along(samples_pair)) {
  s <- samples_pair[i]
  df_s <- pairing_plot_tbl_complete %>% dplyr::filter(sample == s)

  fig_pair <- fig_pair %>%
    plotly::add_bars(
      data = df_s,
      x = ~pairing,
      y = ~n,
      color = ~pairing,
      colors = pairing_colors,
      text = ~paste0(
        "Sample: ", s,
        "<br>Pairing: ", pairing,
        "<br>Cells: ", scales::comma(n)
      ),
      hoverinfo = "text",
      visible = if (i == 1) TRUE else FALSE,
      showlegend = if (i == 1) TRUE else FALSE
    )
}

n_traces_per_sample <- length(pairing_levels)

buttons_pair <- lapply(seq_along(samples_pair), function(i) {
  vis <- rep(FALSE, length(samples_pair) * n_traces_per_sample)
  idx <- ((i - 1) * n_traces_per_sample + 1):(i * n_traces_per_sample)
  vis[idx] <- TRUE

  list(
    method = "update",
    args = list(
      list(visible = vis, showlegend = vis),
      list(
        title = list(text = paste0("Pairing status: ", samples_pair[i]))
      )
    ),
    label = samples_pair[i]
  )
})

fig_pair <- fig_pair %>%
  plotly::layout(
    title = list(text = paste0("Pairing status: ", samples_pair[1])),
    barmode = "group",
    xaxis = list(title = ""),
    yaxis = list(title = "Cells"),
    updatemenus = list(
      list(
        type = "dropdown",
        active = 0,
        x = 0.02,
        y = 1.15,
        xanchor = "left",
        yanchor = "top",
        buttons = buttons_pair
      )
    ),
    annotations = list(
      list(
        x = 0.0,
        y = 1.19,
        xref = "paper",
        yref = "paper",
        text = "<b>Select sample:</b>",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top"
      )
    ),
    legend = list(title = list(text = "Pairing")),
    margin = list(t = 110)
  )

fig_pair

```

#### Multiple productive chains by sample
```{r}
#| label: multichain-plot-dropdown
#| warning: false
#| message: false

multi_chain_long <- pairing_all %>%
  pivot_longer(
    cols = c("multi_alpha", "multi_beta", "multi_gamma", "multi_delta"),
    names_to = "type",
    values_to = "flag"
  ) %>%
  mutate(
    type = recode(
      type,
      multi_alpha = "Multiple TRA",
      multi_beta  = "Multiple TRB",
      multi_gamma = "Multiple TRG",
      multi_delta = "Multiple TRD"
    ),
    flag = factor(flag, levels = c(FALSE, TRUE), labels = c("No", "Yes"))
  ) %>%
  count(sample, type, flag, name = "n")

samples_multi <- sort(unique(multi_chain_long$sample))
type_levels   <- unique(multi_chain_long$type)
flag_levels   <- levels(multi_chain_long$flag)

multi_chain_complete <- tidyr::expand_grid(
  sample = samples_multi,
  type   = type_levels,
  flag   = flag_levels
) %>%
  left_join(multi_chain_long, by = c("sample", "type", "flag")) %>%
  mutate(n = ifelse(is.na(n), 0, n))

# keep a saved static all-samples version for export
n_samples_multi <- dplyr::n_distinct(multi_chain_complete$sample)
facet_ncol <- min(4, max(1, n_samples_multi))
facet_nrow <- ceiling(n_samples_multi / facet_ncol)

plot_width  <- max(14, facet_ncol * 3.8)
plot_height <- max(16, facet_nrow * 5.5)

p_multi_by <- ggplot(multi_chain_complete, aes(x = type, y = n, fill = flag)) +
  geom_col(position = "stack") +
  facet_wrap(~ sample, scales = "free_y", ncol = facet_ncol) +
  labs(
    title = "Multiple Productive Chains",
    subtitle = "Post-QC multiple-chain flags by sample.",
    x = NULL,
    y = "Cells",
    fill = "Flag"
  ) +
  theme_scratch_pub(params$base_size) +
  theme(
    axis.text.x = element_text(size = params$base_size - 4, angle = 45, hjust = 1),
    axis.text.y = element_text(size = params$base_size - 2),
    axis.title.y = element_text(size = params$base_size + 1, face = "bold"),
    strip.text = element_text(size = params$base_size, face = "bold"),
    plot.margin = margin(15, 15, 15, 15)
  )

save_plot_safe(
  p_multi_by,
  glue("multiple_chains_by_sample.{params$figure_format}"),
  width = plot_width,
  height = plot_height
)

flag_colors <- c("No" = "#F8766D", "Yes" = "#00BFC4")

fig_multi <- plotly::plot_ly()

for (i in seq_along(samples_multi)) {
  s <- samples_multi[i]
  df_s <- multi_chain_complete %>% dplyr::filter(sample == s)

  fig_multi <- fig_multi %>%
    plotly::add_bars(
      data = df_s,
      x = ~type,
      y = ~n,
      color = ~flag,
      colors = flag_colors,
      text = ~paste0(
        "Sample: ", s,
        "<br>Type: ", type,
        "<br>Flag: ", flag,
        "<br>Cells: ", scales::comma(n)
      ),
      hoverinfo = "text",
      visible = if (i == 1) TRUE else FALSE,
      showlegend = if (i == 1) TRUE else FALSE
    )
}

n_traces_per_sample <- length(flag_levels)

buttons_multi <- lapply(seq_along(samples_multi), function(i) {
  vis <- rep(FALSE, length(samples_multi) * n_traces_per_sample)
  idx <- ((i - 1) * n_traces_per_sample + 1):(i * n_traces_per_sample)
  vis[idx] <- TRUE

  list(
    method = "update",
    args = list(
      list(visible = vis, showlegend = vis),
      list(
        title = list(text = paste0("Multiple productive chains: ", samples_multi[i]))
      )
    ),
    label = samples_multi[i]
  )
})

fig_multi <- fig_multi %>%
  plotly::layout(
    title = list(text = paste0("Multiple productive chains: ", samples_multi[1])),
    barmode = "stack",
    xaxis = list(title = ""),
    yaxis = list(title = "Cells"),
    updatemenus = list(
      list(
        type = "dropdown",
        active = 0,
        x = 0.02,
        y = 1.15,
        xanchor = "left",
        yanchor = "top",
        buttons = buttons_multi
      )
    ),
    annotations = list(
      list(
        x = 0.0,
        y = 1.19,
        xref = "paper",
        yref = "paper",
        text = "<b>Select sample:</b>",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top"
      )
    ),
    legend = list(title = list(text = "Flag")),
    margin = list(t = 110)
  )

fig_multi
```


### CDR3 properties

::: {.callout-note}
These plots assess whether post-QC CDR3 amino acid lengths are biologically plausible and reasonably stable across samples.
:::

```{r}
#| label: cdr3-length
p_len_all <- ggplot(cdr3_len_all, aes(x = len, fill = chain_grp)) +
  geom_histogram(position = "identity", alpha = 0.7, bins = 40) +
  labs(
    title = "CDR3 Length Distribution",
    subtitle = "Post-QC pooled across all samples.",
    x = "CDR3 amino acid length",
    y = "Count",
    fill = "Chain"
  ) +
  theme_scratch_pub(params$base_size)

p_len_all
save_plot_safe(p_len_all, glue("cdr3_length_hist_all.{params$figure_format}"))
```

#### CDR3 length distribution by sample

```{r}
#| label: cdr3-length-dropdown
#| warning: false
#| message: false

n_samples_facet <- dplyr::n_distinct(cdr3_len_all$sample)
facet_ncol <- min(4, max(1, n_samples_facet))
facet_nrow <- ceiling(n_samples_facet / facet_ncol)

plot_width  <- max(14, facet_ncol * 3.8)
plot_height <- max(16, facet_nrow * 5.5)

# keep static all-samples faceted histogram for export
p_len_facet <- ggplot(cdr3_len_all, aes(x = len, fill = chain_grp)) +
  geom_histogram(position = "identity", alpha = 0.7, bins = 40) +
  facet_wrap(~ sample, scales = "free_y", ncol = facet_ncol) +
  labs(
    title = "CDR3 Length Distribution by Sample",
    subtitle = "Post-QC distribution faceted by sample.",
    x = "CDR3 amino acid length",
    y = "Count",
    fill = "Chain"
  ) +
  theme_scratch_pub(params$base_size) +
  theme(
    axis.text.x = element_text(size = params$base_size - 4, angle = 45, hjust = 1),
    axis.text.y = element_text(size = params$base_size - 2),
    strip.text = element_text(size = params$base_size, face = "bold"),
    plot.margin = margin(15, 15, 15, 15)
  )

save_plot_safe(
  p_len_facet,
  glue("cdr3_length_hist_by_sample.{params$figure_format}"),
  width = plot_width,
  height = plot_height
)

# keep static boxplot for export
cdr3_box <- ggplot(cdr3_len_all, aes(x = sample, y = len, fill = chain_grp)) +
  geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
  labs(
    title = "CDR3 Length by Sample",
    x = NULL,
    y = "CDR3 amino acid length",
    fill = "Chain"
  ) +
  theme_scratch_pub(params$base_size) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

save_plot_safe(
  cdr3_box,
  glue("cdr3_length_boxplot_by_sample.{params$figure_format}"),
  width = 14,
  height = 6
)

samples_cdr3 <- sort(unique(cdr3_len_all$sample))
chain_levels <- unique(cdr3_len_all$chain_grp)

# consistent colors
chain_colors <- c(
  "TRA" = "#F8766D",
  "TRB" = "#00BFC4",
  "TRG" = "#7CAE00",
  "TRD" = "#C77CFF",
  "Other" = "#999999"
)

for (nm in chain_levels) {
  if (!nm %in% names(chain_colors)) {
    chain_colors[[nm]] <- "#999999"
  }
}

# build one histogram per sample
hist_plots <- lapply(samples_cdr3, function(s) {
  df_s <- cdr3_len_all %>% dplyr::filter(sample == s)

  p_one <- ggplot(df_s, aes(x = len, fill = chain_grp)) +
    geom_histogram(position = "identity", alpha = 0.7, bins = 40) +
    labs(
      title = paste0("CDR3 length distribution: ", s),
      x = "CDR3 amino acid length",
      y = "Count",
      fill = "Chain"
    ) +
    scale_fill_manual(values = chain_colors, drop = FALSE) +
    theme_scratch_pub(params$base_size)

  plotly::ggplotly(p_one, tooltip = c("x", "y", "fill")) %>%
    plotly::layout(height = 520)
})

# dropdown selector
fig_cdr3 <- hist_plots[[1]]

buttons_cdr3 <- lapply(seq_along(samples_cdr3), function(i) {
  list(
    method = "restyle",
    args = list("visible", TRUE),
    label = samples_cdr3[i],
    execute = FALSE
  )
})

# use htmlwidgets onRender to switch full plot objects via dropdown is harder than bars;
# instead use subplot traces approach below

fig_cdr3 <- plotly::plot_ly()

trace_counts <- integer(length(samples_cdr3))

for (i in seq_along(samples_cdr3)) {
  s <- samples_cdr3[i]
  df_s <- cdr3_len_all %>% dplyr::filter(sample == s)

  for (ch in chain_levels) {
    df_ch <- df_s %>% dplyr::filter(chain_grp == ch)

    fig_cdr3 <- fig_cdr3 %>%
      plotly::add_histogram(
        data = df_ch,
        x = ~len,
        name = ch,
        marker = list(color = unname(chain_colors[ch])),
        opacity = 0.7,
        nbinsx = 40,
        visible = if (i == 1) TRUE else FALSE,
        showlegend = if (i == 1) TRUE else FALSE,
        hovertemplate = paste0(
          "Sample: ", s,
          "<br>Chain: ", ch,
          "<br>CDR3 length: %{x}",
          "<br>Count: %{y}<extra></extra>"
        )
      )
  }

  trace_counts[i] <- length(chain_levels)
}

buttons_cdr3 <- lapply(seq_along(samples_cdr3), function(i) {
  vis <- rep(FALSE, sum(trace_counts))
  start_idx <- sum(trace_counts[seq_len(i - 1)]) + 1
  end_idx   <- sum(trace_counts[seq_len(i)])
  vis[start_idx:end_idx] <- TRUE

  list(
    method = "update",
    args = list(
      list(visible = vis, showlegend = vis),
      list(
        title = list(text = paste0("CDR3 length distribution: ", samples_cdr3[i]))
      )
    ),
    label = samples_cdr3[i]
  )
})

fig_cdr3 <- fig_cdr3 %>%
  plotly::layout(
    title = list(text = paste0("CDR3 length distribution: ", samples_cdr3[1])),
    barmode = "overlay",
    xaxis = list(title = "CDR3 amino acid length"),
    yaxis = list(title = "Count"),
    updatemenus = list(
      list(
        type = "dropdown",
        active = 0,
        x = 0.02,
        y = 1.15,
        xanchor = "left",
        yanchor = "top",
        buttons = buttons_cdr3
      )
    ),
    annotations = list(
      list(
        x = 0.0,
        y = 1.19,
        xref = "paper",
        yref = "paper",
        text = "<b>Select sample:</b>",
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "top"
      )
    ),
    legend = list(title = list(text = "Chain")),
    margin = list(t = 110)
  )

fig_cdr3

```

#### Cohort-wide CDR3 length boxplot

```{r}
#| label: cdr3-boxplot-display
plotly::plot_ly(
  data = cdr3_len_all,
  x = ~sample,
  y = ~len,
  color = ~chain_grp,
  colors = c("TRA" = "#F8766D", "TRB" = "#00BFC4", "TRG" = "#7CAE00", "TRD" = "#C77CFF", "Other" = "#999999"),
  type = "box",
  boxpoints = "outliers",
  hovertemplate = paste(
    "Sample: %{x}<br>",
    "CDR3 length: %{y}<br>",
    "Chain: %{color}<extra></extra>"
  )
) %>%
  plotly::layout(
    title = list(text = "CDR3 Length by Sample"),
    xaxis = list(title = ""),
    yaxis = list(title = "CDR3 amino acid length")
  )


```


### Gene usage

::: {.callout-note}
These heatmaps summarize post-QC V and J gene segment usage across samples and can highlight biological enrichment or sample-specific receptor bias.
:::

```{r}
#| label: v-gene-heatmap
if (!is.null(v_usage) && nrow(v_usage) > 0) {
  v_mat <- v_usage %>%
    select(sample, v_gene, freq) %>%
    pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
    as.data.frame()

  rownames(v_mat) <- v_mat$v_gene
  v_mat$v_gene <- NULL
  v_mat <- as.matrix(v_mat)

  max_v <- max(v_mat, na.rm = TRUE)
  if (!is.finite(max_v) || max_v <= 0) max_v <- 1

  ht_v <- Heatmap(
    v_mat,
    name = "Freq",
    col = colorRamp2(
      c(0, max_v / 2, max_v),
      c("white", "gold", "firebrick")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 7),
    column_names_rot = 45,
    column_title = "Top V gene usage across samples",
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),
    heatmap_legend_param = list(title = "Frequency")
  )

  draw(ht_v, padding = grid::unit(c(5, 5, 5, 12), "mm"))
}

# #| label: v-gene-heatmap
# if (!is.null(v_usage) && nrow(v_usage) > 0) {
#   v_mat <- v_usage %>%
#     select(sample, v_gene, freq) %>%
#     pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
#     as.data.frame()
# 
#   rownames(v_mat) <- v_mat$v_gene
#   v_mat$v_gene <- NULL
#   v_mat <- as.matrix(v_mat)
# 
#   ht_v <- Heatmap(
#     v_mat,
#     name = "Freq",
#     col = colorRamp2(
#       c(0, max(v_mat, na.rm = TRUE) / 2, max(v_mat, na.rm = TRUE)),
#       c("white", "gold", "firebrick")
#     ),
#     cluster_rows = TRUE,
#     cluster_columns = TRUE,
#     row_names_side = "left",
#     column_title = "Top V gene usage across samples",
#     heatmap_legend_param = list(title = "Frequency")
#   )
# 
#   draw(ht_v)
# }
```

```{r}
#| label: j-gene-heatmap
if (!is.null(j_usage) && nrow(j_usage) > 0) {
  j_mat <- j_usage %>%
    select(sample, j_gene, freq) %>%
    pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
    as.data.frame()

  rownames(j_mat) <- j_mat$j_gene
  j_mat$j_gene <- NULL
  j_mat <- as.matrix(j_mat)

  max_j <- max(j_mat, na.rm = TRUE)
  if (!is.finite(max_j) || max_j <= 0) max_j <- 1

  ht_j <- Heatmap(
    j_mat,
    name = "Freq",
    col = colorRamp2(
      c(0, max_j / 2, max_j),
      c("white", "skyblue", "navy")
    ),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 7),
    column_names_rot = 45,
    column_title = "Top J gene usage across samples",
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),
    heatmap_legend_param = list(title = "Frequency")
  )

  draw(ht_j, padding = grid::unit(c(5, 5, 5, 12), "mm"))
}

# #| label: j-gene-heatmap
# if (!is.null(j_usage) && nrow(j_usage) > 0) {
#   j_mat <- j_usage %>%
#     select(sample, j_gene, freq) %>%
#     pivot_wider(names_from = sample, values_from = freq, values_fill = 0) %>%
#     as.data.frame()
# 
#   rownames(j_mat) <- j_mat$j_gene
#   j_mat$j_gene <- NULL
#   j_mat <- as.matrix(j_mat)
# 
#   ht_j <- Heatmap(
#     j_mat,
#     name = "Freq",
#     col = colorRamp2(
#       c(0, max(j_mat, na.rm = TRUE) / 2, max(j_mat, na.rm = TRUE)),
#       c("white", "skyblue", "navy")
#     ),
#     cluster_rows = TRUE,
#     cluster_columns = TRUE,
#     row_names_side = "left",
#     column_title = "Top J gene usage across samples",
#     heatmap_legend_param = list(title = "Frequency")
#   )
# 
#   draw(ht_j)
# }
```

### Clonality

::: {.callout-note}
Clone rank-abundance summarizes whether the repertoire is broadly distributed or dominated by expanded clonotypes.
:::


```{r}
#| label: clone-rank-plot
#| fig-width: 16
#| fig-height: !expr max(12, 7 + ceiling(dplyr::n_distinct(pairing_all$sample) / 6) * 0.8)

# 1. Create the plotting table
clone_rank_tbl_plot <- pairing_all %>%
  filter(!is.na(clone_id), clone_id != "") %>%
  count(sample, clone_id, name = "n_cells") %>%
  group_by(sample) %>%
  arrange(desc(n_cells), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

save_table_safe(clone_rank_tbl_plot, "clone_rank_abundance.tsv")

# 2. Build the plot
if (nrow(clone_rank_tbl_plot) > 0) {
  n_samples_clone <- dplyr::n_distinct(clone_rank_tbl_plot$sample)
  legend_rows <- ceiling(n_samples_clone / 6)
  
  # Adjusted height calculation for a more compact look
  plot_height <- max(12, 7 + legend_rows * 0.8)

  p_clone_rank <- ggplot(clone_rank_tbl_plot, aes(x = rank, y = n_cells, color = sample)) +
    geom_line(linewidth = 0.8, alpha = 0.8) + 
    geom_point(size = 1.5, alpha = 0.8) +    
    scale_x_log10() +
    scale_y_log10() +
    labs(
      title = "Clone Rank-Abundance Curve",
      subtitle = "All detected clonotypes are shown.",
      x = "Clone rank (log10)",
      y = "Cells per clone (log10)",
      color = "Sample"
    ) +
    theme_scratch_pub(params$base_size) +
    theme(
      # LEGEND POSITION: Change "bottom" to "right" if you want it on the side
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      
      # AXIS TEXT (Tick labels: 1, 10, 100) - Increased to +4
      axis.text = element_text(size = params$base_size + 4, color = "black"),
      
      # AXIS TITLES (Labels: Clone rank, Cells per clone) - Increased to +6
      axis.title = element_text(size = params$base_size + 6, face = "bold"),
      
      # LEGEND TEXT (Sample IDs) - Increased to +2
      legend.text = element_text(size = params$base_size + 2),
      legend.title = element_text(size = params$base_size + 4, face = "bold"),
      
      plot.title = element_text(size = params$base_size + 8, face = "bold"),
      plot.subtitle = element_text(size = params$base_size + 4),
      
      # Log scales look better without x-axis rotation
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.grid.major = element_line(linewidth = 0.5, color = "grey90")
    ) +
    guides(color = guide_legend(nrow = legend_rows, byrow = TRUE))

  print(p_clone_rank)
  save_plot_safe(
    p_clone_rank,
    glue("clone_rank_abundance.{params$figure_format}"),
    width = 16,
    height = plot_height
  )
} else {
  cat("No clonotypes were available to build the clone rank-abundance curve.")
}

# #| label: clone-rank
# 
# 
# clone_rank_tbl_plot <- pairing_all %>%
#   filter(!is.na(clone_id), clone_id != "") %>%
#   count(sample, clone_id, name = "n_cells") %>%
#   group_by(sample) %>%
#   arrange(desc(n_cells), .by_group = TRUE) %>%
#   mutate(rank = row_number()) %>%
#   ungroup()
# 
# save_table_safe(clone_rank_tbl_plot, "clone_rank_abundance.tsv")
# 
# if (nrow(clone_rank_tbl_plot) > 0) {
#   p_clone_rank <- ggplot(clone_rank_tbl_plot, aes(x = rank, y = n_cells, color = sample)) +
#     geom_line(linewidth = 0.9, alpha = 0.8) +
#     geom_point(size = 1.2, alpha = 0.9) +
#     scale_x_log10() +
#     scale_y_log10() +
#     labs(
#       title = "Clone Rank-Abundance Curve",
#       subtitle = "All detected clonotypes are shown.",
#       x = "Clone rank (log10)",
#       y = "Cells per clone (log10)",
#       color = "Sample"
#     ) +
#     theme_scratch_pub(params$base_size)
# 
#   print(p_clone_rank)
#   save_plot_safe(p_clone_rank, glue("clone_rank_abundance.{params$figure_format}"))
# } else {
#   cat("No clonotypes were available to build the clone rank-abundance curve.")
# }
# 
# # if (!is.null(clone_rank_tbl) && nrow(clone_rank_tbl) > 0) {
# #   p_clone_rank <- ggplot(clone_rank_tbl, aes(x = rank, y = n_cells, color = sample)) +
# #     geom_line(linewidth = 0.9, alpha = 0.8) +
# #     geom_point(size = 1.2, alpha = 0.9) +
# #     scale_x_log10() +
# #     scale_y_log10() +
# #     labs(
# #       title = "Clone Rank-Abundance Curve",
# #       subtitle = glue("Only clones with at least {params$min_clone_size_plot} cells are shown."),
# #       x = "Clone rank (log10)",
# #       y = "Cells per clone (log10)",
# #       color = "Sample"
# #     ) +
# #     theme_scratch_pub(params$base_size)
# # 
# #   p_clone_rank
# #   save_plot_safe(p_clone_rank, glue("clone_rank_abundance.{params$figure_format}"))
# # }
```

### Metadata-stratified summaries

::: {.callout-note}
These summaries are shown only when patient, condition, or timepoint metadata are available.
:::

```{r}
#| label: patient-summary
#| results: asis


if (show_patient) {
  patient_source <- sample_qc %>%
    select(sample, patient_id, n_cells_total, n_unique_clones) %>%
    distinct() %>%
    left_join(
      qc_retention %>% select(sample, contigs_before, contigs_after, retained_prop, retained_pct),
      by = "sample"
    )

  patient_tbl <- patient_source %>%
    filter(!is.na(patient_id), patient_id != "") %>%
    group_by(patient_id) %>%
    summarise(
      n_samples = n_distinct(sample),
      contigs_before = sum(contigs_before, na.rm = TRUE),
      contigs_after = sum(contigs_after, na.rm = TRUE),
      retained_prop = contigs_after / pmax(1, contigs_before),
      retained_pct = 100 * retained_prop,
      n_cells_total = sum(n_cells_total, na.rm = TRUE),
      n_unique_clones = sum(n_unique_clones, na.rm = TRUE),
      .groups = "drop"
    )

  print(
    scroll_kable(
      patient_tbl,
      caption = "Patient-level VDJ QC summary.",
      height = "260px"
    )
  )

  p_patient <- patient_tbl %>%
    mutate(patient_id = fct_reorder(patient_id, retained_prop)) %>%
    ggplot(aes(x = patient_id, y = retained_pct)) +
    geom_col(fill = "darkgreen") +
    coord_flip() +
    labs(
      title = "QC Retention by Patient",
      x = NULL,
      y = "Retained (%)"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_patient)
  save_plot_safe(p_patient, glue("retention_by_patient.{params$figure_format}"))

  save_table_safe(patient_tbl, "patient_level_vdj_qc_summary.tsv")
}

```

```{r}
#| label: condition-summary
#| results: asis

if (show_condition) {
  condition_source <- sample_qc %>%
    select(sample, condition, n_cells_total) %>%
    distinct() %>%
    left_join(
      qc_retention %>% select(sample, contigs_before, contigs_after, retained_prop, retained_pct),
      by = "sample"
    )

  condition_tbl <- condition_source %>%
    filter(!is.na(condition), condition != "") %>%
    group_by(condition) %>%
    summarise(
      n_samples = n_distinct(sample),
      contigs_before = sum(contigs_before, na.rm = TRUE),
      contigs_after = sum(contigs_after, na.rm = TRUE),
      retained_prop = contigs_after / pmax(1, contigs_before),
      retained_pct = 100 * retained_prop,
      n_cells_total = sum(n_cells_total, na.rm = TRUE),
      .groups = "drop"
    )

  print(
    scroll_kable(
      condition_tbl,
      caption = "Condition-level VDJ QC summary.",
      height = "260px"
    )
  )

  p_condition <- condition_tbl %>%
    mutate(condition = fct_reorder(condition, retained_prop)) %>%
    ggplot(aes(x = condition, y = retained_pct)) +
    geom_col(fill = "purple4") +
    coord_flip() +
    labs(
      title = "QC Retention by Condition",
      x = NULL,
      y = "Retained (%)"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_condition)
  save_plot_safe(p_condition, glue("retention_by_condition.{params$figure_format}"))

  save_table_safe(condition_tbl, "condition_level_vdj_qc_summary.tsv")
}
```

```{r}
#| label: timepoint-summary
#| results: asis

if (show_timepoint) {
  time_source <- sample_qc %>%
    select(sample, timepoint) %>%
    distinct() %>%
    left_join(
      qc_retention %>% select(sample, contigs_before, contigs_after, retained_prop, retained_pct),
      by = "sample"
    )

  time_tbl <- time_source %>%
    filter(!is.na(timepoint), timepoint != "") %>%
    group_by(timepoint) %>%
    summarise(
      n_samples = n_distinct(sample),
      contigs_before = sum(contigs_before, na.rm = TRUE),
      contigs_after = sum(contigs_after, na.rm = TRUE),
      retained_prop = contigs_after / pmax(1, contigs_before),
      retained_pct = 100 * retained_prop,
      .groups = "drop"
    )

  print(
    scroll_kable(
      time_tbl,
      caption = "Timepoint-level VDJ QC summary.",
      height = "260px"
    )
  )

  p_time <- time_tbl %>%
    mutate(timepoint = fct_inorder(timepoint)) %>%
    ggplot(aes(x = timepoint, y = retained_pct, group = 1)) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_point(size = 3, color = "steelblue4") +
    labs(
      title = "QC Retention Across Timepoints",
      x = "Timepoint",
      y = "Retained (%)"
    ) +
    theme_scratch_pub(params$base_size)

  print(p_time)
  save_plot_safe(p_time, glue("retention_by_timepoint.{params$figure_format}"))

  save_table_safe(time_tbl, "timepoint_level_vdj_qc_summary.tsv")
}
```

## Warnings
```{r}
#| label: warnings
warn_tbl <- tibble(
  warning = c(
    "Low contig retention",
    "High multi-chain burden",
    "Low paired fraction",
    "Missing reads column",
    "Missing UMIs column",
    "Missing patient metadata",
    "Missing condition metadata",
    "Missing timepoint metadata"
  ),
  triggered = c(
    n_contigs_after / max(1, n_contigs_before) < 0.5,
    mean(pairing_all$multi_alpha | pairing_all$multi_beta | pairing_all$multi_gamma | pairing_all$multi_delta, na.rm = TRUE) > 0.1,
    mean(pairing_all$pairing %in% c("Alpha+Beta","Gamma+Delta"), na.rm = TRUE) < 0.3,
    all(is.na(contigs_all_post$reads)),
    all(is.na(contigs_all_post$umis)),
    !show_patient,
    !show_condition,
    !show_timepoint
  ),
  interpretation = c(
    "Less than half of contigs were retained after QC. Inspect productive, high-confidence, and full-length filters as well as barcode/sample path resolution.",
    "More than 10% of cells are multi-chain under the current settings. This may indicate biological dual-chain usage, technical artifacts, or doublets.",
    "Paired-chain fraction is low after QC. Check assay recovery, allowed chain mode, and pairing requirements.",
    "Read-support plots were skipped because no reads column was available.",
    "UMI-support plots were skipped because no UMIs column was available.",
    "Patient-level summaries were skipped because patient metadata were unavailable.",
    "Condition-level summaries were skipped because condition metadata were unavailable.",
    "Timepoint-level summaries were skipped because timepoint metadata were unavailable."
  )
) %>%
  filter(triggered)

if (nrow(warn_tbl) == 0) {
  cat("No major automatic warnings were triggered under the current QC settings.")
} else {
  kable(warn_tbl %>% select(-triggered), caption = "Automatically generated QC warnings and notes.") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover","condensed"))
}
```

## session info
```{r}
#| label: session-info
writeLines(capture.output(sessionInfo()), file.path(params$outdir, "sessionInfo.vdj_qc.txt"))
sessionInfo()
```


